### 2018-05-08 slides/pnwplse.tex
start slides

 \documentclass{beamer}
\usefonttheme{professionalfonts}
\usepackage{agda}
\usepackage{catchfilebetweentags}
\usetheme{CambridgeUS}
\usecolortheme{seahorse}

\usepackage{amssymb}
\usepackage{bbm}
\usepackage[greek,english]{babel}

\beamertemplatenavigationsymbolsempty

\title{Musical Ornaments}
\author{John Leo}
\institute{Halfaya Research}
\date{May 14, 2018}

\begin{document}

\frame{\titlepage}

\begin{frame}\frametitle{Outline}
\begin{itemize}
\item A look toward the future
\item Music Tools
\item Equivalences
\item Ornaments
\end{itemize}

\bigskip 
Sources:
\begin{itemize}
\item {\tt https://github.com/halfaya/MusicTools}
\end{itemize}
\end{frame}

\begin{frame}\frametitle{Robert Harper}
\begin{quote}
Eventually all the arbitrary programming languages are going to be just swept away with the oceans,
and we will have the permanence of constructive, intuistionistic type theory as the master theory
of computation---without doubt, in my mind, no question.  So, from my point of view---this is a personal
statement---working in anything else is a waste of time.
\end{quote}

CMU Homotopy Type Theory lecture 1, 52:56--53:20.
\end{frame}

\begin{frame}\frametitle{Music Representation}
\ExecuteMetaData[../agda/latex/Pnwplse.tex]{pitch}
\ExecuteMetaData[../agda/latex/Pnwplse.tex]{duration}
\ExecuteMetaData[../agda/latex/Pnwplse.tex]{note}
\ExecuteMetaData[../agda/latex/Pnwplse.tex]{music}
\end{frame}

\begin{frame}\frametitle{Conclusion}
Goes here.
\end{frame}

\end{document}

### 2018-05-11 slides/pnwplse.tex
more work on talk

@@
 
-\begin{frame}\frametitle{Music Representation}
+\begin{frame}\frametitle{What will programming look like in 50 years?}
+\begin{itemize}
+\item Convergence of math and computer science
+\item Functional Programming, Algebra of Programming
+\item Dependent Types or a successor (Cubical?)
+\item Who does the programming?
+\end{itemize}
+\end{frame}
+
+\begin{frame}\frametitle{How do we get there from here?}
+\begin{itemize}
+\item Add dependent types to an industrial-strength language (Haskell)
+\item Make a dependently typed language (Agda, Idris) practical to use
+\item Learn how to program using dependent types
+\item Many theoretical and practical advances are still needed
+\end{itemize}
+\end{frame}
+
+\begin{frame}\frametitle{Euterpea}
+\begin{figure}
+\includegraphics[height=0.8\textheight]{hsom}
+\end{figure}
+\end{frame}
+
+\begin{frame}\frametitle{Look vs Time (1997)}
+\begin{figure}
+\includegraphics[width=\textwidth]{lookVsTime}
+\end{figure}
+\end{frame}
+
+\begin{frame}\frametitle{Music Representation \`a la Euterpea}
 \ExecuteMetaData[../agda/latex/Pnwplse.tex]{pitch}

### 2018-05-13 slides/pnwplse.tex
slides complete

@@
 \item Music Tools
-\item Equivalences
-\item Ornaments
+\item Equivalences and Ornaments
 \end{itemize}
@@
 
+\begin{frame}\frametitle{Music Tools}
+\begin{itemize}
+\item Collection of composable tools for synthesis and analysis of music
+\item Originally written in Haskell
+\item Converted to Agda, using Haskell for MIDI interface
+\item Explore programming using dependent types in a circumscribed yet rich domain
+\item Use math, including transport of equivalences (from HoTT) and Ornaments
+\end{itemize}
+\end{frame}
+
 \begin{frame}\frametitle{Look vs Time (1997)}
@@
 
+\begin{frame}\frametitle{Equivalent Representations of Pitch}
+\ExecuteMetaData[../agda/latex/Pnwplse.tex]{pitch}
+\ExecuteMetaData[../agda/latex/Pnwplse.tex]{chromatic}
+\ExecuteMetaData[../agda/latex/Pnwplse.tex]{relpitch}
+\ExecuteMetaData[../agda/latex/Pnwplse.tex]{octave}
+\ExecuteMetaData[../agda/latex/Pnwplse.tex]{pitchoctave}
+\end{frame}
+
+\begin{frame}\frametitle{Equivalences}
+\begin{itemize}
+\item Define an equivalence between {\tt Pitch} and {\tt PitchOctave}
+\item Using HoTT techniques, automatically lift this equivalence to functions defined using \texttt{Pitch}
+\item See \textit{Equivalences for Free!} (Tabareau, Tanter, Sozeau)
+\item Challenge: Defining base equivalences. Can this be automated?
+\end{itemize}
+\end{frame}
+
+\begin{frame}[fragile]\frametitle{Ornaments}
+\begin{semiverbatim}
+data Music a  = ... 
+  |  Modify Control (Music a)
+
+data Control = ...
+  |  Phrase [PhraseAttribute]
+
+data PhraseAttribute = ...
+  |  Orn Ornament
+
+data Ornament =
+  Trill | Mordent | InvMordent | DoubleMordent |
+  Turn | TrilledTurn | ShortTrill ...
+\end{semiverbatim}
+\end{frame}
+
+\begin{frame}\frametitle{Ornaments}
+\begin{itemize}
+\item Functions on a base {\tt Music} structure can be automatically lifted to operate on {\tt Music} ornamented with additional information
+\item See works on Ornaments by McBride, Dagand and others
+\item Challenge: Shallow embedding of ornaments in Agda
+\end{itemize}
+\end{frame}
+
 \begin{frame}\frametitle{Conclusion}
-Goes here.
+\begin{itemize}
+\item Music is a good domain in which to explore  practical application of dependent types
+\item Using math can be more work at first, but should be a big win in the long term
+\item Figure out how to minimize the work and maximize the reward
+\end{itemize}
 \end{frame}

### 2019-05-08 doc/farm/farm-abstract.tex
added skelton for FARM abstract

%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
\documentclass[sigplan,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[acmsmall]{acmart}\settopmatter{}

%% Journal information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{CONF} % CONF = POPL or ICFP or OOPSLA
\acmArticle{1}
\acmYear{2018}
\acmMonth{1}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission;
%% use 'none' for review submission.
\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2018}           %% If different from \acmYear

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
\citestyle{acmauthoryear}   %% For author/year citations

%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption

\begin{document}

%% Title information
\title{Musical Ornaments}
\subtitle{Functional Pearl}

%% Author information
%% Contents and number of authors suppressed with 'anonymous'.

%% Author with single affiliation.
\author{John Leo}
\affiliation{
  \institution{Halfaya Research}            %% \institution is required
  \city{Bellevue}
  \state{WA}
  \country{USA}
}
\email{leo@halfaya.org}          %% \email is recommended

\author{Youyou Cong}
\affiliation{
  \institution{Tokyo Institute of Technology}            %% \institution is required
  \city{Tokyo}
  \country{Japan}
}
\email{cong@c.titech.ac.jp}          %% \email is recommended

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
Text of abstract \ldots.
\end{abstract}

%% 2012 ACM Computing Classification System (CSS) concepts
%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10011007.10011006.10011008</concept_id>
<concept_desc>Software and its engineering~General programming languages</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10003456.10003457.10003521.10003525</concept_id>
<concept_desc>Social and professional topics~History of programming languages</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~General programming languages}
\ccsdesc[300]{Social and professional topics~History of programming languages}
%% End of generated code

%% Keywords
%% comma separated list
\keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission

%% \maketitle
%% Note: \maketitle command must come after title commands, author
%% commands, abstract environment, Computing Classification System
%% environment and commands, and keywords command.
\maketitle

\section{Introduction}

Text of paper \ldots

%% Acknowledgments
\begin{acks}                            %% contents suppressed with 'anonymous'
\end{acks}

%% Bibliography
%\bibliography{bibfile}

\end{document}

### 2019-05-19 doc/farm/farm-abstract.tex
rough draft

@@
 %% Title information
-\title{Musical Ornaments}
-\subtitle{Functional Pearl}
+\title{Demo: Counterpoint by Construction}
+%\subtitle{Functional Pearl}
 
@@
 %% Author with single affiliation.
+\author{Youyou Cong}
+\affiliation{
+  \institution{Tokyo Institute of Technology}            %% \institution is required
+  \city{Tokyo}
+  \country{Japan}
+}
+\email{cong@c.titech.ac.jp}          %% \email is recommended
+
 \author{John Leo}
@@
 
-\author{Youyou Cong}
-\affiliation{
-  \institution{Tokyo Institute of Technology}            %% \institution is required
-  \city{Tokyo}
-  \country{Japan}
-}
-\email{cong@c.titech.ac.jp}          %% \email is recommended
-
 %% Abstract
@@
 \begin{abstract}
-Text of abstract \ldots.
+Western music of the common practice period tends to loosely follow sets of rules which
+were developed over time to help ensure the aethetic quality of the composition. Rules in
+developed in particular for harmony (\cite{piston1987harmony}) and counterpoint (\cite{fux1965study}),
+and continue to be taught to music students
+both to to understand the music of that time and because they continue to provide a foundation for
+modern art and popular music. To help both analyze and synthesize
+tonal music, it is worthwhile to attempt to encode these rules into a programming
+language, with functional programming languages being particularly wells-suited to this task.
+Recently quite a bit of work has been done in Haskell (\cite{fmmh, hihseufha, faamh, hafha, fghm})
+to encode the rules of harmony, but little work seems to have been done with counterpoint (\cite{mezzo}).
+Furthermore Haskell's type system, lacking full dependent types, is not strong enough to encode
+the required logic in a natural way, and requires using a number of extensions.
+
+In this demonstration of work in progress, we present Music Tools (\cite{MusicTools}), a library of
+small tools that can be combined functionally to help analyze and synthesize music. The library is
+written in the dependently typed language Agda (with Haskell used for I/O), which allows a simple
+and natural encoding of rules. We demonstrate its use in writing species counterpoint following
+Fux (\cite{fux1965study}), in which type system ensures by construction that the rules are followed.
+We show how this both aids human composition and allows for computer-generated creation of correct
+counterpoint. We also contrast this work to recent use of machine learnring to generated natural-sounding
+counterpoint which does not necessarily follow specific rules (\cite{CounterpointByConvolution}). Finally
+we indicate future planned work including extension to handle functional harmony.
 \end{abstract}
@@
 
-\section{Introduction}
-
-Text of paper \ldots
-
 %% Acknowledgments
 \begin{acks}                            %% contents suppressed with 'anonymous'
+  Thanks to the participants of the Tokyo Agda Implmentors' Meeting, especially Ulf Norell and Jesper Cockx,
+  for many helpful suggestions that improved our Agda code.
 \end{acks}
@@
 %% Bibliography
-%\bibliography{bibfile}
+\bibliography{farm-abstract.bib}
 

### 2019-05-23 doc/farm/farm-abstract.tex
tweaked abstract

--- 2019-05-08 doc/farm/farm-abstract.tex
@@
 %% before \maketitle command
-\begin{abstract}
-Text of abstract \ldots.
-\end{abstract}
 
@@
 %% comma separated list
-\keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission
+% \keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission
 
@@
 
-\section{Introduction}
+Western music of the common practice period tends to loosely
+follow sets of rules, which were developed over time to help
+ensure the aethetic quality of the composition. 
+Among various rules, those for harmony \citep{piston1987harmony}
+and counterpoint \citep{fux1965study} are continue to be taught
+to music students, not only as a means to understand the music of
+that period, but also as a foundation for modern art and popular music.
+To help analyze and synthesize tonal music, it is worthwhile to encode
+these rules into a programming language, and in particular, functional
+programming languages have proven ideally suited to this task.
+In the past decade, quite a bit of work has been done in Haskell
+\citep{fmmh, hihseufha, faamh, hafha, fghm} to encode the rules of harmony,
+but little progress has been made on counterpoint \citep{mezzo}.
+Furthermore, due to the lack of full dependent types, Haskell's
+type system is not strong enough to encode the logic of counterpoint
+in a natural way; indeed, existing implementation relies on a number
+of language extensions.
 
-Text of paper \ldots
+In this demonstration of work in progress, we present Music Tools
+\citep{MusicTools}, a library of small tools that can be combined
+functionally to help analyze and synthesize music.
+The library is written in Agda (with Haskell used for I/O), which
+allows simple and natural encoding of rules by means of dependent types.
+As an application of the library, we demonstrate an implementation of 
+species counterpoint \citep{fux1965study}, which uses Agda's rich type
+system to ensure (by construction) that all rules are followed.
+We show how this both aids human composition and allows for
+computer-generated creation of correct counterpoint.
+We also contrast our approach to a recent study on generating
+natural-sounding counterpoint by machine learning
+\citep{CounterpointByConvolution}, which does not give us correctness
+guarantees.
+Finally, we discuss further applications of our library, including handling 
+of functional harmony.
 
@@
 \begin{acks}                            %% contents suppressed with 'anonymous'
+ The authors would like to thank the participants of the Tokyo Agda 
+ Implmentors' Meeting, especially Ulf Norell and Jesper Cockx,
+ for many helpful suggestions that improved our Agda code.
 \end{acks}

### 2019-05-24 doc/farm/farm-abstract.tex
more tweaks

--- 2019-05-08 doc/farm/farm-abstract.tex
@@
 
-\section{Introduction}
+Western music of the common practice period tends to loosely
+follow sets of rules, which were developed over time to ensure
+the aethetic quality of the composition. 
+Among various rules, those for harmony \citep{piston1987harmony}
+and counterpoint \citep{fux1965study} (a technique for generating
+multi-voice melodies) are continue to be taught to music students,
+not only as a means to understand the music of that period, but
+also as a foundation for modern art and popular music.
 
-Text of paper \ldots
+To help analyze and synthesize tonal music, it is worthwhile to
+encode these rules into a programming language, and as shown by
+recent studies, functional programming languages are particularly
+suited to this task.
+In the past decade, Haskell has been extensively used to encode
+the rules of harmony \citep{fmmh, hihseufha, faamh, hafha, fghm}
+as well as counterpoint \citep{szamozvancev2017well}.
+Since Haskell is a statically typed language, these frameworks 
+often come with a form of type safety property, namely
+``well-typed music does not sound wrong''.
+Unfortunately, it turns out that the type system of plain Haskell
+is sometimes not powerful enough to guarantee this property.
+To avoid generating music that ``sounds wrong'', previous studies
+have incorporated various language extensions, such as GADTs 
+\citep{cheney2002lightweight} and singleton types 
+\citep{eisenberg2013dependently}, to enable dependently-typed 
+programming.
+
+In this demonstration of work in progress, we present Music Tools
+\citep{MusicTools}, a library of small tools that can be combined
+functionally to help analyze and synthesize music.
+To allow simple and natural encoding of rules, we built the library
+in Agda \citep{norellphd}, which is a functional language with full
+dependent types.
+As an application of the library, we demonstrate an implementation of 
+species counterpoint, based on the rules given by \citet{fux1965study}.
+Thanks to Agda's rich type system, we can ensure by construction that
+well-typed counterpoint satisfies all the required rules.
+We show how the type-based approach both aids human composition and
+allows for computer-generated creation of correct counterpoint.
+We also contrast our work to a recent study on generating
+natural-sounding counterpoint by machine learning
+\citep{CounterpointByConvolution}, which does not give us correctness
+guarantees.
+Finally, we discuss further applications of our library, including
+handling of functional harmony.
 

### 2019-05-25 doc/farm/farm-abstract.tex
added some rules of first species counterpoint

@@
 species counterpoint, based on the rules given by \citet{fux1965study}.
-Thanks to Agda's rich type system, we can ensure by construction that
+To give a rough idea of what the rules look like, here are some of
+the requirements for composing two-part, first-species counterpoint:
+
+\begin{itemize}
+\item The interval between the two voices should not be a perfect fourth
+\item The two voices should not move to a perfect fifth or octave 
+by similar motion
+\item The counterpoint should end with a proper cadence
+\end{itemize}
+
+\noindent Thanks to Agda's rich type system, we can fully express 
+these rules using types, and thus ensure by construction that
 well-typed counterpoint satisfies all the required rules.
-We show how the type-based approach both aids human composition and
-allows for computer-generated creation of correct counterpoint.
-We also contrast our work to a recent study on generating
+
+At the FARM workshop, we intend to give a gentle introduction to
+counterpoint, and describe our Agda implementation, showing how the
+type-based approach both aids human composition and allows for 
+computer-generated creation of correct counterpoint.
+We then contrast our work to a recent study on generating
 natural-sounding counterpoint by machine learning

### 2019-05-25 doc/farm/farm-abstract.tex
add code examples

@@
                         %% http://ctan.org/pkg/subcaption
+\usepackage{listings}
 
@@
 follow sets of rules, which were developed over time to ensure
-the aethetic quality of the composition. 
-Among various rules, those for harmony \citep{piston1987harmony}
-and counterpoint \citep{fux1965study} (a technique for generating
-multi-voice melodies) are continue to be taught to music students,
+the aesthetic quality of the composition. 
+Among these rules, those for harmony \citep{piston1987harmony}
+and counterpoint (harmonically interdependent
+melodies) \citep{fux1965study} are particularly fundamental and continue
+to be taught to music students,
 not only as a means to understand the music of that period, but
@@
 To help analyze and synthesize tonal music, it is worthwhile to
-encode these rules into a programming language, and as shown by
+encode these rules into a programming language. As shown by
 recent studies, functional programming languages are particularly
@@
 as well as counterpoint \citep{szamozvancev2017well}.
-Since Haskell is a statically typed language, these frameworks 
-often come with a form of type safety property, namely
+Haskell's static type sytem is used to encode these rules, so that
 ``well-typed music does not sound wrong''.
-Unfortunately, it turns out that the type system of plain Haskell
-is sometimes not powerful enough to guarantee this property.
-To avoid generating music that ``sounds wrong'', previous studies
-have incorporated various language extensions, such as GADTs 
+Unfortunately the type system of plain Haskell
+is not powerful enough to guarantee these properties. Previous studies
+have thus incorporated language extensions such as GADTs 
 \citep{cheney2002lightweight} and singleton types 
-\citep{eisenberg2013dependently}, to enable dependently-typed 
+\citep{eisenberg2013dependently} to approximate dependently-typed 
 programming.
@@
 species counterpoint, based on the rules given by \citet{fux1965study}.
-To give a rough idea of what the rules look like, here are some of
-the requirements for composing two-part, first-species counterpoint:
+Thanks to Agda's rich type system, we can express 
+these rules naturally, and thus ensure by construction that
+well-typed counterpoint satisfies all the required rules.
 
-\begin{itemize}
-\item The interval between the two voices should not be a perfect fourth
-\item The two voices should not move to a perfect fifth or octave 
-by similar motion
-\item The counterpoint should end with a proper cadence
-\end{itemize}
+As an example, the simplest of Fux's rule systems is first-species
+counterpoint, in which one starts with a base melody (the \textit{cantus firmus})
+and contructs a counterpoint melody note-by-note in the same rhythm.
+``Dissonant'' intervals (2nds, 7ths and 4ths) are probibited, as are
+are parallel and similar motion of perfect intervals (5ths and octaves).
+The latter is encoded in Agda as follows.
 
-\noindent Thanks to Agda's rich type system, we can fully express 
-these rules using types, and thus ensure by construction that
-well-typed counterpoint satisfies all the required rules.
+\begin{lstlisting}
+motionOk : (i1 : Interval)
+           (i2 : Interval) → Set
+motionOk i1 i2 with motion i1 i2
+         | isPerfectInterval i2
+motionOk i1 i2 | contrary | _     = ⊤
+motionOk i1 i2 | oblique  | _     = ⊤
+motionOk i1 i2 | parallel | false = ⊤
+motionOk i1 i2 | parallel | true  = ⊥
+motionOk i1 i2 | similar  | false = ⊤
+motionOk i1 i2 | similar  | true  = ⊥
+\end{lstlisting}
+
+The music must also end with a cadence, which is a final motion from
+the second or seventh degree to the tonic (first degree). These rules
+are encoded in the following type, whose terms are exactly valid
+first-species counterpoint.
+
+\begin{lstlisting}
+data FirstSpecies : PitchInterval →
+                    Set where
+  cadence2 : (p : Pitch) →
+    FirstSpecies
+      (transpose (+ 2) p , maj6)
+  cadence7 : (p : Pitch) →
+    FirstSpecies
+      (transpose -[1+ 0 ] p , min10)
+  _::_ : (pi : PitchInterval) →
+         {pj : PitchInterval} →
+         {_ : motionOk pi pj} →
+         FirstSpecies pj →
+         FirstSpecies pi
+\end{lstlisting}
+
+Explicit conversions from \texttt{PitchInterval} (which ensures
+the inteval is not dissonant) to the general \texttt{Interval} have
+been omitted in the listing above.
+Note that \texttt{motionOk} is an implicit argument of the cons
+constructor, and is able to be resolved automatically by the type checker,
+and thus the proof does not need to be supplied by the user. Valid
+first-species counterpoint can thus be written very naturally as
+in the following simple example.
+
+\begin{lstlisting}
+example : FirstSpecies (g 4 , per8)
+example = 
+  (g 4 , per8) :: (c 5 , maj10) ::
+  (c 5 , per8) :: (c 5 , maj10) ::
+  (e 5 , min10) :: (g 5 , per8) ::
+  (cadence2 (c 6))
+\end{lstlisting}
 
@@
 natural-sounding counterpoint by machine learning
-\citep{CounterpointByConvolution}, which does not give us correctness
+\citep{CounterpointByConvolution}, which does not provide correctness
 guarantees.
 Finally, we discuss further applications of our library, including
-handling of functional harmony.
+representation of functional harmony.
 
@@
  The authors would like to thank the participants of the Tokyo Agda 
- Implmentors' Meeting, especially Ulf Norell and Jesper Cockx,
+ Implementors' Meeting, especially Ulf Norell and Jesper Cockx,
  for many helpful suggestions that improved our Agda code.

### 2019-05-26 doc/farm/farm-abstract.tex
typo
added some more code

@@
 
-As an example, the simplest of Fux's rule systems is first-species
-counterpoint, in which one starts with a base melody (the \textit{cantus firmus})
-and contructs a counterpoint melody note-by-note in the same rhythm.
-``Dissonant'' intervals (2nds, 7ths and 4ths) are probibited, as are
-are parallel and similar motion of perfect intervals (5ths and octaves).
-The latter is encoded in Agda as follows.
+Let us briefly explain how to implement the rule system of 
+first-species counterpoint\footnote{The code is available at
+\url{https://github.com/halfaya/MusicTools/blob/master/agda/Counterpoint.agda}.}.
+In this exercise, we pick a base melody (the \emph{cantus firmus}),
+and contruct a counterpoint melody note-by-note in the same rhythm.
+The two voices are represented as a list of pitch-interval pairs,
+where intervals must not be dissonant ones (2nds, 7ths, or 4ths).
+
+\begin{lstlisting}
+data IntervalQuality : Set where
+  min3  : IntervalQuality
+  maj3  : IntervalQuality
+  per5  : IntervalQuality
+  min6  : IntervalQuality
+  maj6  : IntervalQuality
+  per8  : IntervalQuality
+  min10 : IntervalQuality
+  maj10 : IntervalQuality
+
+PitchInterval : Set
+PitchInterval = Pitch × IntervalQuality
+\end{lstlisting}
+
+In a counterpoint list, there must not be perfect intervals 
+(5ths and octaves) resulted from parallel or similar motion.
+Therefore, we define a predicate that checks whether a motion 
+is allowed or not.
 
@@
 
-The music must also end with a cadence, which is a final motion from
-the second or seventh degree to the tonic (first degree). These rules
-are encoded in the following type, whose terms are exactly valid
-first-species counterpoint.
+The last requirement is that the music must end with a cadence,
+ which is a final motion from the 2nd or 7th degree to the tonic 
+(1st degree). 
+We impose this requirement by declaring two cadence constructors 
+as the base cases of counterpoint.
+Thus, we arrive at the following datatype for well-typed counterpoint\footnote{
+For readability, we have ommited explicit conversions 
+from \texttt{PitchInterval} (which ensures the interval is not dissonant) 
+to the general \texttt{Interval}.}.
 
@@
 
-Explicit conversions from \texttt{PitchInterval} (which ensures
-the inteval is not dissonant) to the general \texttt{Interval} have
-been omitted in the listing above.
-Note that \texttt{motionOk} is an implicit argument of the cons
-constructor, and is able to be resolved automatically by the type checker,
-and thus the proof does not need to be supplied by the user. Valid
-first-species counterpoint can thus be written very naturally as
-in the following simple example.
+\noindent Observe that \texttt{motionOk} is an implicit argument of 
+the \texttt{\_::\_} constructor. 
+The argument can be resolved automatically by the type checker,
+hence there is no need to manually supply this proof.
+
+Now, we can write valid first-species counterpoint as in the 
+example below.
 

### 2019-05-26 doc/farm/farm-abstract.tex
adjust some working

@@
 \url{https://github.com/halfaya/MusicTools/blob/master/agda/Counterpoint.agda}.}.
-In this exercise, we pick a base melody (the \emph{cantus firmus}),
-and contruct a counterpoint melody note-by-note in the same rhythm.
+In first-species counterpoint, one starts with a base melody (the \emph{cantus firmus}),
+and contructs a counterpoint melody note-by-note in the same rhythm.
 The two voices are represented as a list of pitch-interval pairs,
-where intervals must not be dissonant ones (2nds, 7ths, or 4ths).
+where intervals must not be dissonant (2nds, 7ths, or 4ths).
 
@@
 
-In a counterpoint list, there must not be perfect intervals 
-(5ths and octaves) resulted from parallel or similar motion.
+In addition, it is prohibited to move from any interval to
+a perfect interval (5th or octave) via parallel or similar motion.
 Therefore, we define a predicate that checks whether a motion 
@@
 
-Now, we can write valid first-species counterpoint as in the 
+Now we can write valid first-species counterpoint as in the 
 example below.

### 2019-06-21 doc/farm/farm-abstract.tex
typewriter font
reformat code
preparing camera-ready

@@
 %% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
-\documentclass[sigplan,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+\documentclass[sigplan,10pt,screen]{acmart}
+\settopmatter{printfolios=true, printccs=true, printacmref=false}
 %% For double-blind review submission, w/ CCS and ACM Reference
@@
 
-%% Journal information
-%% Supplied to authors by publisher for camera-ready submission;
-%% use defaults for review submission.
-\acmJournal{PACMPL}
-\acmVolume{1}
-\acmNumber{CONF} % CONF = POPL or ICFP or OOPSLA
-\acmArticle{1}
-\acmYear{2018}
-\acmMonth{1}
+\acmConference[FARM 2019]{The ACM SIGPLAN International Workshop on Functional Art, Music, Modelling and Design}{August, 2019}{Berlin, Germany}
+\acmYear{2019}
+\acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
 \acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
@@
                         %% http://ctan.org/pkg/subcaption
-\usepackage{listings}
+\usepackage{alltt}
 
@@
 %% before \maketitle command
+\begin{abstract}
+We present \emph{Music Tools}, an Agda library for analyzing 
+and synthesizing music.  
+The library uses dependent types to simplify encoding of 
+music rules, thus improving existing approaches based on 
+simply typed languages.  
+As an application of the library, we demonstrate an 
+implementation of first-species counterpoint, where we use
+dependent types to constrain the motion of two parallely
+sounding voices.
+\end{abstract}
 
@@
 <concept>
-<concept_id>10011007.10011006.10011008</concept_id>
-<concept_desc>Software and its engineering~General programming languages</concept_desc>
+<concept_id>10010405.10010469.10010475</concept_id>
+<concept_desc>Applied computing~Sound and music computing</concept_desc>
 <concept_significance>500</concept_significance>
@@
 <concept>
-<concept_id>10003456.10003457.10003521.10003525</concept_id>
-<concept_desc>Social and professional topics~History of programming languages</concept_desc>
-<concept_significance>300</concept_significance>
+<concept_id>10011007.10011006.10011008.10011009.10011012</concept_id>
+<concept_desc>Software and its engineering~Functional languages</concept_desc>
+<concept_significance>500</concept_significance>
 </concept>
@@
 
-\ccsdesc[500]{Software and its engineering~General programming languages}
-\ccsdesc[300]{Social and professional topics~History of programming languages}
+\ccsdesc[500]{Applied computing~Sound and music computing}
+\ccsdesc[500]{Software and its engineering~Functional languages}
 %% End of generated code
@@
 
+\section{Introduction}
+
 Western music of the common practice period tends to loosely
@@
 the aesthetic quality of the composition. 
-Among these rules, those for harmony \citep{piston1987harmony}
-and counterpoint (harmonically interdependent
-melodies) \citep{fux1965study} are particularly fundamental and continue
-to be taught to music students,
-not only as a means to understand the music of that period, but
-also as a foundation for modern art and popular music.
+Among such rules, those for harmony \citep{piston1987harmony}
+and counterpoint (harmonically independent melodies) 
+\citep{fux1965study} are particularly fundamental and continue 
+to be taught to music students, not only as a means to understand 
+the music of that period, but also as a foundation for modern art 
+and popular music.
 
-To help analyze and synthesize tonal music, it is worthwhile to
-encode these rules into a programming language. As shown by
-recent studies, functional programming languages are particularly
-suited to this task.
+To analyze and synthesize tonal music, researchers have been
+attempting to encode various rules into a programming language.
+It has proven that functional programming languages are
+ideally suited for this task, and in particular, those with a static
+type system can further guarantee that \emph{well-typed music 
+does not sound wrong}.
 In the past decade, Haskell has been extensively used to encode
@@
 as well as counterpoint \citep{szamozvancev2017well}.
-Haskell's static type sytem is used to encode these rules, so that
-``well-typed music does not sound wrong''.
-Unfortunately the type system of plain Haskell
-is not powerful enough to guarantee these properties. Previous studies
-have thus incorporated language extensions such as GADTs 
-\citep{cheney2002lightweight} and singleton types 
-\citep{eisenberg2013dependently} to approximate dependently-typed 
-programming.
+An interesting observation is that, many of the existing encodings
+rely on some form of \emph{dependent types}, \textit{i.e.},  
+types that depend on terms. 
+Since Haskell is not a dependently typed language, one has to
+use other facilities, such as GADTs \citep{cheney2002lightweight} 
+and singleton types \citep{eisenberg2013dependently},
+to indirectly simulate dependencies.
+While this allows to encode a wide class of music rules, it may
+require duplicating certain pieces of code to reflect terms into 
+types, making the implementation less elegant 
+\citep{monnier2010singleton}.
+This motivates us to explore music programming in a language
+with intrinsic support for dependent types.
 
-In this demonstration of work in progress, we present Music Tools
-\citep{MusicTools}, a library of small tools that can be combined
-functionally to help analyze and synthesize music.
-To allow simple and natural encoding of rules, we built the library
-in Agda \citep{norellphd}, which is a functional language with full
-dependent types.
-As an application of the library, we demonstrate an implementation of 
-species counterpoint, based on the rules given by \citet{fux1965study}.
+We present Music Tools\footnote{\url{https://github.com/halfaya/MusicTools}}, a library of small tools that can be combined functionally to 
+help analyze and synthesize music.
+To allow simple and natural encoding of rules, we build our 
+library in the Agda \citep{norellphd} proof assistant, where
+dependent types are readily available.
+As an application of the library, we demonstrate an
+implementation  of  species counterpoint, based on the rules 
+given by \citet{fux1965study}.
 Thanks to Agda's rich type system, we can express 
-these rules naturally, and thus ensure by construction that
-well-typed counterpoint satisfies all the required rules.
+the rules in a straightforward manner, and thus ensure by 
+construction that well-typed counterpoint satisfies all the 
+required rules.
 
-Let us briefly explain how to implement the rule system of 
-first-species counterpoint\footnote{The code is available at
+\section{The Music Tools Library}
+
+TODO: describe the key ingredients of the library.
+
+\section{Application: First-Species Counterpoint}
+
+Now, let us explain how to implement the rule system of 
+first-species counterpoint\footnote{The code is available at \\
 \url{https://github.com/halfaya/MusicTools/blob/master/agda/Counterpoint.agda}.}.
-In first-species counterpoint, one starts with a base melody (the \emph{cantus firmus}),
-and contructs a counterpoint melody note-by-note in the same rhythm.
+In first-species counterpoint, one starts with a base melody 
+(the \emph{cantus firmus}), and constructs a counterpoint 
+melody note-by-note in the same rhythm.
 The two voices are represented as a list of pitch-interval pairs,
@@
 
-\begin{lstlisting}
+\begin{alltt}
 data IntervalQuality : Set where
@@
 PitchInterval : Set
-PitchInterval = Pitch × IntervalQuality
-\end{lstlisting}
+PitchInterval = Pitch \(\times\) IntervalQuality
+\end{alltt}
 
@@
 
-\begin{lstlisting}
+\begin{alltt}
 motionOk : (i1 : Interval)
-           (i2 : Interval) → Set
+           (i2 : Interval) \(\rightarrow\) Set
 motionOk i1 i2 with motion i1 i2
          | isPerfectInterval i2
-motionOk i1 i2 | contrary | _     = ⊤
-motionOk i1 i2 | oblique  | _     = ⊤
-motionOk i1 i2 | parallel | false = ⊤
-motionOk i1 i2 | parallel | true  = ⊥
-motionOk i1 i2 | similar  | false = ⊤
-motionOk i1 i2 | similar  | true  = ⊥
-\end{lstlisting}
+motionOk i1 i2 | contrary | \_     = \(\top\)
+motionOk i1 i2 | oblique  | \_     = \(\top\)
+motionOk i1 i2 | parallel | false = \(\top\)
+motionOk i1 i2 | parallel | true  = \(\bot\)
+motionOk i1 i2 | similar  | false = \(\top\)
+motionOk i1 i2 | similar  | true  = \(\bot\)
+\end{alltt}
 
@@
 
-\begin{lstlisting}
-data FirstSpecies : PitchInterval →
+\begin{alltt}
+data FirstSpecies : PitchInterval \(\rightarrow\)
                     Set where
-  cadence2 : (p : Pitch) →
-    FirstSpecies
-      (transpose (+ 2) p , maj6)
-  cadence7 : (p : Pitch) →
-    FirstSpecies
-      (transpose -[1+ 0 ] p , min10)
-  _::_ : (pi : PitchInterval) →
-         {pj : PitchInterval} →
-         {_ : motionOk pi pj} →
-         FirstSpecies pj →
+  cadence2 : (p : Pitch) \(\rightarrow\)
+    FirstSpecies (transpose (+ 2) p , maj6)
+  cadence7 : (p : Pitch) \(\rightarrow\)
+    FirstSpecies (transpose -[1+ 0 ] p , min10)
+  \_::\_ : (pi : PitchInterval) \(\rightarrow\)
+         {pj : PitchInterval} \(\rightarrow\)
+         {\_ : motionOk pi pj} \(\rightarrow\)
+         FirstSpecies pj \(\rightarrow\)
          FirstSpecies pi
-\end{lstlisting}
+\end{alltt}
 
@@
 
-\begin{lstlisting}
+\begin{alltt}
 example : FirstSpecies (g 4 , per8)
@@
   (cadence2 (c 6))
-\end{lstlisting}
+\end{alltt}
 
-At the FARM workshop, we intend to give a gentle introduction to
-counterpoint, and describe our Agda implementation, showing how the
-type-based approach both aids human composition and allows for 
-computer-generated creation of correct counterpoint.
-We then contrast our work to a recent study on generating
-natural-sounding counterpoint by machine learning
-\citep{CounterpointByConvolution}, which does not provide correctness
-guarantees.
-Finally, we discuss further applications of our library, including
-representation of functional harmony.
+\section{Future Work}
+
+TODO
+
+%At the FARM workshop, we intend to give a gentle introduction to
+%counterpoint, and describe our Agda implementation, showing how the
+%type-based approach both aids human composition and allows for 
+%computer-generated creation of correct counterpoint.
+%We then contrast our work to a recent study on generating
+%natural-sounding counterpoint by machine learning
+%\citep{CounterpointByConvolution}, which does not provide correctness
+%guarantees.
+%Finally, we discuss further applications of our library, including
+%representation of functional harmony.
 

### 2019-06-21 doc/farm/farm-abstract.tex
fixed a typo

@@
 Thus, we arrive at the following datatype for well-typed counterpoint\footnote{
-For readability, we have ommited explicit conversions 
+For readability, we have omitted explicit conversions 
 from \texttt{PitchInterval} (which ensures the interval is not dissonant) 

### 2019-06-24 doc/farm/farm-abstract.tex
minor fixes

@@
 Among such rules, those for harmony \citep{piston1987harmony}
-and counterpoint (harmonically independent melodies) 
+and counterpoint (harmonically interdependent melodies) 
 \citep{fux1965study} are particularly fundamental and continue 
@@
 
-To analyze and synthesize tonal music, researchers have been
-attempting to encode various rules into a programming language.
-It has proven that functional programming languages are
+To analyze and synthesize tonal music, researchers have
+attempted to encode these rules into programming languages.
+Functional programming languages seem
 ideally suited for this task, and in particular, those with a static
@@
 An interesting observation is that, many of the existing encodings
-rely on some form of \emph{dependent types}, \textit{i.e.},  
+rely on some form of \emph{dependent types}, i.e.,  
 types that depend on terms. 
 Since Haskell is not a dependently typed language, one has to
-use other facilities, such as GADTs \citep{cheney2002lightweight} 
+use language extensions, such as GADTs \citep{cheney2002lightweight} 
 and singleton types \citep{eisenberg2013dependently},
-to indirectly simulate dependencies.
-While this allows to encode a wide class of music rules, it may
-require duplicating certain pieces of code to reflect terms into 
+to simulate dependencies.
+While this allows encoding a wide class of music rules, it can
+require duplicating code to reflect terms into 
 types, making the implementation less elegant 
@@
 
-We present Music Tools\footnote{\url{https://github.com/halfaya/MusicTools}}, a library of small tools that can be combined functionally to 
+We present Music Tools\footnote{\url{https://github.com/halfaya/MusicTools}},
+a library of small tools that can be combined functionally to 
 help analyze and synthesize music.
 To allow simple and natural encoding of rules, we build our 
-library in the Agda \citep{norellphd} proof assistant, where
-dependent types are readily available.
+library in Agda \citep{norellphd}, a functional language with full dependent types.
 As an application of the library, we demonstrate an
@@
 
-Now, let us explain how to implement the rule system of 
+We now explain how to implement the rule system of 
 first-species counterpoint\footnote{The code is available at \\
@@
   \_::\_ : (pi : PitchInterval) \(\rightarrow\)
-         {pj : PitchInterval} \(\rightarrow\)
-         {\_ : motionOk pi pj} \(\rightarrow\)
+         \{pj : PitchInterval\} \(\rightarrow\)
+         \{\_ : motionOk pi pj\} \(\rightarrow\)
          FirstSpecies pj \(\rightarrow\)

### 2019-06-24 doc/farm/farm-abstract.tex
clarify cadence

@@
 We impose this requirement by declaring two cadence constructors 
-as the base cases of counterpoint.
+as the base cases of counterpoint (note that the final interval of
+the cadence is always \texttt{(p, per8)} and is thus not explicitly
+specified).
 Thus, we arrive at the following datatype for well-typed counterpoint\footnote{

### 2019-06-24 doc/farm/farm-abstract.tex
library overview and future work

@@
 
-TODO: describe the key ingredients of the library.
+The goal of Music Tools is to provide a core collection of types and functions
+that can be used to easily build applications to analyze and synthesize music.
+It is intended to evolved into a dependently-typed replacement for Euterpea
+\citep{hudak2018haskell} and borrows many ideas from that library.
+It is currently restricted to the chromatic scale for simplicity, and like
+Euterpea is designed to work well with MIDI.
+
+The fundamental type for melody is \texttt{Pitch} which is just a natural number with $0$
+representing the lowest expressible pitch (it is intended to correspond to MIDI note 0, but
+the interpretation can change depending upon the application). One can also
+express pitch as a pair of an octave and relative pitch within the ocatve, which is
+more convenient for some applications, and convert between this representation and
+absolute pitch. One can then prove that converting back and forth is the identity function.
+
+For rhythm the fundamental type is \texttt{Duration}, also a natural number, which represents
+some unspecified unit of time (when the music is played the unit is then specified).
+A combination of a pitch and a duration gives a \texttt{Note}, and notes in turn are
+made into \texttt{Music} via sequential and parallel composition, as in Euterpea.
+
+Music in turn can be converted to MIDI and and played on synthesizers. Existing Haskell
+MIDI libraries are called via Agda's Haskell FFI. Details can be found in the github
+repository.
 
@@
 
-TODO
+We intend to expand the Music Tools library to include
+richer functionality for analysis and synthesis. We are
+particularly interested in expressing the work done in Haskell
+on functional harmony in the library, to see just how much
+dependent types can simplify the representation.
 
-%At the FARM workshop, we intend to give a gentle introduction to
-%counterpoint, and describe our Agda implementation, showing how the
-%type-based approach both aids human composition and allows for 
-%computer-generated creation of correct counterpoint.
-%We then contrast our work to a recent study on generating
-%natural-sounding counterpoint by machine learning
-%\citep{CounterpointByConvolution}, which does not provide correctness
-%guarantees.
-%Finally, we discuss further applications of our library, including
-%representation of functional harmony.
+Music is a rich yet circumscribed domain in which
+issues of equivalence \citep{tabareau2017equivalences}
+and ornamentation \citep{dagand2017essence} naturally arise.
+For example a
+musical score can be interpreted either horizontally (counterpoint) or
+vertically (harmony) and it is important to be able to seamlessly
+convert between these representations. Similarly one may wish to treat
+pitch or rhythm separately as well as combine them, which is a natural
+application of ornamentation. We plan to examine the extent to which this
+research can be put to use in a practical domain.
+
+For counterpoint specifically, we plan represent higher species counterpoint
+and also to explore automatic generation of species counterpoint,
+contrasting the music created to recent work on generating
+natural-sounding counterpoint by machine learning
+\citep{CounterpointByConvolution}, which does not provide correctness
+guarantees.
 
@@
  for many helpful suggestions that improved our Agda code.
+ We also thank the anonymous reviewers for their thoughtful feedback
+ which improved our presentation.
 \end{acks}

### 2019-06-25 doc/farm/farm-abstract.tex
tweaks

@@
 that can be used to easily build applications to analyze and synthesize music.
-It is intended to evolved into a dependently-typed replacement for Euterpea
+It is intended to evolve into a dependently-typed replacement for Euterpea
 \citep{hudak2018haskell} and borrows many ideas from that library.
@@
 
-The fundamental type for melody is \texttt{Pitch} which is just a natural number with $0$
+The fundamental type for melody is \texttt{Pitch}, which is just a natural number with $0$
 representing the lowest expressible pitch (it is intended to correspond to MIDI note 0, but
 the interpretation can change depending upon the application). One can also
-express pitch as a pair of an octave and relative pitch within the ocatve, which is
+express pitch as a pair of an octave and relative pitch within the octave, which is
 more convenient for some applications, and convert between this representation and
@@
 
-Music in turn can be converted to MIDI and and played on synthesizers. Existing Haskell
-MIDI libraries are called via Agda's Haskell FFI. Details can be found in the github
-repository.
+To play music on synthesizers, one can convert it to MIDI by calling 
+Haskell MIDI libraries via Agda's Haskell FFI. 
+Details can be found in the github repository.
 
@@
 and ornamentation \citep{dagand2017essence} naturally arise.
-For example a
+For example, a
 musical score can be interpreted either horizontally (counterpoint) or
-vertically (harmony) and it is important to be able to seamlessly
-convert between these representations. Similarly one may wish to treat
+vertically (harmony), and it is important to be able to seamlessly
+convert between these representations. Similarly, one may wish to treat
 pitch or rhythm separately as well as combine them, which is a natural
@@
 
-For counterpoint specifically, we plan represent higher species counterpoint
-and also to explore automatic generation of species counterpoint,
-contrasting the music created to recent work on generating
-natural-sounding counterpoint by machine learning
-\citep{CounterpointByConvolution}, which does not provide correctness
-guarantees.
+For counterpoint specifically, we plan to represent higher species 
+counterpoint by extending the rules, 
+and to explore automatic generation of species counterpoint.
+It would be interesting to compare our correct-by-construction counterpoint
+and one created by machine learning \citep{CounterpointByConvolution}, 
+which does not have correctness guarantees.
 

### 2019-06-25 doc/farm/farm-abstract.tex
minor fixes

@@
 implementation of first-species counterpoint, where we use
-dependent types to constrain the motion of two parallely
-sounding voices.
+dependent types to constrain the motion of two parallel
+voices.
 \end{abstract}
@@
 as the base cases of counterpoint (note that the final interval of
-the cadence is always \texttt{(p, per8)} and is thus not explicitly
+the cadence is always \texttt{(p , per8)} and is thus not explicitly
 specified).
@@
 It would be interesting to compare our correct-by-construction counterpoint
-and one created by machine learning \citep{CounterpointByConvolution}, 
+with that created by machine learning \citep{CounterpointByConvolution}, 
 which does not have correctness guarantees.

### 2019-06-28 doc/farm/farm-abstract.tex
camera-ready

@@
 \documentclass[sigplan,10pt,screen]{acmart}
-\settopmatter{printfolios=true, printccs=true, printacmref=false}
+\settopmatter{printfolios=true, printccs=true, printacmref=true}
 %% For double-blind review submission, w/ CCS and ACM Reference
@@
 
-\acmConference[FARM 2019]{The ACM SIGPLAN International Workshop on Functional Art, Music, Modelling and Design}{August, 2019}{Berlin, Germany}
+%%% The following is specific to FARM '19 and the paper
+%%% 'Demo: Counterpoint by Construction'
+%%% by Youyou Cong and John Leo.
+%%%
+\setcopyright{rightsretained}
+\acmPrice{}
+\acmDOI{10.1145/3331543.3342578}
 \acmYear{2019}
-\acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
-\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
-\startPage{1}
-
-%% Copyright information
-%% Supplied to authors (based on authors' rights management selection;
-%% see authors.acm.org) by publisher for camera-ready submission;
-%% use 'none' for review submission.
-\setcopyright{none}
-%\setcopyright{acmcopyright}
-%\setcopyright{acmlicensed}
-%\setcopyright{rightsretained}
-%\copyrightyear{2018}           %% If different from \acmYear
+\copyrightyear{2019}
+\acmISBN{978-1-4503-6811-7/19/08}
+\acmConference[FARM '19]{Proceedings of the 7th ACM SIGPLAN International Workshop on Functional Art, Music, Modeling, and Design}{August 23, 2019}{Berlin, Germany}
+\acmBooktitle{Proceedings of the 7th ACM SIGPLAN International Workshop on Functional Art, Music, Modeling, and Design (FARM '19), August 23, 2019, Berlin, Germany}
 
@@
 does not sound wrong}.
-In the past decade, Haskell has been extensively used to encode
+In the past decade, Haskell has been a popular choice for encoding
 the rules of harmony \citep{fmmh, hihseufha, faamh, hafha, fghm}
@@
 types that depend on terms. 
-Since Haskell is not a dependently typed language, one has to
-use language extensions, such as GADTs \citep{cheney2002lightweight} 
-and singleton types \citep{eisenberg2013dependently},
-to simulate dependencies.
-While this allows encoding a wide class of music rules, it can
-require duplicating code to reflect terms into 
-types, making the implementation less elegant 
-\citep{monnier2010singleton}.
+While Haskell is not a dependently typed language, it has various 
+language extensions (such as GADTs \citep{cheney2002lightweight} 
+and singleton types \citep{eisenberg2013dependently}) for 
+simulating dependent types, and as shown by previous studies,
+the extended type system is expressive enough for 
+a wide class of music applications.
+On the other hand, the need for simulating dependent types 
+can also result in code duplication, making the implementation 
+less elegant \citep{monnier2010singleton}.
 This motivates us to explore music programming in a language
@@
 It is intended to evolve into a dependently-typed replacement for Euterpea
-\citep{hudak2018haskell} and borrows many ideas from that library.
-It is currently restricted to the chromatic scale for simplicity, and like
-Euterpea is designed to work well with MIDI.
+\citep{hudak2018haskell}, and borrows many ideas from that library.
+Currently, Music Tools is restricted to the chromatic scale for simplicity, 
+and like Euterpea is designed to work well with MIDI.
 
@@
 Haskell MIDI libraries via Agda's Haskell FFI. 
-Details can be found in the github repository.
+Details can be found in the GitHub repository.
 
@@
 as the base cases of counterpoint (note that the final interval of
-the cadence is always \texttt{(p , per8)} and is thus not explicitly
+the cadence is always \texttt{(p\ ,\ per8)} and hence is not explicitly
 specified).
@@
 \noindent Observe that \texttt{motionOk} is an implicit argument of 
-the \texttt{\_::\_} constructor. 
+the \texttt{\_::\_} constructor, here denoted by the curly braces. 
 The argument can be resolved automatically by the type checker,
@@
 particularly interested in expressing the work done in Haskell
-on functional harmony in the library, to see just how much
+on functional harmony in the library, to see how much
 dependent types can simplify the representation.
@@
 
-For counterpoint specifically, we plan to represent higher species 
+For counterpoint specifically, we intend to represent higher species 
 counterpoint by extending the rules, 
@@
  for many helpful suggestions that improved our Agda code.
- We also thank the anonymous reviewers for their thoughtful feedback
+ We also thank the anonymous reviewers for their thoughtful feedback,
  which improved our presentation.

### 2020-02-02 doc/icfp20/counterpoint.tex
start counterpoint section

\section{Counterpoint}
\label{sec:cp}

Counterpoint is a technique for combining multiple lines of melodies.
Composing counterpoint is like arranging a song for a choir:
we start with a \emph{cantus firmus}, which serves as a base melody,
and compose counterpoint lines above or below the cantus firmus.
When doing this, we must make sure that the counterpoint lines sound
harmonically pleasing when played with the cantus firmus,
and that the individual melodic lines are distinguishable to the listener.

In this section, we show a dependently typed formalization of
two-voice counterpoint, more specifically,
those variants called the first and second species.
Here we restrict ourselves to \emph{upper} counterpoint
(i.e., composing a counterpoint line above the cantus firmus);
lower counterpoint can be formalized in the exactly same way.

\subsection{First Species Counterpoint}
\label{sec:cp:fs}

In the first species counterpoint, we start with a cantus firmus
consisting only of whole notes,
and set one note against each note in the cantus firmus.
To write good counterpoint, we must follow five rules explained below:

\paragraph{Beginning}
The beginning of the music must be a perfect consonance.
In the case of upper counterpoint, there are three ways to start the music:
using a unison, a 5th, or an 8th.

\paragraph{Intervals}
All intervals appearing in the main body must be consonant.
That is, the body should consist only of 3rds, 5ths, 6ths, and 8ths.

\paragraph{Motion}
For every pair of two adjacent intervals, the cantus firmus and counterpoint
lines must not move in the same direction if the second interval is a perfect one.
This is because such motion would draw too much attention
to the resulting interval.

\paragraph{Unisons}
Unisons should not occur in the main body of the music.
The reason is that they make it harder to distinguish between
the two melodic lines.

\paragraph{Ending}
The ending of the music must be a \emph{cadence}, a progression
of two intervals that induces a sense of resolution.
In our formalization, we fix the last interval to be an 8th,
hence the preceding interval should be either a major 6th or a minor 10th.

### 2020-02-05 doc/icfp20/counterpoint.tex
tweaks

@@
 Counterpoint is a technique for combining multiple lines of melodies.
-Composing counterpoint is like arranging a song for a choir:
-we start with a \emph{cantus firmus}, which serves as a base melody,
+Composing counterpoint is like arranging a song for choir:
+we start with a \emph{cantus firmus}, which serves as the base melody,
 and compose counterpoint lines above or below the cantus firmus.
@@
 
-In this section, we show a dependently typed formalization of
-two-voice counterpoint, more specifically,
-those variants called the first and second species.
-Here we restrict ourselves to \emph{upper} counterpoint
-(i.e., composing a counterpoint line above the cantus firmus);
-lower counterpoint can be formalized in the exactly same way.
+In this section, we show an implementation of species counterpoint, 
+based on the formulation given by Fux~\cite{fux-cp}.
+The idea is to represent rules for composing ``good'' counterpoint
+using dependent types.
+Here we describe two variants of species counterpoint;
+other variants can be formalized in a similar way.
 

### 2020-02-06 doc/icfp20/counterpoint.tex
working on text for first species

@@
 
-In this section, we show an implementation of species counterpoint, 
-based on the formulation given by Fux~\cite{fux-cp}.
+In this section, we present an implementation of species counterpoint,
+based on the formulation given by Fux~\citep{fux-cp}.
 The idea is to represent rules for composing ``good'' counterpoint
 using dependent types.
-Here we describe two variants of species counterpoint;
-other variants can be formalized in a similar way.
+[One more sentence here.]
 
@@
 
-In the first species counterpoint, we start with a cantus firmus
-consisting only of whole notes,
-and set one note against each note in the cantus firmus.
-To write good counterpoint, we must follow five rules explained below:
+The simplest variant of counterpoint is called the first species
+counterpoint.
+In first species, we set one note against each note in the cantus
+firmus, as in the following example:
+
+[Frog]
+
+In our implementation, we represent each bar as a pitch-interval
+pair, and the whole music as a list of such pairs.
+
+\begin{alltt}
+PitchInterval : Set
+PitchInterval = Pitch \(\times\) Interval
+\end{alltt}
+
+First species counterpoint comes with five mandatory rules, 
+constraining the beginning, the main body, and the ending of
+the music.
+Below, we describe these rules one by one, showing the
+corresponding Agda code.
 
 \paragraph{Beginning}
-The beginning of the music must be a perfect consonance.
-In the case of upper counterpoint, there are three ways to start the music:
-using a unison, a 5th, or an 8th.
+The beginning of the music must be a perfect consonance,
+that is, it must be a unison, a 5th, or an 8th.
+In our implementation, this is checked by the predicate
+\texttt{checkBeginning}, which is defined as follows:
+
+\begin{alltt}
+data BeginningError : Set where
+  not158   : PitchInterval \(\rightarrow\) BeginningError
+  tooShort : BeginningError
+
+beginningCheck : PitchInterval \(\rightarrow\) Maybe BeginningError
+beginningCheck pi@(_ , i) =
+  if ((i == per1) \(\vee\) (i == per5) \(\vee\) (i == per8))
+  then nothing
+  else just (not158 pi)
+
+checkBeginning : List PitchInterval \(\rightarrow\) Maybe BeginningError
+checkBeginning []       = just tooShort
+checkBeginning (p :: ps) = beginningCheck p  
+\end{alltt}
+
+\noindent The datatype \texttt{BeginningError} is a kind of
+``type error'' designed specifically for counterpoint.
+The first constructor \texttt{not158} tells us that the music
+begins with a wrong interval, whereas the second constructor
+\texttt{tooShort} says that the whole music is empty.
+With this datatype, we can obtain an informative error message
+when we accidentally break the rule at the beginning of the music.
+Thus, what we are doing here can be seen as writing a
+customized typechecker.
 
 \paragraph{Intervals}
-All intervals appearing in the main body must be consonant.
-That is, the body should consist only of 3rds, 5ths, 6ths, and 8ths.
+In the main body of the music, all intervals must be consonant.
+That is, we should only use 3rds, 5ths, 6ths, and 8ths.
+This rule can be encoded as the \texttt{checkIntervals} predicate,
+which again uses a new datatype \texttt{IntervalError} representing
+the presence of a dissonant interval:
+
+\begin{alltt}
+data IntervalError : Set where
+  dissonant : Interval \(\rightarrow\) IntervalError
+
+intervalCheck : Interval \(\rightarrow\) Maybe IntervalError
+intervalCheck i = if isConsonant i then nothing else just (dissonant i)
+
+checkIntervals : List PitchInterval \(\rightarrow\) List IntervalError
+checkIntervals = mapMaybe (intervalCheck \(\circ\) proj\(\sb{\mathtt{2}}\))
+\end{alltt}
 
 \paragraph{Motion}
-For every pair of two adjacent intervals, the cantus firmus and counterpoint
-lines must not move in the same direction if the second interval is a perfect one.
-This is because such motion would draw too much attention
-to the resulting interval.
+For every pair of two adjacent intervals, the cantus firmus and
+counterpoint lines must not move in the same direction if the second
+interval is a perfect one (i.e., unison, 5th, or 8th).
+This is because such motion would draw too much attention to the
+resulting interval.
 
@@
 The ending of the music must be a \emph{cadence}, a progression
-of two intervals that induces a sense of resolution.
-In our formalization, we fix the last interval to be an 8th,
-hence the preceding interval should be either a major 6th or a minor 10th.
+of two intervals that gives rise to a sense of resolution.

### 2020-02-07 doc/icfp20/counterpoint.tex
fix checkUnison; more on counterpoint

@@
 based on the formulation given by Fux~\citep{fux-cp}.
-The idea is to represent rules for composing ``good'' counterpoint
-using dependent types.
-[One more sentence here.]
+The idea is to represent ``good'' counterpoint  as a dependent record,
+whose fields are the proofs that the counterpoint follows certain rules.
+For space reason, we only describe two variants of species counterpoint;
+other variants can be formalized in an analogous way.
 
@@
 
-The simplest variant of counterpoint is called the first species
-counterpoint.
-In first species, we set one note against each note in the cantus
-firmus, as in the following example:
+First species counterpoint is the simplest variant of counterpoint.
+In first species, we set one note against each note in the cantus firmus,
+which starts with a tonic and consists only of whole notes.
+Here is an example of first species counterpoint, where the lower line
+stands for the cantus firmus and the upper line represents the
+counterpoint\footnote{The cantus firmus is part of a German song
+  called ``Froschgesang'' (Frog's song).}:
 
-[Frog]
+[Frog score]
 
-In our implementation, we represent each bar as a pitch-interval
-pair, and the whole music as a list of such pairs.
+We represent such music as a vector of pitch-interval pairs \texttt{(p , i)},
+where \texttt{p} is a note in the cantus firmus, and \texttt{i} is the
+distance between \texttt{p} and the corresponding counterpoint note.
 
@@
 PitchInterval = Pitch \(\times\) Interval
+
+frog-cp1 : Vector PitchInterval 7
+frog-cp1 = (c 5 , per8) :: (d 5 , maj6) :: (e 5 , min6) :: (f 5 , maj3) :: (e 5 , min3) :: (d 5 , maj6) :: (c 5 , per8) :: []
 \end{alltt}
 
-First species counterpoint comes with five mandatory rules, 
-constraining the beginning, the main body, and the ending of
-the music.
-Below, we describe these rules one by one, showing the
-corresponding Agda code.
+Note that the vector representation does not carry any correctness
+guarantee of the music;
+it requires extrinsic proofs stating that the music follows all the
+rules for being good counterpoint.
+Below, we discuss five such rules, showing what role they play
+and how they can be represented using dependent types.
 
@@
 \paragraph{Intervals}
-In the main body of the music, all intervals must be consonant.
+In the main body of the music, we must ensure that vertical intervals
+are harmonically pleasing.
+all intervals must be consonant.
 That is, we should only use 3rds, 5ths, 6ths, and 8ths.
@@
 \paragraph{Motion}
+
+\begin{alltt}
+data Motion : Set where
+contrary : Motion
+parallel : Motion
+similar  : Motion
+oblique  : Motion
+\end{alltt}
+
 For every pair of two adjacent intervals, the cantus firmus and
@@
 
+\begin{alltt}
+data MotionError : Set where
+  parallel : PitchInterval \(\rightarrow\) PitchInterval \(\rightarrow\) MotionError
+  similar  : PitchInterval \(\rightarrow\) PitchInterval \(\rightarrow\) MotionError
+
+motionCheck : PitchInterval \(\rightarrow\) PitchInterval \(\rightarrow\) Maybe MotionError
+motionCheck i1 i2 with motion i1 i2 | isPerfect (proj\(\sb{\mathtt{2}}\) i2)
+motionCheck i1 i2 | contrary | \_     = nothing
+motionCheck i1 i2 | oblique  | \_     = nothing
+motionCheck i1 i2 | parallel | false = nothing
+motionCheck i1 i2 | parallel | true  = just (parallel i1 i2)
+motionCheck i1 i2 | similar  | false = nothing
+motionCheck i1 i2 | similar  | true  = just (similar i1 i2)
+
+checkMotion : List PitchInterval → List MotionError
+checkMotion = mapMaybe (uncurry motionCheck) ∘ pairs  
+\end{alltt}
+
 \paragraph{Unisons}
@@
 
+\begin{alltt}
+data UnisonError : Set where
+  unison : Pitch \(\rightarrow\) UnisonError
+
+unisonCheck : PitchInterval \(\rightarrow\) Maybe UnisonError
+unisonCheck (p , i) = if (i == per1) then just (unison p) else nothing  
+\end{alltt}
+
 \paragraph{Ending}

### 2020-02-08 doc/icfp20/counterpoint.tex
renaming & text

@@
 The idea is to represent ``good'' counterpoint  as a dependent record,
-whose fields are the proofs that the counterpoint follows certain rules.
+whose fields encode the musical content as well as proofs that the
+counterpoint follows certain rules.
 For space reason, we only describe two variants of species counterpoint;
@@
 
-We represent such music as a vector of pitch-interval pairs \texttt{(p , i)},
-where \texttt{p} is a note in the cantus firmus, and \texttt{i} is the
-distance between \texttt{p} and the corresponding counterpoint note.
+We represent such music in the following way: we keep the first and
+last bars as separate elements, and encode the middle bars as a vector
+of pitch-interval pairs \texttt{(p , i)}, where \texttt{p} is a note in the
+cantus firmus, and \texttt{i} is the distance between \texttt{p} and the
+corresponding counterpoint note.
+For instance, the main body of the above example is represented as
+follows:
 
@@
 
-frog-cp1 : Vector PitchInterval 7
-frog-cp1 = (c 5 , per8) :: (d 5 , maj6) :: (e 5 , min6) :: (f 5 , maj3) :: (e 5 , min3) :: (d 5 , maj6) :: (c 5 , per8) :: []
+cfcp1 : List PitchInterval
+cfcp1 = (d 5 , maj6) :: (e 5 , min6) :: (f 5 , maj3) :: (e 5 , min3) :: (d 5 , maj6) :: []
 \end{alltt}
 
-Note that the vector representation does not carry any correctness
-guarantee of the music;
-it requires extrinsic proofs stating that the music follows all the
-rules for being good counterpoint.
-Below, we discuss five such rules, showing what role they play
-and how they can be represented using dependent types.
+The reason we separate different parts of the music is that they are
+subject to different rules.
+In the reminder of this section, we elaborate these rules one by one.
 
@@
   not158   : PitchInterval \(\rightarrow\) BeginningError
-  tooShort : BeginningError
 
-beginningCheck : PitchInterval \(\rightarrow\) Maybe BeginningError
-beginningCheck pi@(_ , i) =
+checkBeginning : PitchInterval \(\rightarrow\) Maybe BeginningError
+checkBeginning pi@(_ , i) =
   if ((i == per1) \(\vee\) (i == per5) \(\vee\) (i == per8))
@@
   else just (not158 pi)
-
-checkBeginning : List PitchInterval \(\rightarrow\) Maybe BeginningError
-checkBeginning []       = just tooShort
-checkBeginning (p :: ps) = beginningCheck p  
 \end{alltt}
@@
 unisonCheck (p , i) = if (i == per1) then just (unison p) else nothing
+
+checkUnison : List PitchInterval \(\rightarrow\) List UnisonError
+checkUnison = mapMaybe unisonCheck
 \end{alltt}

### 2020-02-08 doc/icfp20/counterpoint.tex
integrate unisonCheck into intervalCheck

@@
 which starts with a tonic and consists only of whole notes.
-Here is an example of first species counterpoint, where the lower line
-stands for the cantus firmus and the upper line represents the
-counterpoint\footnote{The cantus firmus is part of a German song
-  called ``Froschgesang'' (Frog's song).}:
+Figure~\ref{fig:frog} is an example of first species counterpoint,
+where the lower line stands for the cantus firmus and the upper line
+represents the counterpoint\footnote{The cantus firmus is part of a
+  German song called ``Froschgesang'' (Frog's song).}:
 
-[Frog score]
-
-We represent such music in the following way: we keep the first and
-last bars as separate elements, and encode the middle bars as a vector
-of pitch-interval pairs \texttt{(p , i)}, where \texttt{p} is a note in the
-cantus firmus, and \texttt{i} is the distance between \texttt{p} and the
-corresponding counterpoint note.
+Among different options of how to encode such music, we adopt
+the following strategy.
+First, we represent each bar as a pitch-interval pair \texttt{(p ,  i)},
+where \texttt{p} is a note in the cantus firmus, and \texttt{i} is the
+distance between \texttt{p} and the corresponding counterpoint note.
+Second, we keep the first and last bars as separate elements, and
+represent the middle bars (often referred to as the \emph{main body})
+as a list of pitch-interval pairs.
 For instance, the main body of the above example is represented as
@@
 \paragraph{Beginning}
-The beginning of the music must be a perfect consonance,
-that is, it must be a unison, a 5th, or an 8th.
-In our implementation, this is checked by the predicate
-\texttt{checkBeginning}, which is defined as follows:
+The beginning of the music should express perfection.
+Therefore, the first interval must be a unison, a 5th, or an 8th.
+In our formalization, we implement this rule as a function
+\texttt{checkBeginning}, which reports an error \texttt{not158 i}
+when the first interval \texttt{i} is an invalid one.
 
@@
 
-\noindent The datatype \texttt{BeginningError} is a kind of
-``type error'' designed specifically for counterpoint.
-The first constructor \texttt{not158} tells us that the music
-begins with a wrong interval, whereas the second constructor
-\texttt{tooShort} says that the whole music is empty.
-With this datatype, we can obtain an informative error message
-when we accidentally break the rule at the beginning of the music.
-Thus, what we are doing here can be seen as writing a
-customized typechecker.
-
 \paragraph{Intervals}
-In the main body of the music, we must ensure that vertical intervals
-are harmonically pleasing.
-all intervals must be consonant.
-That is, we should only use 3rds, 5ths, 6ths, and 8ths.
-This rule can be encoded as the \texttt{checkIntervals} predicate,
-which again uses a new datatype \texttt{IntervalError} representing
-the presence of a dissonant interval:
+The main body of the music should maintain harmonical consonance
+and independence of melodic lines.
+This means, all constituent intervals must be chosen from the 3rd,
+5th, 6th, and 8th.
+We encode this rule as the \texttt{checkIntervals} function,
+which returns a list of errors correponding to the occurrences of
+dissonant intervals and unisons.
 
@@
   dissonant : Interval \(\rightarrow\) IntervalError
+  unison    : Pitch \(\rightarrow\) IntervalError
 
-intervalCheck : Interval \(\rightarrow\) Maybe IntervalError
-intervalCheck i = if isConsonant i then nothing else just (dissonant i)
+intervalCheck : PitchInterval \(\rightarrow\) Maybe IntervalError
+intervalCheck (p , i) with isConsonant i | isUnison i
+... | false | _    = just (dissonant i)
+... | _     | true = just (unison p)
+... | _     | _    = nothing
 
 checkIntervals : List PitchInterval \(\rightarrow\) List IntervalError
-checkIntervals = mapMaybe (intervalCheck \(\circ\) proj\(\sb{\mathtt{2}}\))
+checkIntervals = mapMaybe intervalCheck
 \end{alltt}
@@
 \paragraph{Motion}
-
-\begin{alltt}
-data Motion : Set where
-contrary : Motion
-parallel : Motion
-similar  : Motion
-oblique  : Motion
-\end{alltt}
-
-For every pair of two adjacent intervals, the cantus firmus and
-counterpoint lines must not move in the same direction if the second
-interval is a perfect one (i.e., unison, 5th, or 8th).
-This is because such motion would draw too much attention to the
-resulting interval.
+The independence of melodic lines is also affected by \emph{motion},
+i.e., the way one interval moves to another interval.
+As shown in Figure~\ref{fig:motion}, there are four kinds of motion,
+and among them, parallel and similar motion makes it harder to
+distinguish between two voices, especially when the second interval
+is a perfect one.
+To avoid this problematic motion, we define the \texttt{checkMotion}
+function, which lists all invalid motion approaching a perfect interval.
 
@@
 
-checkMotion : List PitchInterval → List MotionError
-checkMotion = mapMaybe (uncurry motionCheck) ∘ pairs  
-\end{alltt}
-
-\paragraph{Unisons}
-Unisons should not occur in the main body of the music.
-The reason is that they make it harder to distinguish between
-the two melodic lines.
-
-\begin{alltt}
-data UnisonError : Set where
-  unison : Pitch \(\rightarrow\) UnisonError
-
-unisonCheck : PitchInterval \(\rightarrow\) Maybe UnisonError
-unisonCheck (p , i) = if (i == per1) then just (unison p) else nothing
-
-checkUnison : List PitchInterval \(\rightarrow\) List UnisonError
-checkUnison = mapMaybe unisonCheck
+checkMotion : List PitchInterval \(\rightarrow\) List MotionError
+checkMotion = mapMaybe (uncurry motionCheck) \(\circ\) pairs  
 \end{alltt}
@@
 \paragraph{Ending}
-The ending of the music must be a \emph{cadence}, a progression
-of two intervals that gives rise to a sense of resolution.
+The ending of the music should express relaxation.
+This means, the last interval must be chosen from the unison and
+the 8th, which are both considered very stable.
+The last interval should also be approached by a \emph{cadence},
+a progression that gives rise to a sense of resolution.
+We encode these rules as the \texttt{checkEnding} function;
+note that it returns the \texttt{tooShort} error when the music does
+not have enough number of bars to form a valid ending.
+
+\begin{alltt}
+data EndingError : Set where
+  not18    : PitchInterval \(\rightarrow\) EndingError
+  not27    : PitchInterval \(\rightarrow\) EndingError
+  tooShort : List PitchInterval \(\rightarrow\) EndingError
+
+endingCheck : PitchInterval \(\rightarrow\) PitchInterval \(\rightarrow\) Maybe EndingError
+endingCheck pi1@(pitch p , i) (pitch q , interval 0)  = 
+  if ((p + 1 \(\equiv\sp{\mathtt{b}}\) q) \(\wedge\) (i == min3)) then nothing else just (not27 pi1)
+endingCheck pi1@(pitch p , i) (pitch q , interval 12) =
+  if ((q + 2 \(\equiv\sp{\mathtt{b}}\) p) \(\wedge\) (i == maj6) \(\vee\) (p + 1 \(\equiv\sp{\mathtt{b}}\) q) \(\wedge\) (i == min10))
+  then nothing
+  else just (not27 pi1)
+endingCheck pi1               pi2                     =
+  just (not18 pi2)
+
+checkEnding : List PitchInterval \(\rightarrow\) PitchInterval \(\rightarrow\) Maybe EndingError
+checkEnding []        \_ = just (tooShort [])
+checkEnding (p :: []) q = endingCheck p q
+checkEnding (p :: ps) q = checkEnding ps q
+\end{alltt}

### 2020-02-09 doc/icfp20/counterpoint.tex
explain FirstSpecies record type

@@
 which starts with a tonic and consists only of whole notes.
-Figure~\ref{fig:frog} is an example of first species counterpoint,
-where the lower line stands for the cantus firmus and the upper line
-represents the counterpoint\footnote{The cantus firmus is part of a
-  German song called ``Froschgesang'' (Frog's song).}:
+Figure~\ref{fig:frog} is an example of first species counterpoint.
+The lower line is the cantus firmus, which we take from a famous
+German song called ``Froschgesang'' (Frog's song).
+The upper line is the counterpoint composed by the second author.
 
-Among different options of how to encode such music, we adopt
-the following strategy.
-First, we represent each bar as a pitch-interval pair \texttt{(p ,  i)},
-where \texttt{p} is a note in the cantus firmus, and \texttt{i} is the
-distance between \texttt{p} and the corresponding counterpoint note.
-Second, we keep the first and last bars as separate elements, and
-represent the middle bars (often referred to as the \emph{main body})
-as a list of pitch-interval pairs.
-For instance, the main body of the above example is represented as
-follows:
+\subsubsection{Representing Music}
+
+In our implementation, we represent each bar as a pitch-interval
+pair \texttt{(p ,  i)}, where \texttt{p} is a note in the cantus firmus,
+and \texttt{i} is the distance between \texttt{p} and the corresponding
+counterpoint note.
+We then represent the sequence of bars as a list of pitch-interval pairs,
+but with one proviso: we separate the first and last bars from the
+middle bars.
+Therefore, the Frog's song in Figure~\ref{fig:frog} is represented as a
+compound of the following three elements:
 
@@
 
-cfcp1 : List PitchInterval
-cfcp1 = (d 5 , maj6) :: (e 5 , min6) :: (f 5 , maj3) :: (e 5 , min3) :: (d 5 , maj6) :: []
+first : PitchInterval
+first = (c 5 , per8)
+
+middle : List PitchInterval
+middle = (d 5 , maj6) :: (e 5 , min6) :: (f 5 , maj3) :: (e 5 , min3) :: (d 5 , maj6) :: []
+
+last : PitchInterval
+last = (c 5 , per8)
 \end{alltt}
 
-The reason we separate different parts of the music is that they are
-subject to different rules.
-In the reminder of this section, we elaborate these rules one by one.
+\subsubsection{Representing Rules}
+
+The reason why we separate different parts of music is that
+they are subject to different rules.
+Below, we elaborate these rules one by one.
 
 \paragraph{Beginning}
+
 The beginning of the music should express perfection.
@@
 \paragraph{Intervals}
-The main body of the music should maintain harmonical consonance
-and independence of melodic lines.
+
+The middle bars of the music should maintain harmonical
+consonance and independence of melodic lines.
 This means, all constituent intervals must be chosen from the 3rd,
@@
 \paragraph{Motion}
+
 The independence of melodic lines is also affected by \emph{motion},
@@
 \paragraph{Ending}
+
 The ending of the music should express relaxation.
@@
 \end{alltt}
+
+\subsubsection{Putting Things Together}
+
+Using the encoding of music and rules we have seen so far,
+we define \texttt{FirstSpecies}, a record type inhabited by correct
+counterpoint.
+The first three fields of this record type represent the first, middle,
+and last bars, respectively.
+The last four fields stand for the proofs that the music satisfies
+all the required properties, that is:
+
+\begin{itemize}
+  \item The beginning is a perfect consonance
+  \item The middle bars have no dissonant interval or unison
+  \item The middle bars have no perfect interval approached by
+    parallel or similar motion
+  \item The ending is a unison or an octave approached by a cadence
+\end{itemize}
+
+\begin{alltt}
+record FirstSpecies : Set where
+  constructor firstSpecies
+  field
+    firstBar    : PitchInterval
+    middleBars  : List PitchInterval
+    lastBar     : PitchInterval
+    beginningOk : checkBeginning firstBar \(\equiv\) nothing
+    intervalsOk : checkIntervals middleBars \(\equiv\) []
+    motionOk    : checkMotion middleBars \(\equiv\) []
+    endingOk    : checkEnding middleBars lastBar \(\equiv\) nothing
+\end{alltt}
+
+With this record type, we can show that the counterpoint composed
+for the Frog's song is correct, since the music inhabits the
+\texttt{FIrstSpecies} type.
+
+\begin{alltt}
+fs : FirstSpecies
+fs = firstSpecies first1 middle1 last1 refl refl refl refl
+\end{alltt}

### 2020-02-10 doc/icfp20/counterpoint.tex
add figures & start second species section

@@
+\newcommand{\FS}{
+  \begin{figure}
+    \includegraphics[width=10cm]{fs.png}
+    \caption{First Species Counterpoint}
+    \label{fig:fs}
+  \end{figure}
+}
+
+\newcommand{\SS}{
+  \begin{figure}
+    \includegraphics[width=10cm]{ss.png}
+    \caption{Second Species Counterpoint}
+    \label{fig:ss}
+  \end{figure}
+}
+
 \section{Counterpoint}
@@
 and compose counterpoint lines above or below the cantus firmus.
-When doing this, we must make sure that the counterpoint lines sound
-harmonically pleasing when played with the cantus firmus,
-and that the individual melodic lines are distinguishable to the listener.
+When doing this, we must make sure that the whole music sounds
+harmonically pleasing, and that the individual melodic lines are
+distinguishable to the listener.
 
@@
 
+\FS
+
 First species counterpoint is the simplest variant of counterpoint.
@@
 which starts with a tonic and consists only of whole notes.
-Figure~\ref{fig:frog} is an example of first species counterpoint.
+Figure~\ref{fig:fs} is an example of first species counterpoint.
 The lower line is the cantus firmus, which we take from a famous
@@
 counterpoint note.
-We then represent the sequence of bars as a list of pitch-interval pairs,
+We then represent a sequence of bars as a list of pitch-interval pairs,
 but with one proviso: we separate the first and last bars from the
 middle bars.
-Therefore, the Frog's song in Figure~\ref{fig:frog} is represented as a
+Therefore, the Frog's song in Figure~\ref{fig:fs} is represented as a
 compound of the following three elements:
@@
 
-The reason why we separate different parts of music is that
-they are subject to different rules.
-Below, we elaborate these rules one by one.
+The reason behind our three-part representation of music is that
+different parts are subject to different rules.
+Let us go through the
 
@@
 when the first interval \texttt{i} is an invalid one.
+Since Agda does not have exceptions, we wrap this error around
+the \texttt{just} constructor for option values.
 
@@
 
+\noindent Recall that the functions representing these rules return
+either an option value or a list of found errors.
+Therefore, the correctness proofs either take the form
+\texttt{func args $\equiv$ nothing}, or look like \texttt{func args $\equiv$ []}.
+
 \begin{alltt}
@@
 \end{alltt}
+
+\subsection{Second Species Counterpoint}
+\label{sec:cp:ss}
+
+\SS
+
+We next discuss a more complex variant of counterpoint, called the
+second species.
+In second species, we set \emph{two} half notes against every cantus
+firmus note.
+Figure~\ref{fig:ss} shows an example of such two-against-one
+counterpoint, again composed for the Frog's song.
+To represent second species counterpoint, we define a new type
+\texttt{PitchInterval2} as a triple of a pitch and two intervals:
+
+\begin{alltt}
+PitchInterval2 : Set
+PitchInterval2 = Pitch \(\times\) Interval \(\times\) Interval
+\end{alltt}
+
+The rhythmic structure of secound species counterpoint gives rise to
+the distinction between \emph{strong beats} (the first interval in a bar)
+and \emph{weak beats} (the second interval).
+Roughly speaking, strong beats in middle bars are constrained by the
+same rules as in first species.
+That is, they must all be consonant, non-unison intervals, and in the 
+case of perfect intervals, the motion from the preceding interval (on
+the weak beat) must be either contrary or oblique.
+Weak beats, on the other hand, ...

### 2020-02-11 doc/icfp20/counterpoint.tex
more on second species

@@
   \begin{figure}
-    \includegraphics[width=10cm]{fs.png}
+    \includegraphics[width=11cm]{fs.png}
     \caption{First Species Counterpoint}
@@
 
-\newcommand{\SS}{
+\renewcommand{\SS}{
   \begin{figure}
@@
 middle : List PitchInterval
-middle = (d 5 , maj6) :: (e 5 , min6) :: (f 5 , maj3) :: (e 5 , min3) :: (d 5 , maj6) :: []
+middle = (d 5 , maj6) :: (e 5 , min6) :: (f 5 , maj3) ::
+         (e 5 , min3) :: (d 5 , maj6) :: []
 
@@
 counterpoint, again composed for the Frog's song.
-To represent second species counterpoint, we define a new type
-\texttt{PitchInterval2} as a triple of a pitch and two intervals:
+As can be seen from this example, the first and last bars may have
+a different structure from middle bars.
+More specifically, it is encouraged to begin the counterpoint line with
+a half rest and end with a whole note.
+Therefore, in our implementation, we reuse the \texttt{PitchInterval} 
+type for the first and last bars, and define a new type
+\texttt{PitchInterval2} for middle bars:
 
@@
 PitchInterval2 = Pitch \(\times\) Interval \(\times\) Interval
+
+first2 : PitchInterval
+first2 = (c 5 , per5)
+
+middle2 : List PitchInterval2
+middle2 =
+  (d 5 , min3 , per5) :: (e 5 , min3 , min6) :: (f 5 , maj3 , aug4) ::
+  (e 5 , min6 , min3) :: (d 5 , min3 , maj6) :: []
+
+last2 : PitchInterval
+last2 = (c 5 , per8)
 \end{alltt}
@@
 the weak beat) must be either contrary or oblique.
-Weak beats, on the other hand, ...
+Weak beats, on the other hand, are constrained in a looser manner.
+First, they are allowed to be dissonant if they are created by a
+\emph{passing tone}, i.e., a note in the middle of two step-wise
+motions in the same direction (as in bars 4-5 of Figure~\ref{fig:ss}).
+Second, they are allowed to be a unison if they are left by step in the
+opposite direction from their approach (as in bars 5-6 of Figure~\ref{fig:ss}).
+
+With these rules in mind, let us look at the definition of
+\texttt{SecondSpecies}, a record type inhabited by correct second
+species counterpoint:
+
+\begin{alltt}
+record SecondSpecies : Set where
+  constructor secondSpecies
+  field
+    firstBar      : PitchInterval 
+    middleBars    : List PitchInterval2
+    lastBar       : PitchInterval 
+    beginningOk   : checkBeginning2 firstBar \(\equiv\) nothing
+    strongBeatsOk : checkStrongBeats middleBars \(\equiv\) []
+    weakBeatsOk   : checkWeakBeats middleBars (secondPitch lastBar) \(\equiv\) []
+    motionOk      : checkMotion2 (firstBar ::
+                                  (expandPitchIntervals2 middleBars) ++
+                                  (lastBar :: [])) \(\equiv\) []
+    endingOk      : checkEnding2 middleBars lastBar \(\equiv\) nothing
+\end{alltt}
+
+\noindent As in \texttt{FirstSpecies}, we have three fields holding
+the musical content, followed by five fields carrying the proofs of
+various properties.
+Among the latter group of fields, \texttt{beginningOK} is defined
+analogously to the corresponding field of \texttt{FirstSpecies};
+the only difference is that the interval must not be a unison,
+as it prevents the listener from recognizing the beginning of the
+counterpoint line.
+The next two fields, \texttt{strongBeatsOK} and \texttt{weakBeatsOK},
+guarantee the correctness of strong and weak beats, respectively.
+What follows is \texttt{motionOK}, which converts a list of
+\texttt{PitchInterval2} into a list of \texttt{PitchInterval}, and checks
+motion accross bars.
+Lastly, we have \texttt{endingOK}, which forces the last two intervals
+to have a cadence structure.
+
+Using \texttt{SecondSpecies}, we can show that our second species
+counterpoint for the Frog's song is correct.
+
+\begin{alltt}
+ss : SecondSpecies
+ss = SecondSpecies first2 middle2 last2 refl refl refl refl refl
+\end{alltt}

### 2020-02-13 doc/icfp20/music.tex
musical preliminaries

\section{Musical Preliminaries}
\label{sec:music}

\paragraph{Pitches}
In music, a \emph{pitch} is a value denoting how high or low a sound is.
Pitches can either be represented either as an absolute value or as a
relative value paired with the octave the sound belongs to.
In our implementation, the two representations of pitches are defined
as \texttt{Pitch} and \texttt{PitchOctave}, respectively, with the proof
that converting back and forth between the two is an identity.

\begin{alltt}
data Pitch : Set where
  pitch : \(\mathbb{N}\) \(\rightarrow\) Pitch

data Octave : Set where
  octave : \(\mathbb{N}\) \(\rightarrow\) Octave

PitchOctave : Set
PitchOctave = Fin 12 \(\times\) Octave
\end{alltt}

\paragraph{Duration}
\emph{Duration} denotes a certain unit of time during which a sound
or silence lasts.
The unit is unspecified at the composition time; it is instantiated to
a concrete value when the music is played at a specific tempo.

\begin{alltt}
data Duration : Set where
  duration : \(\mathbb{N}\) \(\rightarrow\) Duration
\end{alltt}

\paragraph{Note}
Combining pitches and duration gives us \emph{notes}.
In our implementation, we represent notes with sound as \texttt{tone}s
and those without as \texttt{rest}s.

\begin{alltt}
data Note : Set where
  tone : Duration \(\rightarrow\) Pitch \(\rightarrow\) Note
  rest : Duration         \(\rightarrow\) Note
\end{alltt}

\paragraph{Intervals}
\emph{Intervals} are yet another key concept in music.
An interval denotes the difference in pitch between two notes.
There are 13 kinds of interval within an octave, and these intervals
may be classified from several different perspectives: (i) major or minor;
(ii) consonant or dissonant; and (iii) perfect or imperfect.
For convenience, we let the names of intervals carry the information
about the first and third classifications.

\begin{alltt}
data Interval : Set where
interval : \(\mathbb{N}\) \(\rightarrow\) Interval

per1  = interval 0 -- consonant
min2  = interval 1
maj2  = interval 2
min3  = interval 3 -- consonant
maj3  = interval 4 -- consonant
per4  = interval 5 
aug4  = interval 6
per5  = interval 7 -- consonant
min6  = interval 8 -- consonant
maj6  = interval 9 -- consonant
min7  = interval 10
maj7  = interval 11
per8  = interval 12 -- consonant
\end{alltt}

### 2020-02-14 doc/icfp20/counterpoint.tex
tweaks

@@
 In first species, we set one note against each note in the cantus firmus,
-which starts with a tonic and consists only of whole notes.
-Figure~\ref{fig:fs} is an example of first species counterpoint.
-The lower line is the cantus firmus, which we take from a famous
+which is required to start with a tonic and consist only of whole notes.
+Figure~\ref{fig:fs} shows an example of first species counterpoint.
+The lower line is the cantus firmus, which we excerpt from a famous
 German song called ``Froschgesang'' (Frog's song).
@@
 The reason behind our three-part representation of music is that
-different parts are subject to different rules.
-Let us go through the
+different parts are subject to different rules, as we detail below.
 
@@
 The beginning of the music should express perfection.
-Therefore, the first interval must be a unison, a 5th, or an 8th.
-In our formalization, we implement this rule as a function
-\texttt{checkBeginning}, which reports an error \texttt{not158 i}
-when the first interval \texttt{i} is an invalid one.
-Since Agda does not have exceptions, we wrap this error around
-the \texttt{just} constructor for option values.
+As we saw in Section~\ref{sec:music}, there are three intervals that
+are classified as perfect: the unison, the 5th, and the 8th.
+The first interval of the music must then be one of these intervals.
+In our formalization, we implement this rule as the
+\texttt{checkBeginning} function, which reports an error
+\texttt{not158 i} when the first interval \texttt{i} is an invalid one.
+Since Agda does not have exceptions, we turn the error into an option
+value by wrapping it around the \texttt{just} constructor.
 
@@
 
-The middle bars of the music should maintain harmonical
-consonance and independence of melodic lines.
-This means, all constituent intervals must be chosen from the 3rd,
-5th, 6th, and 8th.
-We encode this rule as the \texttt{checkIntervals} function,
-which returns a list of errors correponding to the occurrences of
-dissonant intervals and unisons.
+The middle bars of the music should maintain harmonical consonance
+and independence of melodic lines.
+As we saw before, consonant intervals include the unison, 3rd, 5th,
+6th, and 8th, and among these, the unison is clearly an obstacle to
+distinguish between the two lines of music.
+Therefore, the middle bars must consist of the latter four intervals.
+We encode this rule as the \texttt{checkIntervals} function, which
+returns a list of errors correponding to the occurrences of dissonant
+intervals and unisons.
 
@@
 checkMotion = mapMaybe (uncurry motionCheck) \(\circ\) pairs
+-- where pairs returns a list of all adjacent pairs in a given list
 \end{alltt}
@@
 The ending of the music should express relaxation.
-This means, the last interval must be chosen from the unison and
-the 8th, which are both considered very stable.
+Among different intervals, the unison and the 8th are considered
+most stable, hence the last interval of the music must be either of
+these intervals.
 The last interval should also be approached by a \emph{cadence},
 a progression that gives rise to a sense of resolution.
-We encode these rules as the \texttt{checkEnding} function;
-note that it returns the \texttt{tooShort} error when the music does
+We encode these rules as the \texttt{checkEnding} function.
+Note that it returns the \texttt{tooShort} error when the music does
 not have enough number of bars to form a valid ending.
@@
 
-Using the encoding of music and rules we have seen so far,
-we define \texttt{FirstSpecies}, a record type inhabited by correct
-counterpoint.
+Using the encoding of music and rules we have seen so far, we define
+\texttt{FirstSpecies}, a record type inhabited by correct counterpoint.
 The first three fields of this record type represent the first, middle,
 and last bars, respectively.
-The last four fields stand for the proofs that the music satisfies
-all the required properties, that is:
+The last four fields stand for the proofs that the music satisfies all
+the required properties, that is:
 
@@
 Therefore, the correctness proofs either take the form
-\texttt{func args $\equiv$ nothing}, or look like \texttt{func args $\equiv$ []}.
+\texttt{func args $\equiv$ nothing}, or look like
+\texttt{func args $\equiv$ []}.
 
@@
 the weak beat) must be either contrary or oblique.
+
 Weak beats, on the other hand, are constrained in a looser manner.

### 2020-02-14 doc/icfp20/music.tex
tweaks

@@
 In music, a \emph{pitch} is a value denoting how high or low a sound is.
-Pitches can either be represented either as an absolute value or as a
-relative value paired with the octave the sound belongs to.
+Pitches can be represented either as an absolute value or as a relative
+value paired with the octave the sound belongs to.
 In our implementation, the two representations of pitches are defined
-as \texttt{Pitch} and \texttt{PitchOctave}, respectively, with the proof
-that converting back and forth between the two is an identity.
+as \texttt{Pitch} and \texttt{PitchOctave}, respectively, accompagnied
+by a proof that converting back and forth between them is an identity.
 
@@
 about the first and third classifications.
+As a convention, we refer to \texttt{min2} and \texttt{maj2}  uniformly
+as ``2nd'', and similarly for other intervals.
+We also call \texttt{per1} the \emph{unison} in the rest of this section.
 

### 2020-02-16 doc/icfp20/counterpoint.tex
figures
move figures to fig/

@@
   \begin{figure}
-    \includegraphics[width=11cm]{fs.png}
+    \includegraphics[width=6cm]{fig/fs.png}
     \caption{First Species Counterpoint}
@@
 
+\newcommand{\Motion}{
+  \begin{figure}
+    \includegraphics[width=6cm]{fig/motion.png}
+    \hspace{1cm}
+    \includegraphics[width=3cm]{fig/badmotion.png} \\
+    %\vspace{-1cm}
+    \begin{flushleft}
+    \begin{small}
+      \hspace{1.95cm} contrary \ \ \ \ parallel \ \ \ \ \ \ \ similar \ \ \ \ \ \ oblique
+      \hspace{2.2cm} 5th \hspace{9.5mm} 8th
+    \end{small}
+    \end{flushleft}
+    \caption{Left: Four Kinds of Motion, Right: Invalid Motion}
+    \label{fig:motion}
+  \end{figure}
+}
+
 \renewcommand{\SS}{
   \begin{figure}
-    \includegraphics[width=10cm]{ss.png}
+    \includegraphics[width=7cm]{fig/ss.png}
     \caption{Second Species Counterpoint}
@@
 
-\FS
-
 First species counterpoint is the simplest variant of counterpoint.
@@
 which is required to start with a tonic and consist only of whole notes.
-Figure~\ref{fig:fs} shows an example of first species counterpoint.
-The lower line is the cantus firmus, which we excerpt from a famous
-German song called ``Froschgesang'' (Frog's song).
+Figure~\ref{fig:fs} shows an example of first species counterpoint:
+
+\begin{figure}[h]
+  \includegraphics[width=6cm]{fig/fs.png}
+  \caption{First Species Counterpoint}
+  \label{fig:fs}
+\end{figure}
+
+\noindent The lower line is the cantus firmus, which we excerpt from
+a famous German song called ``Froschgesang'' (Frog's song).
 The upper line is the counterpoint composed by the second author.
@@
 i.e., the way one interval moves to another interval.
-As shown in Figure~\ref{fig:motion}, there are four kinds of motion,
-and among them, parallel and similar motion makes it harder to
-distinguish between two voices, especially when the second interval
-is a perfect one.
-To avoid this problematic motion, we define the \texttt{checkMotion}
+
+\begin{figure}
+    \includegraphics[width=6cm]{fig/motion.png}
+    \hspace{1cm}
+    \includegraphics[width=3cm]{fig/badmotion.png} \\
+    \begin{flushleft}
+    \begin{small}
+      \hspace{1.95cm} contrary \ \ \ \ parallel \ \ \ \ \ \ \ similar \ \ \ \ \ \ oblique
+      \hspace{2.2cm} 5th \hspace{9.5mm} 8th
+    \end{small}
+    \end{flushleft}
+    \caption{Four Kinds of Motion (left) and Bad Motion (right)}
+    \label{fig:motion}
+  \end{figure}
+
+As can be seen from the figure on the left, there are four kinds of
+motion: \emph{contrary} (two lines go in different directions),
+\emph{parallel} (two lines go in the same direction by the same
+distance), \emph{similar} (two lines go in the same direction),
+and \emph{oblique} (one line plays the same note).
+Among these variants, parallel and similar motion makes it harder
+to distinguish between two voices, especially when the second
+interval is a perfect one, as in the figure on the right.
+To avoid problematic motion, we define the \texttt{checkMotion}
 function, which lists all invalid motion approaching a perfect interval.
@@
 
-\SS
-
 We next discuss a more complex variant of counterpoint, called the
@@
 firmus note.
-Figure~\ref{fig:ss} shows an example of such two-against-one
-counterpoint, again composed for the Frog's song.
-As can be seen from this example, the first and last bars may have
+Figure~\ref{fig:ss} is an example of such two-against-one counterpoint,
+again composed for the Frog's song:
+
+\begin{figure}[h]
+  \includegraphics[width=7cm]{fig/ss.png}
+  \caption{Second Species Counterpoint}
+  \label{fig:ss}
+\end{figure}
+
+As we can see from this example, the first and last bars may have
 a different structure from middle bars.

### 2020-02-20 doc/icfp20/conclusion.tex
start introduction

\section{Conclusion}
\label{sec:conclusion}

Text here.

### 2020-02-20 doc/icfp20/harmony.tex
start introduction

\section{Harmony}
\label{sec:harmony}

Text here.

### 2020-02-20 doc/icfp20/intro.tex
start introduction

\section{Introduction}
\label{sec:intro}

Edgar Var\`{e}se describes music as ``organized sound'', and
throughtout history cultures have created and applied systems of rules
and guidelines to govern the music they create, most notably the
Common Practice Period of Western music spanning the 17th to early
20th centuries. These rule systems are seldom absolute, and indeed
deliberate breaking the rules is often part of the aesthetic, but they
roughly constrain the music they apply to and give it a common form
and sound.

Artists and theoreticians have attempted to capture and codify these
rule systems, always informally in natural language, and typically
accompanied by examples from the existing literature. The intent is
both to analysize existing music and then to use these principles to
guide the creation of new music, in other words for synthesis.

Starting in the 20th century computers have become ubiquitous in
music in every area including sound sythesis, composition and
production (\citep{roads-tutorial}).

[more here]

### 2020-02-21 doc/icfp20/abstract.tex
put abstract in a separate file; fix typo

Throughout history, music has been composed following general rules
and guidelines, expressed informally through natural language and
examples. The expressiveness of dependent type theory allows us to
capture these rules formally, and then use them to automate
analysis and synthesis of music.

In this experience report, we explore expressing a small subset of the
rules of the common practice period in Agda, a functional programming
language with full dependent types. We focus on the construction of
species counterpoint as well as four-part harmonization of melody. We
point out both the advantages of using dependent types to express
music theory and some of the challenges that remain to make
languages like Agda more practical as a tool for musical exploration.

### 2020-02-21 doc/icfp20/intro.tex
put abstract in a separate file; fix typo

@@
 20th centuries. These rule systems are seldom absolute, and indeed
-deliberate breaking the rules is often part of the aesthetic, but they
+deliberate breaking of the rules is often part of the aesthetic, but they
 roughly constrain the music they apply to and give it a common form
@@
 
-Starting in the 20th century computers have become ubiquitous in
-music in every area including sound sythesis, composition and
+Starting in the 20th century, computers have become ubiquitous in
+music in every area, including sound sythesis, composition and
 production (\citep{roads-tutorial}).

### 2020-02-21 doc/icfp20/counterpoint.tex
remove some code

@@
     \includegraphics[width=3cm]{fig/badmotion.png} \\
-    %\vspace{-1cm}
     \begin{flushleft}
@@
 which is required to start with a tonic and consist only of whole notes.
-Figure~\ref{fig:fs} shows an example of first species counterpoint:
-
-\begin{figure}[h]
-  \includegraphics[width=6cm]{fig/fs.png}
-  \caption{First Species Counterpoint}
-  \label{fig:fs}
-\end{figure}
-
-\noindent The lower line is the cantus firmus, which we excerpt from
-a famous German song called ``Froschgesang'' (Frog's song).
+Figure~\ref{fig:fs} shows an example of first species counterpoint.
+The lower line is the cantus firmus, which we excerpt from a famous
+German song called ``Froschgesang'' (Frog's song).
 The upper line is the counterpoint composed by the second author.
@@
 
-In our implementation, we represent each bar as a pitch-interval
-pair \texttt{(p ,  i)}, where \texttt{p} is a note in the cantus firmus,
-and \texttt{i} is the distance between \texttt{p} and the corresponding
+In our implementation, we represent each bar as a pitch-interval pair
+\texttt{(p ,  i)}, where \texttt{p} is the pitch of a cantus firmus note,
+and \texttt{i} is the interval between \texttt{p} and the corresponding
 counterpoint note.
@@
 
+\Motion
+
 The independence of melodic lines is also affected by \emph{motion},
 i.e., the way one interval moves to another interval.
-
-\begin{figure}
-    \includegraphics[width=6cm]{fig/motion.png}
-    \hspace{1cm}
-    \includegraphics[width=3cm]{fig/badmotion.png} \\
-    \begin{flushleft}
-    \begin{small}
-      \hspace{1.95cm} contrary \ \ \ \ parallel \ \ \ \ \ \ \ similar \ \ \ \ \ \ oblique
-      \hspace{2.2cm} 5th \hspace{9.5mm} 8th
-    \end{small}
-    \end{flushleft}
-    \caption{Four Kinds of Motion (left) and Bad Motion (right)}
-    \label{fig:motion}
-  \end{figure}
-
-As can be seen from the figure on the left, there are four kinds of
-motion: \emph{contrary} (two lines go in different directions),
+As can be seen from Figure~\ref{fig:motion} (left), there are four
+kinds of motion: \emph{contrary} (two lines go in different directions),
 \emph{parallel} (two lines go in the same direction by the same
@@
 to distinguish between two voices, especially when the second
-interval is a perfect one, as in the figure on the right.
+interval is a perfect one, as in Figure~\ref{fig:motion} (right).
 To avoid problematic motion, we define the \texttt{checkMotion}
@@
 
+\SS
+
 We next discuss a more complex variant of counterpoint, called the
@@
 Figure~\ref{fig:ss} is an example of such two-against-one counterpoint,
-again composed for the Frog's song:
-
-\begin{figure}[h]
-  \includegraphics[width=7cm]{fig/ss.png}
-  \caption{Second Species Counterpoint}
-  \label{fig:ss}
-\end{figure}
-
+again composed for the Frog's song.
 As we can see from this example, the first and last bars may have

### 2020-02-21 doc/icfp20/music.tex
figures
move figures to fig/
remove some code

@@
 In our implementation, the two representations of pitches are defined
-as \texttt{Pitch} and \texttt{PitchOctave}, respectively, accompagnied
-by a proof that converting back and forth between them is an identity.
+as the \texttt{Pitch} and \texttt{PitchOctave} datatypes, respectively,
+accompanied by a proof that converting back and forth between them
+is an identity.
 
-\begin{alltt}
-data Pitch : Set where
-  pitch : \(\mathbb{N}\) \(\rightarrow\) Pitch
+% \begin{alltt}
+% data Pitch : Set where
+%   pitch : \(\mathbb{N}\) \(\rightarrow\) Pitch
 
-data Octave : Set where
-  octave : \(\mathbb{N}\) \(\rightarrow\) Octave
+% data Octave : Set where
+%   octave : \(\mathbb{N}\) \(\rightarrow\) Octave
 
-PitchOctave : Set
-PitchOctave = Fin 12 \(\times\) Octave
-\end{alltt}
+% PitchOctave : Set
+% PitchOctave = Fin 12 \(\times\) Octave
+% \end{alltt}
 
@@
 
-\begin{alltt}
-data Duration : Set where
-  duration : \(\mathbb{N}\) \(\rightarrow\) Duration
-\end{alltt}
+% \begin{alltt}
+% data Duration : Set where
+%   duration : \(\mathbb{N}\) \(\rightarrow\) Duration
+% \end{alltt}
 
-\paragraph{Note}
+\paragraph{Notes}
 Combining pitches and duration gives us \emph{notes}.
-In our implementation, we represent notes with sound as \texttt{tone}s
-and those without as \texttt{rest}s.
+In our implementation, we represent notes as a datatype \texttt{Note}
+with two constructors: \texttt{tone} for notes with sound and
+\texttt{rest} for those without.
 
-\begin{alltt}
-data Note : Set where
-  tone : Duration \(\rightarrow\) Pitch \(\rightarrow\) Note
-  rest : Duration         \(\rightarrow\) Note
-\end{alltt}
+% \begin{alltt}
+% data Note : Set where
+%   tone : Duration \(\rightarrow\) Pitch \(\rightarrow\) Note
+%   rest : Duration         \(\rightarrow\) Note
+% \end{alltt}
 
 \paragraph{Intervals}
+
 \emph{Intervals} are yet another key concept in music.
@@
 (ii) consonant or dissonant; and (iii) perfect or imperfect.
-For convenience, we let the names of intervals carry the information
-about the first and third classifications.
 As a convention, we refer to \texttt{min2} and \texttt{maj2}  uniformly
@@
 
-\begin{alltt}
-data Interval : Set where
-interval : \(\mathbb{N}\) \(\rightarrow\) Interval
-
-per1  = interval 0 -- consonant
-min2  = interval 1
-maj2  = interval 2
-min3  = interval 3 -- consonant
-maj3  = interval 4 -- consonant
-per4  = interval 5 
-aug4  = interval 6
-per5  = interval 7 -- consonant
-min6  = interval 8 -- consonant
-maj6  = interval 9 -- consonant
-min7  = interval 10
-maj7  = interval 11
-per8  = interval 12 -- consonant
-\end{alltt}
+\begin{figure}[h]
+  \includegraphics[width=12cm]{fig/interval.png} \\
+  \begin{flushleft}
+    \begin{footnotesize}
+      \hspace{1.45cm} \texttt{per1 \hspace{0.5mm} min2  \hspace{2.5mm}
+        maj2 \hspace{1.2mm} min3 \hspace{0.5mm} maj3 \ per4
+        \hspace{1mm}aug4
+        \hspace{0.5mm}  per5\hspace{1.2mm}  min6\hspace{1.6mm} maj6
+        \hspace{1.6mm}min7\hspace{1.2mm} maj7\hspace{1.2mm} per8} \\
+      consonant? \hspace{1.5mm} yes \hspace{4mm} no \hspace{7mm}  no
+      \hspace{4.5mm} yes \hspace{3.8mm}  yes \hspace{3.5mm} no \hspace{3.5mm}
+      no \hspace{4mm} yes \hspace{4mm} yes \hspace{3.2mm} yes
+      \hspace{3.7mm} no \hspace{4mm} no \hspace{3.8mm} yes
+    \end{footnotesize}
+  \end{flushleft}
+\end{figure}

### 2020-02-22 doc/icfp20/music.tex
fix min6

@@
         \hspace{1.6mm}min7\hspace{1.2mm} maj7\hspace{1.2mm} per8} \\
-      consonant? \hspace{1.5mm} yes \hspace{4mm} no \hspace{7mm}  no
-      \hspace{4.5mm} yes \hspace{3.8mm}  yes \hspace{3.5mm} no \hspace{3.5mm}
-      no \hspace{4mm} yes \hspace{4mm} yes \hspace{3.2mm} yes
+      consonant? \hspace{1.5mm} yes \hspace{4.2mm} no \hspace{6.5mm}  no
+      \hspace{4.7mm} yes \hspace{3.8mm}  yes \hspace{3.7mm} no \hspace{3.5mm}
+      no \hspace{4.3mm} yes \hspace{3.8mm} yes \hspace{3.2mm} yes
       \hspace{3.7mm} no \hspace{4mm} no \hspace{3.8mm} yes

### 2020-02-22 doc/icfp20/music.tex
improve music section

@@
 \paragraph{Pitches}
-In music, a \emph{pitch} is a value denoting how high or low a sound is.
-Pitches can be represented either as an absolute value or as a relative
-value paired with the octave the sound belongs to.
-In our implementation, the two representations of pitches are defined
-as the \texttt{Pitch} and \texttt{PitchOctave} datatypes, respectively,
-accompanied by a proof that converting back and forth between them
-is an identity.
+A \emph{pitch} is a value denoting how high or low a sound is.
+In our implementation, pitches have type \texttt{Pitch} and take the
+form \texttt{name octave}, where \texttt{name} is a name ranging
+over \texttt{c}, \texttt{d}, \texttt{e}, etc. and \texttt{octave} is a
+natural number denoting which octave the pitch belongs to.
+For instance, the middle C is represented as \texttt{c 5}.
 
@@
 \paragraph{Duration}
-\emph{Duration} denotes a certain unit of time during which a sound
-or silence lasts.
-The unit is unspecified at the composition time; it is instantiated to
-a concrete value when the music is played at a specific tempo.
+\emph{Duration} denotes an unspecified unit of time during which a
+sound or silence lasts.
+In our implementation, duration (of type \texttt{Duration} is simply
+a natural number, denoting the relative length within a bar (such as
+whole and half).
+When the music is played at a specific tempo, the number is 
+multiplied by a velocity and turned into an absolute length.
 
@@
 
-\emph{Intervals} are yet another key concept in music.
-An interval denotes the difference in pitch between two notes.
+\emph{Intervals} are a concept denoting the difference in pitch
+between two notes.
 There are 13 kinds of interval within an octave, and these intervals
-may be classified from several different perspectives: (i) major or minor;
-(ii) consonant or dissonant; and (iii) perfect or imperfect.
-As a convention, we refer to \texttt{min2} and \texttt{maj2}  uniformly
-as ``2nd'', and similarly for other intervals.
-We also call \texttt{per1} the \emph{unison} in the rest of this section.
+can be classified from several different perspectives: (i) major or
+minor; (ii) consonant or dissonant; and (iii) perfect or imperfect.
+We define intervals as a datatype \texttt{Interval}, where each
+constructor represents one of the 13 intervals.
 

### 2020-02-22 doc/icfp20/counterpoint.tex
improve counterpoint section

@@
 As we saw in Section~\ref{sec:music}, there are three intervals that
-are classified as perfect: the unison, the 5th, and the 8th.
+are classified as perfect: the 1st (more commonly called the
+\emph{unison}), the 5th, and the 8th.
 The first interval of the music must then be one of these intervals.
@@
 and independence of melodic lines.
-As we saw before, consonant intervals include the unison, 3rd, 5th,
-6th, and 8th, and among these, the unison is clearly an obstacle to
-distinguish between the two lines of music.
+As we saw before, consonant intervals include the unison,
+3rd\footnote{We use ``3rd'' to mean both the major and minor variants
+of the 3rd interval, and similarly for the 6th.}, 5th, 6th, and 8th,
+and among these, the unison is clearly an obstacle to distinguishing
+between the two lines of music.
 Therefore, the middle bars must consist of the latter four intervals.
@@
 i.e., the way one interval moves to another interval.
-As can be seen from Figure~\ref{fig:motion} (left), there are four
+As can be seen from Figure~\ref{fig:motion} left, there are four
 kinds of motion: \emph{contrary} (two lines go in different directions),
@@
 and \emph{oblique} (one line plays the same note).
-Among these variants, parallel and similar motion makes it harder
-to distinguish between two voices, especially when the second
-interval is a perfect one, as in Figure~\ref{fig:motion} (right).
-To avoid problematic motion, we define the \texttt{checkMotion}
-function, which lists all invalid motion approaching a perfect interval.
+There is a consensus that approaching a perfect interval by parallel
+or similar motion (as in Figure~\ref{fig:motion} right) destroys the
+independence of lines.
+To rule out this, we define the \texttt{checkMotion} function, which
+lists all occurrences of such invalid motion.
 
@@
 
-With this record type, we can show that the counterpoint composed
-for the Frog's song is correct, since the music inhabits the
-\texttt{FIrstSpecies} type.
+With this record type, we can show that the counterpoint in
+Figure~\ref{fig:fs} is correct, since the music is an inhabitant of
+the \texttt{FIrstSpecies} type.
 
@@
 again composed for the Frog's song.
-As we can see from this example, the first and last bars may have
-a different structure from middle bars.
+
+\subsubsection{Representing Music}
+
+In second species, the first and last bars may have a different structure
+from middle bars.
 More specifically, it is encouraged to begin the counterpoint line with
@@
 
+\subsubsection{Representing Rules}
+
+In second species counterpoint, some parts of the music are applied
+the same rules as first species, while other parts are constrained by
+new rules.
+
+\paragraph{Beginning}
+
+The beginning of the music is \emph{not} allowed to be the unison,
+as it prevents the listener from recognizing the beginning of the
+counterpoint line.
+We restrict the first interval by defining a new error
+\texttt{BeginningError2} and a function \texttt{checkBeginning2}
+that works analogously to \texttt{checkBeginning} for first species.
+
+\paragraph{Strong Beats}
+
 The rhythmic structure of secound species counterpoint gives rise to
@@
 and \emph{weak beats} (the second interval).
-Roughly speaking, strong beats in middle bars are constrained by the
-same rules as in first species.
+Strong beats in middle bars are constrained by the same rules as in
+first species.
 That is, they must all be consonant, non-unison intervals, and in the 
@@
 the weak beat) must be either contrary or oblique.
+We encode these restrictions as the \texttt{checkStrongBeats}
+function, which, as the \texttt{checkIntervals} function for first
+species, returns a list of found errors.
 
-Weak beats, on the other hand, are constrained in a looser manner.
+\paragraph{Weak Beats}
+
+Compared to strong beasts, weak beats are constrained in a looser
+manner.
 First, they are allowed to be dissonant if they are created by a
@@
 motions in the same direction (as in bars 4-5 of Figure~\ref{fig:ss}).
-Second, they are allowed to be a unison if they are left by step in the
-opposite direction from their approach (as in bars 5-6 of Figure~\ref{fig:ss}).
+Second, they are allowed to be the unison if they are left by step in
+the opposite direction from their approach (as in bars 5-6 of
+Figure~\ref{fig:ss}).
+We implement these rules as the \texttt{checkWeakBeats} function,
+which checks the validity of every three successive beats in the
+middle bars.
 
-With these rules in mind, let us look at the definition of
-\texttt{SecondSpecies}, a record type inhabited by correct second
-species counterpoint:
+\paragraph{Motion}
+
+Parallel and similar motion towards a perfect interval is prohibited
+across bars.
+We enforce this rule using the \texttt{checkMotion2} function,
+which uses \texttt{expandPitchIntervals2} to convert a list of
+\texttt{PitchInterval2} into a list of \texttt{PitchInterval}, and
+checks every pair of weak beat and strong beat intervals.
+
+\subsubsection{Putting Things Together}
+
+Now we define \texttt{SecondSpecies}, a record type inhabited by
+correct second species counterpoint.
+As in \texttt{FirstSpecies}, we have three fields holding the musical
+content, followed by five fields carrying the proofs of various
+properties.
 
@@
 
-\noindent As in \texttt{FirstSpecies}, we have three fields holding
-the musical content, followed by five fields carrying the proofs of
-various properties.
-Among the latter group of fields, \texttt{beginningOK} is defined
-analogously to the corresponding field of \texttt{FirstSpecies};
-the only difference is that the interval must not be a unison,
-as it prevents the listener from recognizing the beginning of the
-counterpoint line.
-The next two fields, \texttt{strongBeatsOK} and \texttt{weakBeatsOK},
-guarantee the correctness of strong and weak beats, respectively.
-What follows is \texttt{motionOK}, which converts a list of
-\texttt{PitchInterval2} into a list of \texttt{PitchInterval}, and checks
-motion accross bars.
-Lastly, we have \texttt{endingOK}, which forces the last two intervals
-to have a cadence structure.
-
-Using \texttt{SecondSpecies}, we can show that our second species
-counterpoint for the Frog's song is correct.
+Using \texttt{SecondSpecies}, we can show that the second species
+counterpoint in Figure~\ref{fig:ss} is correct.
 

### 2020-02-23 doc/icfp20/music.tex
better text for music

@@
 \paragraph{Pitches}
-A \emph{pitch} is a value denoting how high or low a sound is.
-In our implementation, pitches have type \texttt{Pitch} and take the
-form \texttt{name octave}, where \texttt{name} is a name ranging
-over \texttt{c}, \texttt{d}, \texttt{e}, etc. and \texttt{octave} is a
-natural number denoting which octave the pitch belongs to.
+A \emph{pitch} tells us how high or low a sound is.
+This notion is represented as natural numbers (of type
+\texttt{Pitch}) in our implementation, with zero representing the
+lowest possible sound.
+Throughout the paper, we use the notation \texttt{name octave}
+for pitches, where \texttt{name} ranges over \texttt{c}, \texttt{d},
+\texttt{e}, etc., and \texttt{octave} denotes which octave the pitch
+belongs to.
 For instance, the middle C is represented as \texttt{c 5}.
@@
 \paragraph{Duration}
-\emph{Duration} denotes an unspecified unit of time during which a
-sound or silence lasts.
-In our implementation, duration (of type \texttt{Duration} is simply
-a natural number, denoting the relative length within a bar (such as
-whole and half).
-When the music is played at a specific tempo, the number is 
-multiplied by a velocity and turned into an absolute length.
+\emph{Duration} denotes an unspecified unit of time during which
+a sound or silence lasts.
+In our implementation, duration is again a natural number (of type
+\texttt{Duration}), with names such as \texttt{whole}, \texttt{half},
+and \texttt{quarter}.
+These values are turned into an absolute length when the music is
+played at a specific tempo.
+As an example, \texttt{whole} has value \texttt{16} and corresponds
+to 4 seconds if the tempo is 60 beats per minute.
 
@@
 Combining pitches and duration gives us \emph{notes}.
-In our implementation, we represent notes as a datatype \texttt{Note}
-with two constructors: \texttt{tone} for notes with sound and
-\texttt{rest} for those without.
+We construct notes (of type \texttt{Note}) with the \texttt{tone}
+and \texttt{rest} constructors.
+For example, \texttt{tone whole (c 5)} is a whole note whose
+sound is the middle C, and \texttt{rest half} is a half rest.
 
@@
 
-\emph{Intervals} are a concept denoting the difference in pitch
-between two notes.
-There are 13 kinds of interval within an octave, and these intervals
-can be classified from several different perspectives: (i) major or
-minor; (ii) consonant or dissonant; and (iii) perfect or imperfect.
+An \emph{interval} represents the difference in pitch between
+two notes.
+There are 13 kinds of interval within an octave, and these
+intervals can be classified from several different perspectives:
+(i) major or minor; (ii) consonant or dissonant; and (iii) perfect
+or imperfect.
 We define intervals as a datatype \texttt{Interval}, where each

### 2020-02-29 doc/icfp20/intro.tex

@@
 deliberate breaking of the rules is often part of the aesthetic, but they
-roughly constrain the music they apply to and give it a common form
-and sound.
+roughly constrain the music they apply to and
+give it a common form and sound.
 
 Artists and theoreticians have attempted to capture and codify these
-rule systems, always informally in natural language, and typically
+rule systems, informally in natural language, and typically
 accompanied by examples from the existing literature. The intent is
-both to analysize existing music and then to use these principles to
+both to analyze existing music and then to use these principles to
 guide the creation of new music, in other words for synthesis.
@@
 music in every area, including sound sythesis, composition and
-production (\citep{roads-tutorial}).
+production~\citep{roads-tutorial}. In terms of music theory, there has
+been a line of recent work on using functional programming for
+harmonic
+analysis~\citep{magalhaes-harmtrace,dehaas-harmtrace-a,dehaas-harmtrace-b},
+harmonization of a melody and generation of melodies based on a
+harmonization~\citep{koops-fharm,magalhaes-fcomp} and
+counterpoint~\citep{szamozvancev-welltyped}. There is also an
+established Haskell library Euterpea~\citep{hudak-haskell} for general music
+and sound exploration.
 
-[more here]
+To describe rules of basic harmonic structure,
+\citet{magalhaes-harmtrace} and its successors use dependent
+types. (TODO: Give an example.) However Haskell currently has limited
+support for dependent types, and requires many extentions and tricks
+such as the use of singleton types~\citep{eisenberg-singleton} to
+simulate a restricted form of dependent types. (TODO: fix wording)
+
+We would like to explore what can be done in the context of music by
+using a programming language that offers full dependent types. We use
+the language Agda~\citep{norell-phd} since it is fairly mature and
+aims to be a functional programming language as well as a proof
+assistant. It also features a Haskell FFI so we can take advantage of
+existing Haskell libraries, in partiular for MIDI.
+
+Full dependent types allow a fully expressive logic, and it is
+tempting to take a standard textbook on music theory such as
+\citet{piston-harmony} or \citet{aldwell2018harmony} and attempt to
+formalize it in type theory.
+
+-- people are starting to use computers for analysis and synthesis of rule systems
+(reference Haskell work). In particular FP, and some rudimentary dep types
+
+-- With full dep types, expressive logic, tempting to formalize all the rules
+of some theory book in dep types. Point out difficulties....
+
+-- Music Tools, compare to Euterpea + other Haskell work, mention FARM.
+
+-- In this paper, focus on counterpoint plus basic start to some harmony.
+
+Counterpoint section: contrast old and new methods
+
+Harmony section: Harmonize melody, use counterpoint for voice leading and to contrain harmony.
+Mention grid for harmonic analsis, makes use of vectors.
+
+conclusion: easy to express rules, break into parts, reuse in other contexts, extend.
+Downsides: hard to do some things (equality, sets, etc), Proof emphasis can get in the way, but
+may want to use in the future. Future work: equivalence, ornaments, ML.
+
+TODO: Fix bib style.

### 2020-02-29 doc/icfp20/intro.tex
wip

@@
 
-Full dependent types allow a fully expressive logic, and it is
+Full dependent types allow expression of full predicate logic, and it is
 tempting to take a standard textbook on music theory such as
 \citet{piston-harmony} or \citet{aldwell2018harmony} and attempt to
-formalize it in type theory.
+formalize it in type theory. However this is difficult since most
+``rules'' in music theory are not absolute, but rather suggestions of
+varying degrees (themselves vaguely specified) of importance. Since
+ultimately the resulting created music is more important than the
+theory, an alternative is to attempt to indirectly deduce the rules
+from music samples, using for example machine
+learning~\citep{huang-cp}. However.... TODO
+TODO: Perhaps create rules and weight them, then use ML.
 
--- people are starting to use computers for analysis and synthesis of rule systems
-(reference Haskell work). In particular FP, and some rudimentary dep types
+Although formalization of a large amount of music theory is a goal,
+for now we start with the modest task of expressing a small,
+relatively strict rule set known as species
+counterpoint~\cite{fux-cp}. TODO--more here
+We also present preliminary work on harmonizing a melody using
+a subset of rules based on \citet{piston-harmony}, contrasting with
+existing work by \citet{koops-fharm}. Notably, since counterpoint and
+harmony are not separate concepts but in fact deeply intertwined, we
+are able to reuse counterpoint rules to help develop natural-sounding
+harmonizations.
 
--- With full dep types, expressive logic, tempting to formalize all the rules
-of some theory book in dep types. Point out difficulties....
+We make use of and extend~\cite{MusicTools}, a library of small tools
+for analysis and synthesis of music written in Agda, with the goal
+of eventually being a dependently-typed alternative to Euterpea. 
+Previous work by ~\cite{cong-cp} using Music Tools expressed
+first-species counterpoint with the Agda type system, so that an error
+in counterpoint resulted in an Agda type error. Although this takes
+advantage of Agda's native type checking, it has several
+downsides. One is that the Agda errors may be difficult to interpret
+as the corresponding musical error. Another is TODO: flexibility and
+composability.
 
--- Music Tools, compare to Euterpea + other Haskell work, mention FARM.
+This is an experience report, and we highlight both the advantages and
+disadvantages of using Agda
 
--- In this paper, focus on counterpoint plus basic start to some harmony.
-
-Counterpoint section: contrast old and new methods
-
-Harmony section: Harmonize melody, use counterpoint for voice leading and to contrain harmony.
-Mention grid for harmonic analsis, makes use of vectors.
-
-conclusion: easy to express rules, break into parts, reuse in other contexts, extend.
-Downsides: hard to do some things (equality, sets, etc), Proof emphasis can get in the way, but
-may want to use in the future. Future work: equivalence, ornaments, ML.
-
-TODO: Fix bib style.
+TODO: check that bib style is allowed.

### 2020-02-29 doc/icfp20/intro.tex
add diagrams

@@
 Edgar Var\`{e}se describes music as ``organized sound'', and
-throughtout history cultures have created and applied systems of rules
+throughout history cultures have created and applied systems of rules
 and guidelines to govern the music they create, most notably the
@@
 
-We would like to explore what can be done in the context of music by
+In this paper we explore what can be done in the context of music by
 using a programming language that offers full dependent types. We use
@@
 
-Full dependent types allow expression of full predicate logic, and it is
+Full dependent types allow expression of predicate logic, and it is
 tempting to take a standard textbook on music theory such as
@@
 
-We make use of and extend~\cite{MusicTools}, a library of small tools
+We make use of and extend~\citet{MusicTools}, a library of small tools
 for analysis and synthesis of music written in Agda, with the goal
 of eventually being a dependently-typed alternative to Euterpea. 
-Previous work by ~\cite{cong-cp} using Music Tools expressed
+Previous work by ~\citet{cong-cp} using Music Tools expressed
 first-species counterpoint with the Agda type system, so that an error
@@
 This is an experience report, and we highlight both the advantages and
-disadvantages of using Agda
+disadvantages of using Agda. TODO: MORE
+
+TODO: Talk about conclusion.
 

### 2020-02-29 doc/icfp20/counterpoint.tex
wip
better figure for motion
separate figures of 4 motions and 2 bad patterns
tweak counterpoint section

@@
   \begin{figure}
-    \includegraphics[width=6cm]{fig/motion.png}
-    \hspace{1cm}
-    \includegraphics[width=3cm]{fig/badmotion.png} \\
+    \includegraphics[width=7cm]{fig/motion1.png}
+    \hspace{8mm}
+    \includegraphics[width=3.35cm]{fig/motion2.png}
     \begin{flushleft}
     \begin{small}
-      \hspace{1.95cm} contrary \ \ \ \ parallel \ \ \ \ \ \ \ similar \ \ \ \ \ \ oblique
-      \hspace{2.2cm} 5th \hspace{9.5mm} 8th
+      \hspace{1.55cm} contrary \ \ \ \ \ \ \ \ parallel \ \ \ \ \ \ \
+      \ \ \,similar \ \ \ \ \ \ \ \ \ oblique
+      \hspace{1.3cm} paralell-5th \ \ \ similar-8th
     \end{small}
     \end{flushleft}
-    \caption{Left: Four Kinds of Motion, Right: Invalid Motion}
+    \caption{Four Kinds of Motion (left) and Invalid Patterns (right)}
     \label{fig:motion}
@@
 
+TODO: contrast old and new methods
+
 Counterpoint is a technique for combining multiple lines of melodies.
@@
 we start with a \emph{cantus firmus}, which serves as the base melody,
-and compose counterpoint lines above or below the cantus firmus.
+and compose a counterpoint line above or below the cantus firmus.
 When doing this, we must make sure that the whole music sounds
@@
 In this section, we present an implementation of species counterpoint,
-based on the formulation given by Fux~\citep{fux-cp}.
+based on the formulation given by \citet{fux-cp}.
 The idea is to represent ``good'' counterpoint  as a dependent record,
@@
 In first species, we set one note against each note in the cantus firmus,
-which is required to start with a tonic and consist only of whole notes.
+which is required to start with a tonic (the first note of a scale;
+e.g., C in the case of C major) and consist only of whole notes.
 Figure~\ref{fig:fs} shows an example of first species counterpoint.
-The lower line is the cantus firmus, which we excerpt from a famous
-German song called ``Froschgesang'' (Frog's song).
+The lower line is the cantus firmus, which we excerpt from a German
+song called \emph{Froschgesang} (Frog's song).
 The upper line is the counterpoint composed by the second author.
@@
 As we saw in Section~\ref{sec:music}, there are three intervals that
-are classified as perfect: the 1st (more commonly called the
-\emph{unison}), the 5th, and the 8th.
+are classified as perfect: the 1st (usually called the \emph{unison}),
+t5th, and 8th.
 The first interval of the music must then be one of these intervals.
@@
 i.e., the way one interval moves to another interval.
-As can be seen from Figure~\ref{fig:motion} left, there are four
-kinds of motion: \emph{contrary} (two lines go in different directions),
+As Figure~\ref{fig:motion} left shows, there are four kinds of motion:
+\emph{contrary} (two lines go in different directions),
 \emph{parallel} (two lines go in the same direction by the same
@@
 and \emph{oblique} (one line plays the same note).
-There is a consensus that approaching a perfect interval by parallel
-or similar motion (as in Figure~\ref{fig:motion} right) destroys the
-independence of lines.
-To rule out this, we define the \texttt{checkMotion} function, which
-lists all occurrences of such invalid motion.
+It is known that approaching a perfect interval by parallel or similar
+motion (as in Figure~\ref{fig:motion} right) destroys the independence
+of lines.
+To rule out such motion, we define the \texttt{checkMotion}
+function, which checks all pairs of two adjacent intervals and
+returns a list of erroneous patterns.
 
@@
 checkMotion = mapMaybe (uncurry motionCheck) \(\circ\) pairs
--- where pairs returns a list of all adjacent pairs in a given list
+-- where pairs l returns a list of pairs consisting of two adjacent elements of l
 \end{alltt}
@@
 The ending of the music should express relaxation.
-Among different intervals, the unison and the 8th are considered
+Among different intervals, the unison and 8th are considered
 most stable, hence the last interval of the music must be either of
@@
 a progression that gives rise to a sense of resolution.
-We encode these rules as the \texttt{checkEnding} function.
-Note that it returns the \texttt{tooShort} error when the music does
-not have enough number of bars to form a valid ending.
+This in turn suggests that a valid ending requires the middle bars
+to be non-empty.
+We encode these rules as the \texttt{checkEnding} function, which,
+upon finding an invalid ending, tells us which rules are not satisfied.
 

### 2020-03-01 doc/icfp20/counterpoint.tex
more tweaks

@@
 firmus note.
-Figure~\ref{fig:ss} is an example of such two-against-one counterpoint,
+This gives rise to the distinction between \emph{strong} beats (the
+first interval in a bar) and \emph{weak beats} (the second interval).
+Figure~\ref{fig:ss} is an example of two-against-one counterpoint,
 again composed for the Frog's song.
@@
 type for the first and last bars, and define a new type
-\texttt{PitchInterval2} for middle bars:
+\texttt{PitchInterval2} for middle bars.
 
@@
 
-In second species counterpoint, some parts of the music are applied
-the same rules as first species, while other parts are constrained by
-new rules.
+The rules for second species counterpoint can be obtained by
+tweaking those for first species and adding a few new ones.
+Here we go through the rules without showing the corresponding
+Agda functions, as they are largely similar to what we have seen
+before.
 
@@
 counterpoint line.
-We restrict the first interval by defining a new error
-\texttt{BeginningError2} and a function \texttt{checkBeginning2}
-that works analogously to \texttt{checkBeginning} for first species.
+% We restrict the first interval by defining a new error
+% \texttt{BeginningError2} and a function \texttt{checkBeginning2}
+% that works analogously to \texttt{checkBeginning} for first species.
 
@@
 
-The rhythmic structure of secound species counterpoint gives rise to
-the distinction between \emph{strong beats} (the first interval in a bar)
-and \emph{weak beats} (the second interval).
 Strong beats in middle bars are constrained by the same rules as in
-first species.
-That is, they must all be consonant, non-unison intervals, and in the 
-case of perfect intervals, the motion from the preceding interval (on
-the weak beat) must be either contrary or oblique.
-We encode these restrictions as the \texttt{checkStrongBeats}
-function, which, as the \texttt{checkIntervals} function for first
-species, returns a list of found errors.
+first species: they must all be consonant, non-unison intervals.
+% We encode this restriction as the \texttt{checkStrongBeats}
+% function, which, as the \texttt{checkIntervals} function for first
+% species, returns a list of found errors.
 
@@
 
-Compared to strong beasts, weak beats are constrained in a looser
-manner.
+Weak beats are constrained in a looser manner.
 First, they are allowed to be dissonant if they are created by a
@@
 Figure~\ref{fig:ss}).
-We implement these rules as the \texttt{checkWeakBeats} function,
-which checks the validity of every three successive beats in the
-middle bars.
+% We implement these rules as the \texttt{checkWeakBeats} function,
+% which checks the validity of every three successive beats in the
+% middle bars.
 
@@
 across bars.
-We enforce this rule using the \texttt{checkMotion2} function,
-which uses \texttt{expandPitchIntervals2} to convert a list of
-\texttt{PitchInterval2} into a list of \texttt{PitchInterval}, and
-checks every pair of weak beat and strong beat intervals.
+% We enforce this rule using the \texttt{checkMotion2} function,
+% which uses \texttt{expandPitchIntervals2} to convert a list of
+% \texttt{PitchInterval2} into a list of \texttt{PitchInterval}, and
+% checks every pair of weak beat and strong beat intervals.
 
@@
 As in \texttt{FirstSpecies}, we have three fields holding the musical
-content, followed by five fields carrying the proofs of various
-properties.
+content, followed by five fields carrying the proofs of the properties
+disucssed above.
 

### 2020-03-01 doc/icfp20/counterpoint.tex
comparison with farm work

@@
     intervalsOk : checkIntervals middleBars \(\equiv\) []
-    motionOk    : checkMotion middleBars \(\equiv\) []
+    motionOk    : checkMotion (firstBar :: middleBars) \(\equiv\) []
     endingOk    : checkEnding middleBars lastBar \(\equiv\) nothing
@@
 Here we go through the rules without showing the corresponding
-Agda functions, as they are largely similar to what we have seen
-before.
+Agda functions, as they are largely similar to what we defined for
+first species.
 
@@
 
-The beginning of the music is \emph{not} allowed to be the unison,
-as it prevents the listener from recognizing the beginning of the
-counterpoint line.
+The beginning of the music may be either the 5th or 8th, but
+\emph{not} the unison, as it prevents the listener from recognizing
+the beginning of the counterpoint line.
 % We restrict the first interval by defining a new error
@@
 
-Weak beats are constrained in a looser manner.
+Weak beats are constrained in a looser manner than strong beats.
 First, they are allowed to be dissonant if they are created by a
@@
 
+\paragraph{Ending}
+
+The last interval must again be the unison or 8th, preceded by an
+appropriate interval that constitutes a cadence structure.
+
 \subsubsection{Putting Things Together}
@@
 As in \texttt{FirstSpecies}, we have three fields holding the musical
-content, followed by five fields carrying the proofs of the properties
-disucssed above.
+content, followed by five fields carrying the proofs of the required
+properties disucssed above\footnote{
+  As the reader would have imagined, each \texttt{checkXXX} function
+  corresponds to one rule that must be satisfied by the counterpoint.
+  The auxiliary function \texttt{secondPitch} in \texttt{weakBeatsOk}
+  extracts the counterpoint note of the last interval, and
+  \texttt{expandPitchIntervals2} turns a list of \texttt{PitchInterval2}s
+  into a list of \texttt{PitchInterval}s.
+}.
 
@@
     motionOk      : checkMotion2 (firstBar ::
-                                  (expandPitchIntervals2 middleBars) ++
-                                  (lastBar :: [])) \(\equiv\) []
+                                  (expandPitchIntervals2 middleBars)) \(\equiv\) []
     endingOk      : checkEnding2 middleBars lastBar \(\equiv\) nothing
@@
 \end{alltt}
+
+\subsection{Comparison with Previous Work}
+\label{sec:cp:comp}
+
+\citet{cong-cp} implement correct-by-construction counterpoint
+using an older version of Music Tools.
+The idea is to represent counterpoint as a list-like datatype,
+where the base cases enforce a valid ending and the inductive case
+guarantees correct uses of intervals and motion (by means of
+implicit arguments representing proofs).
+
+Our representation of counterpoint improves on the existing
+implementation in two ways.
+First, while \citet{cong-cp} rely on the Agda type checker to report
+errors in the counterpoint, we use our own type checker (implemented
+as the \texttt{checkXXX} functions) and type errors (defined as the
+\texttt{XXXError} datatypes).
+This allows us to produce error messages that clearly state what
+kind of mistake the programmer has made.
+Suppose we have replaced the second interval of Figure~\ref{fig:fs} by
+an octave (8th) of D, breaking the rule that perfect intervals cannot
+be approached by parallel motion.
+In our implementation, this results in an error in the
+\texttt{motionOk} field, which requires a proof of
+\texttt{checkMotion (firstBar :: middleBars) \(\equiv\) []}:
+
+\begin{alltt}
+(parallel (c 5 , per8) (d 5 , per8) :: []) != [] of type (List MotionError)
+\end{alltt}
+
+\noindent In \citet{cong-cp}'s implementation, on the other hand,
+the error results in the first \texttt{PitchInterval} being
+highlighted in yellow, with the following error message displayed
+in the Agda buffer:
+
+\begin{alltt}
+\_13 : motionOk (c 5 , c 6) (d 5 , d 6)
+\end{alltt}
+
+\noindent This indicates the existence of an unsolved metavariable.
+To a non-expert user, it may be difficult to connect this message to
+the musical error.
+
+The second improvement from \citet{cong-cp} is that, instead of
+incorporating all the rules into the counterpoint datatype, we define
+each of them as a separate function.
+This makes it possible to reuse certain rules in a different context,
+as we will see in the next section.

### 2020-03-01 doc/icfp20/conclusion.tex
wip

@@
-\section{Conclusion}
+\section{Conclusion and Future Work}
 \label{sec:conclusion}
 
-Text here.
+We believe that functional programming, and in particular functional
+programming with dependent types, is the most promising way to do software
+engineering going forward. The languages and tools are perhaps not yet
+as finely polished and rich as those typically used in production
+software, but they are fundamentally far more powerful, and thus once
+the infrastructure (and training of programmers) catches up should
+become the languages of choice.
+
+Music is an especially valuable domain in which to explore these
+ideas. We have found it to be a microcosm of issues that arise
+everywhere in software engineering: modularity, composability,
+correctness, and data issues such as handling equivalent formulations of
+the same data as well slightly different formulations. If we can
+demonstrate that functional programming with dependent types works
+well in the domain of music, the same techniques can be used in
+software in general.
+
+Data issues are particularly prevalent at API boundaries between
+systems, which may internally represent the data in different formats
+(satisfying modularity and abstraction) but then rely on tedious and
+potentially error-prone conversions between the formats. In music
+we see this in wanting to work sometimes with absolute pitch and others
+with relative pitch (a pair of an octave and a pitch within that
+octave), or choosing between
+a pair of pitches and a \texttt{PitchInterval}. Also we would like to
+reuse a function that just works on a pitch to work on a note, which
+consists of a pitch plus a duration.
+
+For now we have explicity converted between equivalent representations
+and explictly lifted functions, to get a feel for the amount of work
+required.  However in the world of dependent types there are known
+research techniques for handling these cases, namely transporting
+across equivalences as in Homotopy Type Theory~\citep{hottbook} and
+ornaments~\citep{dagand-ornaments}. Cubical
+Agda~\citep{vezzosi-cubical}, although still early in development,
+should be an excellent tool for exploring the extent to which these
+research ideas can be applied in a practical context.
+
+Aside from these more fundamental ideas there is simply the work of
+further extending the formalization of music theory to encompass
+harmonic analysis, more advanced counterpoint, voice leading, and even
+composition~\citep{schoenberg-fundamentals}. The results so far have
+been encouraging, but much more needs to be done to find the right
+abstractions to express the rules in the simplest, clearest and most
+composeable way possible.
+We expect the world of music theory to worthy of a long exploration.

### 2020-03-01 doc/icfp20/intro.tex

@@
 To describe rules of basic harmonic structure,
-\citet{magalhaes-harmtrace} and its successors use dependent
-types. (TODO: Give an example.) However Haskell currently has limited
-support for dependent types, and requires many extentions and tricks
-such as the use of singleton types~\citep{eisenberg-singleton} to
-simulate a restricted form of dependent types. (TODO: fix wording)
+\citet{magalhaes-harmtrace} and its successors use dependent types,
+for example to index chords by major or minor mode.  However Haskell
+currently has limited support for dependent types, and requires many
+extentions and tricks such as the use of singleton
+types~\citep{eisenberg-singleton}.
 
@@
 from music samples, using for example machine
-learning~\citep{huang-cp}. However.... TODO
-TODO: Perhaps create rules and weight them, then use ML.
+learning~\citep{huang-cp}. However there is the danger that since
+the rules were never explicit they could end up being violated with
+poor results; for example parallel fifths and octaves that are
+extremely rare (and almost always due to special allowed
+circumstances) in Bach chorales used as training data resulted
+in a much higher frequency of their use in generated
+harmonizations~\citep{roberts-cp}.
 
-Although formalization of a large amount of music theory is a goal,
-for now we start with the modest task of expressing a small,
+Ultimately it seems some kind of mixture of rules and statistics may
+be the best way to faithfully capture an effective theory of
+music. One simple approach would be to develop a logic of rules and
+their interactions, but attach weights to the rules which could be
+determined via machine learning techniques.
+
+Our long-term goal is formalization of a large amount of music theory.
+As a first step we start with the modest task of expressing a small,
 relatively strict rule set known as species
-counterpoint~\cite{fux-cp}. TODO--more here
+counterpoint~\cite{fux-cp}, intended for combining interdependent
+melody lines to produce a pleasant-sounding result.
+
+We make use of and extend~\citet{MusicTools}, a library of small tools
+for analysis and synthesis of music written in Agda, with the goal of
+eventually being a dependently-typed alternative to Euterpea.
+Previous work by \citet{cong-cp} using Music Tools expressed
+first-species counterpoint with the Agda type system, so that an error
+in counterpoint resulted in an Agda type error. Although this takes
+advantage of Agda's native type checking, it has several
+downsides. One is that the Agda errors may be difficult to interpret
+as the corresponding musical error. Another is that it is difficult to
+keep the types both simple and flexible.
+
+In this work we present an alternative representation of counterpoint
+using what could be considered a custom musical type checker written
+in Agda. This allows us to easily describe fine-grained rules, write
+custom type errors, and then combine subsets of rules together for
+larger-scale checking. Music can be easily combined with rules it
+satisfies as a dependent record type. This record represents a proof
+certificate that the music follows the asserted rules.
+Rules can also easily be used in
+other contexts, satisfying modularity and composability.
+
 We also present preliminary work on harmonizing a melody using
@@
 
-We make use of and extend~\citet{MusicTools}, a library of small tools
-for analysis and synthesis of music written in Agda, with the goal
-of eventually being a dependently-typed alternative to Euterpea. 
-Previous work by ~\citet{cong-cp} using Music Tools expressed
-first-species counterpoint with the Agda type system, so that an error
-in counterpoint resulted in an Agda type error. Although this takes
-advantage of Agda's native type checking, it has several
-downsides. One is that the Agda errors may be difficult to interpret
-as the corresponding musical error. Another is TODO: flexibility and
-composability.
-
 This is an experience report, and we highlight both the advantages and
-disadvantages of using Agda. TODO: MORE
-
-TODO: Talk about conclusion.
-
-TODO: check that bib style is allowed.
+disadvantages of using Agda for music theory. On one hand the
+expressiveness of dependent types makes it easy and natural to
+describe music theory rules. However we find the emphasis on proof
+construction and particularly the extra work needed for decidable
+equality can add an extra burden which is not always welcome. However
+overall we feel the positives far outweigh the negatives, and in the
+final section we describe future work which may help reduce the tedium.

### 2020-03-01 doc/icfp20/counterpoint.tex
typo

@@
 Figure~\ref{fig:fs} is correct, since the music is an inhabitant of
-the \texttt{FIrstSpecies} type.
+the \texttt{FirstSpecies} type.
 

### 2020-03-01 doc/icfp20/counterpoint.tex
improve figures

@@
 
-\newcommand{\Motion}{
+\newcommand{\MotionCadence}{
   \begin{figure}
-    \includegraphics[width=7cm]{fig/motion1.png}
-    \hspace{8mm}
-    \includegraphics[width=3.35cm]{fig/motion2.png}
+    \includegraphics[width=7cm]{fig/motion.png}
+    \hspace{1.2cm}
+    \includegraphics[width=3.33cm]{fig/cadence.png}
     \begin{flushleft}
     \begin{small}
-      \hspace{1.55cm} contrary \ \ \ \ \ \ \ \ parallel \ \ \ \ \ \ \
-      \ \ \,similar \ \ \ \ \ \ \ \ \ oblique
-      \hspace{1.3cm} paralell-5th \ \ \ similar-8th
+      \hspace{1.2cm} contrary \ \ \ \ \ \ \ \ \,parallel \ \ \ \ \ \ \
+      \ \ \ \,similar \ \ \ \ \ \ \ \ \ \ oblique
+      %\hspace{1.3cm} paralell-5th \ \ \ similar-8th
     \end{small}
     \end{flushleft}
-    \caption{Four Kinds of Motion (left) and Invalid Patterns (right)}
-    \label{fig:motion}
+    \caption{Motion (left) and Cadence (right)}
+    \label{fig:motioncadence}
   \end{figure}
@@
 \label{sec:cp}
-
-TODO: contrast old and new methods
 
@@
 
-\Motion
+\MotionCadence
 
@@
 It is known that approaching a perfect interval by parallel or similar
-motion (as in Figure~\ref{fig:motion} right) destroys the independence
-of lines.
+motion tends to fuse the two lines of melody into one.
 To rule out such motion, we define the \texttt{checkMotion}
-function, which checks all pairs of two adjacent intervals and
-returns a list of erroneous patterns.
+function, which inspects all pairs of two adjacent intervals (using an
+auxiliary function \texttt{pairs}) and returns a list of erroneous patterns.
 
@@
 checkMotion = mapMaybe (uncurry motionCheck) \(\circ\) pairs
--- where pairs l returns a list of pairs consisting of two adjacent elements of l
 \end{alltt}
@@
 The ending of the music should express relaxation.
-Among different intervals, the unison and 8th are considered
-most stable, hence the last interval of the music must be either of
-these intervals.
+There is a concensus that the unison and 8th are the most stable
+intervals, hence the music must end with either of these intervals.
 The last interval should also be approached by a \emph{cadence},
-a progression that gives rise to a sense of resolution.
+a progression that gives rise to a sense of resolution
+(Figure~\ref{fig:motion} right).
 This in turn suggests that a valid ending requires the middle bars
@@
 We encode these rules as the \texttt{checkEnding} function, which,
-upon finding an invalid ending, tells us which rules are not satisfied.
+upon finding an invalid ending, reports one of the three possible
+errors.
 

### 2020-03-02 doc/icfp20/intro.tex
tweak intro

@@
 Edgar Var\`{e}se describes music as ``organized sound'', and
-throughout history cultures have created and applied systems of rules
+throughout history cultures have developed and applied systems of rules
 and guidelines to govern the music they create, most notably the
@@
 To describe rules of basic harmonic structure,
-\citet{magalhaes-harmtrace} and its successors use dependent types,
+\citet{magalhaes-harmtrace} and their successors use dependent types,
 for example to index chords by major or minor mode.  However Haskell
@@
 using a programming language that offers full dependent types. We use
-the language Agda~\citep{norell-phd} since it is fairly mature and
-aims to be a functional programming language as well as a proof
+Agda~\citep{norell-phd} since it is fairly mature and
+aims to be both a functional programming language and a proof
 assistant. It also features a Haskell FFI so we can take advantage of
-existing Haskell libraries, in partiular for MIDI.
+existing Haskell libraries, in partiular for MIDI generation.
 
@@
 tempting to take a standard textbook on music theory such as
-\citet{piston-harmony} or \citet{aldwell2018harmony} and attempt to
-formalize it in type theory. However this is difficult since most
+\citet{piston-harmony} or \citet{aldwell2018harmony} and 
+formalize it in type theory. However this is difficult because most
 ``rules'' in music theory are not absolute, but rather suggestions of
@@
 theory, an alternative is to attempt to indirectly deduce the rules
-from music samples, using for example machine
-learning~\citep{huang-cp}. However there is the danger that since
-the rules were never explicit they could end up being violated with
-poor results; for example parallel fifths and octaves that are
-extremely rare (and almost always due to special allowed
-circumstances) in Bach chorales used as training data resulted
-in a much higher frequency of their use in generated
-harmonizations~\citep{roberts-cp}.
+from music samples, using machine learning~\citep{huang-cp} or
+some other statistical methods. Unfortunately the set of deduced
+rules could sometimes be insufficient for generating good music.
+As an example, \citet{roberts-cp} develop the Bach Doodle
+harmonization tool using Bach chorals as training data, but still
+observe high frequency of parallel 5th and 8th that are extremely
+rare in Bach's music.
 
@@
 eventually being a dependently-typed alternative to Euterpea.
-Previous work by \citet{cong-cp} using Music Tools expressed
+Previous work by \citet{cong-cp} using Music Tools expresses
 first-species counterpoint with the Agda type system, so that an error

### 2020-03-02 doc/icfp20/intro.tex
more tweaks

@@
 
-We make use of and extend~\citet{MusicTools}, a library of small tools
+We make use of and extend Music Tools~\citep{MusicTools},
+a library of small tools
 for analysis and synthesis of music written in Agda, with the goal of
 eventually being a dependently-typed alternative to Euterpea.
-Previous work by \citet{cong-cp} using Music Tools expresses
+Previous work by \citet{cong-cp} expresses
 first-species counterpoint with the Agda type system, so that an error
@@
 certificate that the music follows the asserted rules.
-Rules can also easily be used in
-other contexts, satisfying modularity and composability.
+Rules can also be used in multiple different contexts, satisfying
+modularity and composability.
 
@@
 expressiveness of dependent types makes it easy and natural to
-describe music theory rules. However we find the emphasis on proof
+describe music theory rules. On the other hand we find the emphasis on proof
 construction and particularly the extra work needed for decidable

### 2020-03-02 doc/icfp20/intro.tex
comment out some text

@@
 
-Full dependent types allow expression of predicate logic, and it is
-tempting to take a standard textbook on music theory such as
-\citet{piston-harmony} or \citet{aldwell2018harmony} and 
-formalize it in type theory. However this is difficult because most
-``rules'' in music theory are not absolute, but rather suggestions of
-varying degrees (themselves vaguely specified) of importance. Since
-ultimately the resulting created music is more important than the
-theory, an alternative is to attempt to indirectly deduce the rules
-from music samples, using machine learning~\citep{huang-cp} or
-some other statistical methods. Unfortunately the set of deduced
-rules could sometimes be insufficient for generating good music.
-As an example, \citet{roberts-cp} develop the Bach Doodle
-harmonization tool using Bach chorals as training data, but still
-observe high frequency of parallel 5th and 8th that are extremely
-rare in Bach's music.
+[NOTE: I've commented out the discussion on explicit / implicit
+handling of rules, to make the flow smoother.]
 
-Ultimately it seems some kind of mixture of rules and statistics may
-be the best way to faithfully capture an effective theory of
-music. One simple approach would be to develop a logic of rules and
-their interactions, but attach weights to the rules which could be
-determined via machine learning techniques.
+% Full dependent types allow expression of predicate logic, and it is
+% tempting to take a standard textbook on music theory such as
+% \citet{piston-harmony} or \citet{aldwell2018harmony} and 
+% formalize it in type theory. However this is difficult because most
+% ``rules'' in music theory are not absolute, but rather suggestions of
+% varying degrees (themselves vaguely specified) of importance. Since
+% ultimately the resulting created music is more important than the
+% theory, an alternative is to attempt to indirectly deduce the rules
+% from music samples, using machine learning~\citep{huang-cp} or
+% some other statistical methods. Unfortunately the set of deduced
+% rules could sometimes be insufficient for generating good music.
+% As an example, \citet{roberts-cp} develop the Bach Doodle
+% harmonization tool using Bach chorals as training data, but still
+% observe high frequency of parallel 5th and 8th that are extremely
+% rare in Bach's music.
 
-Our long-term goal is formalization of a large amount of music theory.
+% Ultimately it seems some kind of mixture of rules and statistics may
+% be the best way to faithfully capture an effective theory of
+% music. One simple approach would be to develop a logic of rules and
+% their interactions, but attach weights to the rules which could be
+% determined via machine learning techniques.
+
+Our long-term goal is to formalize a large amount of music theory.
 As a first step we start with the modest task of expressing a small,
@@
 melody lines to produce a pleasant-sounding result.
+Roughly speaking, what we do is to write a custom musical type
+checker in Agda.
+Compared to the previous work by \citet{cong-cp}, where rules are
+expressed directly as Agda types, our approach makes it easier to
+describe fine-grained rules, produce readable error messages, and
+add or drop rules depending on the circumstances.
 
-We make use of and extend Music Tools~\citep{MusicTools},
-a library of small tools
-for analysis and synthesis of music written in Agda, with the goal of
-eventually being a dependently-typed alternative to Euterpea.
-Previous work by \citet{cong-cp} expresses
-first-species counterpoint with the Agda type system, so that an error
-in counterpoint resulted in an Agda type error. Although this takes
-advantage of Agda's native type checking, it has several
-downsides. One is that the Agda errors may be difficult to interpret
-as the corresponding musical error. Another is that it is difficult to
-keep the types both simple and flexible.
+[NOTE: I've squeezed the description of our new counterpoint
+implementation.]
 
-In this work we present an alternative representation of counterpoint
-using what could be considered a custom musical type checker written
-in Agda. This allows us to easily describe fine-grained rules, write
-custom type errors, and then combine subsets of rules together for
-larger-scale checking. Music can be easily combined with rules it
-satisfies as a dependent record type. This record represents a proof
-certificate that the music follows the asserted rules.
-Rules can also be used in multiple different contexts, satisfying
-modularity and composability.
+% We make use of and extend Music Tools~\citep{MusicTools},
+% a library of small tools
+% for analysis and synthesis of music written in Agda, with the goal of
+% eventually being a dependently-typed alternative to Euterpea.
+% Previous work by \citet{cong-cp} expresses
+% first-species counterpoint with the Agda type system, so that an error
+% in counterpoint results in an Agda type error. Although this takes
+% advantage of Agda's native type checking, it has several
+% downsides. One is that the Agda errors may be difficult to interpret
+% as the corresponding musical error. Another is that it is difficult to
+% keep the types both simple and flexible.
+
+% In this work we present an alternative representation of counterpoint
+% using what could be considered a custom musical type checker written
+% in Agda. This allows us to easily describe fine-grained rules, write
+% custom type errors, and then combine subsets of rules together for
+% larger-scale checking. Music can be combined with rules it
+% satisfies as a dependent record type. This record represents a proof
+% certificate that the music follows the asserted rules.
+% Rules can also be used in multiple different contexts, satisfying
+% modularity and composability.
 
@@
 a subset of rules based on \citet{piston-harmony}, contrasting with
-existing work by \citet{koops-fharm}. Notably, since counterpoint and
+existing work by \citet{koops-fharm}. Since counterpoint and
 harmony are not separate concepts but in fact deeply intertwined, we
-are able to reuse counterpoint rules to help develop natural-sounding
-harmonizations.
+would wish to reuse counterpoint rules to develop natural-sounding
+harmonizations.  We show that, with our custom type checker, it is
+straightforward to reuse rules in different contexts.  Put differently,
+our representation of rules satisfies modularity.
+
+The rest of this paper structured as follows.  Section~\ref{sec:music}
+provides a brief glossary of musical terms, to prepare the
+reader with no familiarity in music theory.  Sections~\ref{sec:cp} and
+\ref{sec:harmony} describe our formalization of counterpoint and
+harmony, focusing on the composable and modular aspects of the
+proposed approach.  Lastly, Section~\ref{sec:conclusion} concludes
+the paper with future perspectives.
 
@@
 overall we feel the positives far outweigh the negatives, and in the
-final section we describe future work which may help reduce the tedium.
+final section we discuss possible ways to reduce the tedium.

### 2020-03-02 doc/icfp20/intro.tex

@@
 
-[NOTE: I've commented out the discussion on explicit / implicit
-handling of rules, to make the flow smoother.]
-
-% Full dependent types allow expression of predicate logic, and it is
-% tempting to take a standard textbook on music theory such as
-% \citet{piston-harmony} or \citet{aldwell2018harmony} and 
-% formalize it in type theory. However this is difficult because most
+% However this is difficult because most
 % ``rules'' in music theory are not absolute, but rather suggestions of
@@
 
-Our long-term goal is to formalize a large amount of music theory.
-As a first step we start with the modest task of expressing a small,
-relatively strict rule set known as species
+Full dependent types allow expression of predicate logic, and it is
+tempting to take a standard textbook on music theory such as
+\citet{piston-harmony} or \citet{aldwell2018harmony} and formalize it
+in type theory.  As a first step we start with the modest task of
+expressing a small, relatively strict rule set known as species
 counterpoint~\cite{fux-cp}, intended for combining interdependent
-melody lines to produce a pleasant-sounding result.
-Roughly speaking, what we do is to write a custom musical type
-checker in Agda.
+melody lines to produce a pleasant-sounding result.  Roughly speaking,
+what we do is to write a custom musical type checker in Agda.
 Compared to the previous work by \citet{cong-cp}, where rules are
 expressed directly as Agda types, our approach makes it easier to
-describe fine-grained rules, produce readable error messages, and
-add or drop rules depending on the circumstances.
+describe fine-grained rules, produce readable error messages, and add
+or drop rules depending on the circumstances.
 
-[NOTE: I've squeezed the description of our new counterpoint
-implementation.]
+% [NOTE: I've squeezed the description of our new counterpoint
+% implementation.]
 

### 2020-03-02 doc/icfp20/counterpoint.tex
make anonymous and add CCS concepts
minor fixes

@@
     \caption{First Species Counterpoint}
+    \Description{First Species Counterpoint}
     \label{fig:fs}
@@
     \caption{Motion (left) and Cadence (right)}
+    \Description{Motion (left) and Cadence (right)}
     \label{fig:motioncadence}
@@
     \caption{Second Species Counterpoint}
+    \Description{Second Species Counterpoint}
     \label{fig:ss}
@@
 i.e., the way one interval moves to another interval.
-As Figure~\ref{fig:motion} left shows, there are four kinds of motion:
+As Figure~\ref{fig:motioncadence} left shows, there are four kinds of motion:
 \emph{contrary} (two lines go in different directions),
@@
 a progression that gives rise to a sense of resolution
-(Figure~\ref{fig:motion} right).
+(Figure~\ref{fig:motioncadence} right).
 This in turn suggests that a valid ending requires the middle bars

### 2020-03-02 doc/icfp20/counterpoint.tex
proof read sections 1 - 3.2

@@
 In first species, we set one note against each note in the cantus firmus,
-which is required to start with a tonic (the first note of a scale;
-e.g., C in the case of C major) and consist only of whole notes.
+which is required to start with a tonic (the first note of a scale,
+such as C in C major) and consist only of whole notes.
 Figure~\ref{fig:fs} shows an example of first species counterpoint.
@@
 \begin{alltt}  
-PitchInterval : Set
-PitchInterval = Pitch \(\times\) Interval
-
-first : PitchInterval
+first : PitchInterval -- Pitch \(\times\) Interval
 first = (c 5 , per8)
@@
 are classified as perfect: the 1st (usually called the \emph{unison}),
-t5th, and 8th.
+5th, and 8th.
 The first interval of the music must then be one of these intervals.
@@
 \texttt{not158 i} when the first interval \texttt{i} is an invalid one.
-Since Agda does not have exceptions, we turn the error into an option
-value by wrapping it around the \texttt{just} constructor.
+Since Agda does not have exceptions, we turn the error into a
+\texttt{Maybe} value by wrapping it around the \texttt{just} constructor.
 
@@
 between the two lines of music.
-Therefore, the middle bars must consist of the latter four intervals.
+Therefore, the middle bars must consist only of the latter four intervals.
 We encode this rule as the \texttt{checkIntervals} function, which
@@
 intervalCheck (p , i) with isConsonant i | isUnison i
-... | false | _    = just (dissonant i)
-... | _     | true = just (unison p)
-... | _     | _    = nothing
+intervalCheck (p , i) | false | _    = just (dissonant i)
+intervalCheck (p , i) | _     | true = just (unison p)
+intervalCheck (p , i) | _     | _    = nothing
 
@@
 \emph{parallel} (two lines go in the same direction by the same
-distance), \emph{similar} (two lines go in the same direction),
-and \emph{oblique} (one line plays the same note).
+distance), \emph{similar} (two lines go in the same direction by
+different distances), and \emph{oblique} (one line plays the same note).
 It is known that approaching a perfect interval by parallel or similar
-motion tends to fuse the two lines of melody into one.
+motion tends to fuse the melodic lines into one.
 To rule out such motion, we define the \texttt{checkMotion}
@@
 Using the encoding of music and rules we have seen so far, we define
-\texttt{FirstSpecies}, a record type inhabited by correct counterpoint.
+\texttt{FirstSpecies}, a record type inhabited by correct first species
+counterpoint.
 The first three fields of this record type represent the first, middle,
@@
 The last four fields stand for the proofs that the music satisfies all
-the required properties, that is:
-
-\begin{itemize}
-  \item The beginning is a perfect consonance
-  \item The middle bars have no dissonant interval or unison
-  \item The middle bars have no perfect interval approached by
-    parallel or similar motion
-  \item The ending is a unison or an octave approached by a cadence
-\end{itemize}
-
-\noindent Recall that the functions representing these rules return
-either an option value or a list of found errors.
-Therefore, the correctness proofs either take the form
-\texttt{func args $\equiv$ nothing}, or look like
-\texttt{func args $\equiv$ []}.
+the required properties.
 
@@
 Figure~\ref{fig:fs} is correct, since the music is an inhabitant of
-the \texttt{FirstSpecies} type.
+\texttt{FirstSpecies}.
 
@@
 \begin{alltt}
-PitchInterval2 : Set
-PitchInterval2 = Pitch \(\times\) Interval \(\times\) Interval
-
 first2 : PitchInterval
@@
 
-middle2 : List PitchInterval2
+middle2 : List PitchInterval2 -- Pitch \(\times\) Interval \(\times\) Interval
 middle2 =
@@
 
-Weak beats are constrained in a looser manner than strong beats.
-First, they are allowed to be dissonant if they are created by a
+Weak beats are allowed to be dissonant if they are created by a
 \emph{passing tone}, i.e., a note in the middle of two step-wise
 motions in the same direction (as in bars 4-5 of Figure~\ref{fig:ss}).
-Second, they are allowed to be the unison if they are left by step in
-the opposite direction from their approach (as in bars 5-6 of
-Figure~\ref{fig:ss}).
+They may also be the unison if they are left by step in the opposite
+direction from their approach (as in bars 5-6 of Figure~\ref{fig:ss}).
 % We implement these rules as the \texttt{checkWeakBeats} function,
@@
 
-The last interval must again be the unison or 8th, preceded by an
+The last interval must be the unison or 8th, preceded by an
 appropriate interval that constitutes a cadence structure.
@@
 properties disucssed above\footnote{
-  As the reader would have imagined, each \texttt{checkXXX} function
-  corresponds to one rule that must be satisfied by the counterpoint.
   The auxiliary function \texttt{secondPitch} in \texttt{weakBeatsOk}
-  extracts the counterpoint note of the last interval, and
+  extracts the counterpoint note of a given interval, and
   \texttt{expandPitchIntervals2} turns a list of \texttt{PitchInterval2}s

### 2020-03-02 doc/icfp20/intro.tex
proof read sections 1 - 3.2

@@
 \citet{piston-harmony} or \citet{aldwell2018harmony} and formalize it
-in type theory.  As a first step we start with the modest task of
+in type theory.  As a first step towards this goal, we start with the modest task of
 expressing a small, relatively strict rule set known as species
@@
 The rest of this paper structured as follows.  Section~\ref{sec:music}
-provides a brief glossary of musical terms, to prepare the
-reader with no familiarity in music theory.  Sections~\ref{sec:cp} and
-\ref{sec:harmony} describe our formalization of counterpoint and
-harmony, focusing on the composable and modular aspects of the
-proposed approach.  Lastly, Section~\ref{sec:conclusion} concludes
-the paper with future perspectives.
+introduces basic musical concepts as well as their Agda representation.
+Sections~\ref{sec:cp} and \ref{sec:harmony} describe our formalization
+of counterpoint and harmony, focusing on the composable and
+modular aspects of the proposed approach.  Lastly,
+Section~\ref{sec:conclusion} concludes the paper with future perspectives.
 

### 2020-03-02 doc/icfp20/music.tex
minor fixes
proof read sections 1 - 3.2

--- 2020-02-22 doc/icfp20/music.tex
@@
-\section{Musical Preliminaries}
+\newcommand{\Interval}{
+  \begin{figure}
+  \includegraphics[width=12cm]{fig/interval.png} \\
+  \begin{flushleft}
+    \begin{footnotesize}
+      \hspace{1.45cm} \texttt{per1 \hspace{0.5mm} min2  \hspace{2.5mm}
+        maj2 \hspace{1.2mm} min3 \hspace{0.5mm} maj3 \ per4
+        \hspace{1mm}aug4
+        \hspace{0.5mm}  per5\hspace{1.2mm}  min6\hspace{1.6mm} maj6
+        \hspace{1.6mm}min7\hspace{1.2mm} maj7\hspace{1.2mm} per8} \\
+      consonant? \hspace{1.5mm} yes \hspace{4.2mm} no \hspace{6.5mm}  no
+      \hspace{4.7mm} yes \hspace{3.8mm}  yes \hspace{3.7mm} no \hspace{3.5mm}
+      no \hspace{4.3mm} yes \hspace{3.8mm} yes \hspace{3.2mm} yes
+      \hspace{3.7mm} no \hspace{4mm} no \hspace{3.8mm} yes
+    \end{footnotesize}
+  \end{flushleft}
+  \caption{13 Kinds of Intervals}
+  \label{fig:interval}
+ \end{figure}
+}
+
+\section{Basic Musical Terms and Their Representation in Agda}
 \label{sec:music}
@@
 \paragraph{Pitches}
-A \emph{pitch} is a value denoting how high or low a sound is.
-In our implementation, pitches have type \texttt{Pitch} and take the
-form \texttt{name octave}, where \texttt{name} is a name ranging
-over \texttt{c}, \texttt{d}, \texttt{e}, etc. and \texttt{octave} is a
-natural number denoting which octave the pitch belongs to.
+
+A \emph{pitch} tells us how high or low a sound is.
+We represent pitches as natural numbers, but throughout the paper,
+we use a more readable notation \texttt{name octave}, where
+\texttt{name} ranges over \texttt{c}, \texttt{d}, \texttt{e}, etc.,
+and \texttt{octave} denotes which octave the pitch belongs to.
 For instance, the middle C is represented as \texttt{c 5}.
@@
 \paragraph{Duration}
-\emph{Duration} denotes an unspecified unit of time during which a
-sound or silence lasts.
-In our implementation, duration (of type \texttt{Duration} is simply
-a natural number, denoting the relative length within a bar (such as
-whole and half).
-When the music is played at a specific tempo, the number is 
-multiplied by a velocity and turned into an absolute length.
+
+\emph{Duration} denotes an unspecified unit of time during which
+a sound or silence lasts.
+We represent duration as a natural number, and use this value to
+calculate the absolute length when the music is played at a specific
+tempo.
+As an example, the duration constant \texttt{whole} has value
+\texttt{16} and corresponds to 4 seconds if the tempo is 60 beats
+per minute.
 
@@
 \paragraph{Notes}
+
 Combining pitches and duration gives us \emph{notes}.
-In our implementation, we represent notes as a datatype \texttt{Note}
-with two constructors: \texttt{tone} for notes with sound and
-\texttt{rest} for those without.
+We represent notes as a datatype \texttt{Note} with two constructors:
+\texttt{tone} for notes with sound and \texttt{rest} for those without.
+For example, \texttt{tone whole (c 5)} is a whole note whose sound
+is the middle C.
 
@@
 
-\emph{Intervals} are a concept denoting the difference in pitch
-between two notes.
-There are 13 kinds of interval within an octave, and these intervals
-can be classified from several different perspectives: (i) major or
-minor; (ii) consonant or dissonant; and (iii) perfect or imperfect.
+\Interval
+
+An \emph{interval} represents the difference in pitch between
+two notes.
+There are 13 kinds of interval within an octave
+(Figure~\ref{fig:interval}, and these intervals can be classified from
+several different perspectives:
+(i) major or minor; (ii) consonant or dissonant; and (iii) perfect
+or imperfect.
 We define intervals as a datatype \texttt{Interval}, where each
 constructor represents one of the 13 intervals.
-
-\begin{figure}[h]
-  \includegraphics[width=12cm]{fig/interval.png} \\
-  \begin{flushleft}
-    \begin{footnotesize}
-      \hspace{1.45cm} \texttt{per1 \hspace{0.5mm} min2  \hspace{2.5mm}
-        maj2 \hspace{1.2mm} min3 \hspace{0.5mm} maj3 \ per4
-        \hspace{1mm}aug4
-        \hspace{0.5mm}  per5\hspace{1.2mm}  min6\hspace{1.6mm} maj6
-        \hspace{1.6mm}min7\hspace{1.2mm} maj7\hspace{1.2mm} per8} \\
-      consonant? \hspace{1.5mm} yes \hspace{4.2mm} no \hspace{6.5mm}  no
-      \hspace{4.7mm} yes \hspace{3.8mm}  yes \hspace{3.7mm} no \hspace{3.5mm}
-      no \hspace{4.3mm} yes \hspace{3.8mm} yes \hspace{3.2mm} yes
-      \hspace{3.7mm} no \hspace{4mm} no \hspace{3.8mm} yes
-    \end{footnotesize}
-  \end{flushleft}
-\end{figure}

### 2020-03-03 doc/icfp20/counterpoint.tex
proof read Section 3.3; add discussion on Haskell '17

@@
 
-\citet{cong-cp} implement correct-by-construction counterpoint
-using an older version of Music Tools.
-The idea is to represent counterpoint as a list-like datatype,
-where the base cases enforce a valid ending and the inductive case
-guarantees correct uses of intervals and motion (by means of
+The type-theoretical formalization of counterpoint has previously
+been attempted by \citet{cong-cp}.
+They represent correct counterpoint as an Agda datatype akin to
+a list, where the base cases enforce a valid ending and the inductive
+case guarantees correct uses of intervals and motion (by means of
 implicit arguments representing proofs).
@@
 be approached by parallel motion.
-In our implementation, this results in an error in the
-\texttt{motionOk} field, which requires a proof of
-\texttt{checkMotion (firstBar :: middleBars) \(\equiv\) []}:
+In our implementation, this causes an error in the \texttt{motionOk}
+field, with the following message:
 
@@
 \noindent In \citet{cong-cp}'s implementation, on the other hand,
-the error results in the first \texttt{PitchInterval} being
-highlighted in yellow, with the following error message displayed
-in the Agda buffer:
+the invalid motion results in an unsolved metavariable \texttt{\_13},
+representing the proof that the two intervals form a valid motion:
 
 \begin{alltt}
-\_13 : motionOk (c 5 , c 6) (d 5 , d 6)
+\_13 : motionOk (c 5 , per8) (d 5 , per8)
 \end{alltt}
 
-\noindent This indicates the existence of an unsolved metavariable.
-To a non-expert user, it may be difficult to connect this message to
-the musical error.
+\noindent Compared to the error message produced by our
+implementation, this lacks the exact information about the motion
+between the two intervals.
+
+We would also like to note that the attempt to precisely reporting
+musical errors has appeared in the work by \citet{szamozvancev-welltyped}.
+They formalize various musical rules, including those for intervals
+and motion, using GHC's custom type error feature.
+However, they do not aim to fully support species counterpoint,
+and thus do not cover as many rules as we do.
 
 The second improvement from \citet{cong-cp} is that, instead of
-incorporating all the rules into the counterpoint datatype, we define
-each of them as a separate function.
-This makes it possible to reuse certain rules in a different context,
-as we will see in the next section.
+incorporating all the rules into a counterpoint datatype, we define
+each of them as a separate function and combine them using a
+record type.
+This makes it easy to switch between strict and relaxed rule sets:
+we just need to add or remove certain fields.
+It also allows us to reuse some of the rules in a context other than
+counterpoint, as we will see in the next section.

### 2020-03-03 doc/icfp20/music.tex
add figure desciption and remove CCSXML
proof read Section 3.3; add discussion on Haskell '17

@@
   \caption{13 Kinds of Intervals}
+  \Description{13 Kinds of Intervals}
   \label{fig:interval}
@@
 There are 13 kinds of interval within an octave
-(Figure~\ref{fig:interval}, and these intervals can be classified from
+(Figure~\ref{fig:interval}), and these intervals can be classified from
 several different perspectives:

### 2020-03-03 doc/icfp20/cpharmony.tex
start harmony

\section{Representations of Counterpoint and Harmony}
\label{sec:cpharmony}

Harmony refers to simultaneously sounding tones (chords) and harmonic progression
a series of chords. It is in essence the of dual of counterpoint. A
fundamental representation used in Music Tools is a grid of
\texttt{Point}s, where a \texttt{Point} represents either a pitch, a
hold of a previous pitch, or a rest.

\begin{alltt}
data Point : Set where
  tone : Pitch \(\rightarrow\) Point
  hold : Pitch \(\rightarrow\) Point
  rest : Point
\end{alltt}

Music is thus considered to be quantized to some minimal duration,
with the horizontal dimension representing time and the vertical
dimension representing pitch. A slice in the horizontal direction is a
\texttt{Melody}, represented as a vector of points of some length $n$. Similarly a slice
in the vertical direction is a \texttt{Chord}, again of some fixed length.

\begin{alltt}
data Melody (n : \(\mathbb{N}\)) : Set where
  melody : Vec Point n \(\rightarrow\) Melody n

data Chord (n : \(\mathbb{N}\)) : Set where
  chord : Vec Point n \(\rightarrow\) Chord n
\end{alltt}

One can think of the grid as several melodies in parallel of the same
duration. This is essentially counterpoint, and thus given that
name. Note the crucial use of dependent types here to ensure all
melodies have the same duration $d$. The counterpoint is specified to
have $v$ voices.

\begin{alltt}
data Counterpoint (v : \(\mathbb{N}\)) (d : \(\mathbb{N}\)): Set where
  cp : Vec (Melody d) v \(\rightarrow\) Counterpoint v d
\end{alltt}

Dually one can think of the grid as a series of chords, in another
word a harmonic progression, although this is named \texttt{Harmony}
in the library for conciseness. Again dependent types enforce every
chord has the same number of voices.

\begin{alltt}
data Harmony (v : \(\mathbb{N}\)) (d : \(\mathbb{N}\)): Set where
  harmony : Vec (Chord v) d \(\rightarrow\) Harmony v d
\end{alltt}

Converting between \texttt{Counterpoint} and \texttt{Harmony} is
simply matrix transposition. These are fundamental representations
that can be used directly for analysis, for example, and there are
functions to convert them to MIDI for sound generation.

There are many other ways to represent music, which have advantages in
certain contexts. For example in Section~\ref{sec:cp}
\texttt{PitchInterval} and \texttt{PitchInterval2} were used instead
of \texttt{Counterpoint} since they more explictly represent the
structure of the music in species counterpoint. It is straightforward
to convert between the representations, and worth using the most
appropriate type for a given situation.

### 2020-03-03 doc/icfp20/harmony.tex
wip
add diagrams
rename & combine figures
improve figures
make anonymous and add CCS concepts
start harmony

@@
+\newcommand{\Harmonization}{
+  \begin{figure}
+    \includegraphics[width=5.5cm]{fig/piston.png}
+    \hspace{1cm}
+    \includegraphics[width=5cm]{fig/fharm.png} \\
+    \begin{flushleft}
+    \begin{small}
+     \hspace{2.35cm} V \ \ \ \ \ \ \ \,I \ \ \ \ \ \ \ \,IV \,VI \ \,III \ IV \ \ \,V
+     \hspace{2.6cm} V \ \ \ \ \ \ \ \,I \ \ \ \ \ \,IV VI \,III\,IV \ \ V
+    \end{small}
+    \end{flushleft}
+    \caption{Harmonization by Piston (left) and \fharm (right)}
+    \Description{Harmonization by Piston (left) and \fharm (right)}
+    \label{fig:harmonization}
+  \end{figure}
+}
+
 \section{Harmony}
@@
 
-Text here.
+Harmony is a particularly fundamental area of music theory with numerous texts, for
+example \citet{piston-harmony} and \citet{aldwell2018harmony}, devoted
+to it. It has also been the focus of most of the Haskell-based
+work~\citep{magalhaes-harmtrace,koops-fharm,magalhaes-fcomp}, and as a
+first step into this huge area it worth trying to implement some of
+this functionality in Agda. Indeed \citet{magalhaes-harmtrace} states in its final section
+``It would be interesting to see if we could easily port our system to
+a dependently-typed setting'' and explicitly mentions Agda and Idris
+as candidates. Our work, while currently very preliminary, provides a
+glipse into what that can look like.
+
+\citet{magalhaes-harmtrace} is devoted to creating a grammar for
+harmonic analysis and a corresponding parser that takes as input a
+series of chords and tries to produce a harmonic analysis that fits
+the progression. The grammar makes crucial use of dependent types and
+is thus awkard to represent in Haskell. We have implemented those
+parts in Agda and not surprisingly they are simple and
+elegant. However \citet{magalhaes-harmtrace} also makes use of
+sophisticated Haskell libraries for parser combinators and generics,
+which are unavailable in Agda and would either have to be natively
+implemented or called through Agda's Haskell FFI, both of which involve
+considerable work, so we have not yet pursued that route.
+
+Instead we first focus on FHARM~\citep{koops-fharm}, which aims to create a
+harmonize a given melody. This is the subject of Chapter~9 of
+\citet{piston-harmony}, and, along with three exercises,
+FHARM harmonizes the main Example~9.1 of the chapter,
+shown in Figure~\ref{fig:harmonization}.
+
+\Harmonization
+
+Their techinique is to first require the melody notes be members of
+the harmonizing chords, then use their previous work to try to
+create a harmonic analysis of every possible sequence of chords that
+arises, chosing the one that has most closely fits their
+grammar. Although the result is not bad, it clearly leaves room for
+improvement. Noted in the paper is that no attenton is given to
+voicing (how each of the four parts flows horizontally and how the
+lines interact---in other words the counterpoint). Not noted is that
+the melody note is always doubled, whereas typically in four-part
+harmony it is prefered to double the root note. Also mimimizing the
+number of parse errors as a means to choose the best harmonization
+does not necessarily have a musical meaning.
+
+Not having an impelmentation of the grammar and parser we explore a
+different approach, based closer on Piston's own.
+
+TODO: More here.

### 2020-03-03 doc/icfp20/harmony.tex
tweak harmony

@@
 as candidates. Our work, while currently very preliminary, provides a
-glipse into what that can look like.
+glimpse into what that can look like.
 
-\citet{magalhaes-harmtrace} is devoted to creating a grammar for
+\citet{magalhaes-harmtrace} aim to creating a grammar for
 harmonic analysis and a corresponding parser that takes as input a
@@
 is thus awkard to represent in Haskell. We have implemented those
-parts in Agda and not surprisingly they are simple and
+parts in Agda (\texttt{Fmmh.agda}) and not surprisingly they are simple and
 elegant. However \citet{magalhaes-harmtrace} also makes use of
@@
 
-Instead we first focus on FHARM~\citep{koops-fharm}, which aims to create a
+Instead we first focus on \fharm~\citep{koops-fharm}, which aims to create a
 harmonize a given melody. This is the subject of Chapter~9 of 
 \citet{piston-harmony}, and, along with three exercises,
-FHARM harmonizes the main Example~9.1 of the chapter,
+\fharm harmonizes the main Example~9.1 of the chapter,
 shown in Figure~\ref{fig:harmonization}.
@@
 
-Their techinique is to first require the melody notes be members of
+Their techinique is to first require the melody notes to be members of
 the harmonizing chords, then use their previous work to try to
 create a harmonic analysis of every possible sequence of chords that
-arises, chosing the one that has most closely fits their
+arises, chosing the one that most closely fits their
 grammar. Although the result is not bad, it clearly leaves room for
-improvement. Noted in the paper is that no attenton is given to
+improvement. Noted in the paper is that no attention is given to
 voicing (how each of the four parts flows horizontally and how the
@@
 Not having an impelmentation of the grammar and parser we explore a
-different approach, based closer on Piston's own.
+different approach, based closer on Piston's.
 

### 2020-03-03 doc/icfp20/conclusion.tex
fix typos

@@
 
-For now we have explicity converted between equivalent representations
-and explictly lifted functions, to get a feel for the amount of work
+For now we have explicitly converted between equivalent representations
+and explicitly lifted functions, to get a feel for the amount of work
 required.  However in the world of dependent types there are known
@@
 abstractions to express the rules in the simplest, clearest and most
-composeable way possible.
-We expect the world of music theory to worthy of a long exploration.
+composable way possible.
+We expect the world of music theory to be worthy of a long exploration.

### 2020-03-03 doc/icfp20/counterpoint.tex
fix typos

@@
 
-The middle bars of the music should maintain harmonical consonance
+The middle bars of the music should maintain consonance of intervals
 and independence of melodic lines.
@@
 We encode this rule as the \texttt{checkIntervals} function, which
-returns a list of errors correponding to the occurrences of dissonant
+returns a list of errors corresponding to the occurrences of dissonant
 intervals and unisons.
@@
 The ending of the music should express relaxation.
-There is a concensus that the unison and 8th are the most stable
+There is a consensus that the unison and 8th are the most stable
 intervals, hence the music must end with either of these intervals.
@@
 content, followed by five fields carrying the proofs of the required
-properties disucssed above\footnote{
+properties discussed above\footnote{
   The auxiliary function \texttt{secondPitch} in \texttt{weakBeatsOk}

### 2020-03-03 doc/icfp20/harmony.tex
fix typos

@@
 the progression. The grammar makes crucial use of dependent types and
-is thus awkard to represent in Haskell. We have implemented those
+is thus awkward to represent in Haskell. We have implemented those
 parts in Agda (\texttt{Fmmh.agda}) and not surprisingly they are simple and
@@
 
-Their techinique is to first require the melody notes to be members of
+Their technique is to first require the melody notes to be members of
 the harmonizing chords, then use their previous work to try to
 create a harmonic analysis of every possible sequence of chords that
-arises, chosing the one that most closely fits their
+arises, choosing the one that most closely fits their
 grammar. Although the result is not bad, it clearly leaves room for
@@
 the melody note is always doubled, whereas typically in four-part
-harmony it is prefered to double the root note. Also mimimizing the
+harmony it is preferred to double the root note. Also minimizing the
 number of parse errors as a means to choose the best harmonization
@@
 
-Not having an impelmentation of the grammar and parser we explore a
+Not having an implementation of the grammar and parser we explore a
 different approach, based closer on Piston's.

### 2020-03-03 doc/icfp20/intro.tex
fix typos

@@
 Starting in the 20th century, computers have become ubiquitous in
-music in every area, including sound sythesis, composition and
+music in every area, including sound synthesis, composition and
 production~\citep{roads-tutorial}. In terms of music theory, there has
@@
 assistant. It also features a Haskell FFI so we can take advantage of
-existing Haskell libraries, in partiular for MIDI generation.
+existing Haskell libraries, in particular for MIDI generation.
 

### 2020-03-03 doc/icfp20/conclusion.tex
the most promising way -> a promising way

@@
 We believe that functional programming, and in particular functional
-programming with dependent types, is the most promising way to do software
+programming with dependent types, is a promising way to do software
 engineering going forward. The languages and tools are perhaps not yet

### 2020-03-03 doc/icfp20/counterpoint.tex
prepare code for submission

@@
 motionCheck : PitchInterval \(\rightarrow\) PitchInterval \(\rightarrow\) Maybe MotionError
-motionCheck i1 i2 with motion i1 i2 | isPerfect (proj\(\sb{\mathtt{2}}\) i2)
+motionCheck i1 i2 with motion i1 i2 | isPerfect (proj\(\sb{2}\) i2)
 motionCheck i1 i2 | contrary | \_     = nothing
@@
 fs : FirstSpecies
-fs = firstSpecies first1 middle1 last1 refl refl refl refl
+fs = firstSpecies first middle last refl refl refl refl
 \end{alltt}
@@
 the invalid motion results in an unsolved metavariable \texttt{\_13},
-representing the proof that the two intervals form a valid motion:
+meaning that Agda failed to construct a proof required by the
+counterpoint constructor:
 
@@
 
-\noindent Compared to the error message produced by our
-implementation, this lacks the exact information about the motion
-between the two intervals.
-
-We would also like to note that the attempt to precisely reporting
-musical errors has appeared in the work by \citet{szamozvancev-welltyped}.
-They formalize various musical rules, including those for intervals
-and motion, using GHC's custom type error feature.
-However, they do not aim to fully support species counterpoint,
-and thus do not cover as many rules as we do.
+\noindent For a non-expert user, it may be difficult to connect
+this message to the corresponding musical error \footnote{
+  The attempt to produce customized musical errors has also
+  appeared in \citet{szamozvancev-welltyped}.
+  They formalize various musical rules, including those for
+  intervals and motion, using GHC's custom type error feature.
+  However, their error messages are not as precise as ours:
+  e.g., for the parallel octave example, they simply report
+  ``\texttt{Direct motion into a perfect octave is forbidden}'',
+  without the precise information about the two intervals and
+  the motion between them.}.
 

### 2020-03-03 doc/icfp20/music.tex
add piston example to Main.agda; explain chords

@@
 \label{sec:music}
+
+[TODO] Maybe need to explain degrees, instead of duration and notes.
 
@@
 constructor represents one of the 13 intervals.
+
+\paragraph{Chords}
+
+A \emph{chord} refers to a set of simultaneously sounding pitches.
+The most common pattern of chords is called \emph{triads},
+consisting of three notes separated by (major or minor) 3rd intervals.
+Chords are usually denoted by roman numerals representing the
+lowest note (the \emph{root}).
+We define triads as a datatype \texttt{Triad} with constructors
+\texttt{I}, \texttt{II}, etc., and obtain their constituent notes
+(packaged into a three-element vector) via a function
+\texttt{triadDegrees}.

### 2020-03-03 doc/icfp20/counterpoint.tex
improve glossary

@@
 and independence of melodic lines.
-As we saw before, consonant intervals include the unison,
-3rd\footnote{We use ``3rd'' to mean both the major and minor variants
-of the 3rd interval, and similarly for the 6th.}, 5th, 6th, and 8th,
-and among these, the unison is clearly an obstacle to distinguishing
-between the two lines of music.
+As we saw in Figure~\ref{fig:interval}, consonant intervals include
+the unison, 3rd, 5th, 6th, and 8th, and among these, the unison is
+clearly an obstacle to distinguishing between the two lines of music.
 Therefore, the middle bars must consist only of the latter four intervals.

### 2020-03-03 doc/icfp20/music.tex
improve glossary

--- 2020-03-03 doc/icfp20/music.tex
@@
 or imperfect.
-We define intervals as a datatype \texttt{Interval}, where each
-constructor represents one of the 13 intervals.
+We represent intervals as natural numbers, but in this paper, we
+refer to them using more informative names such as \texttt{maj3}
+and \texttt{per8}.
+
+\paragraph{Chords and Triads}
+
+A \emph{chord} refers to a set of simultaneously sounding notes.
+In this paper we are particularly interested in \emph{triads},
+which are chords consisting of three notes separated by the 3rd
+intervals (i.e., \texttt{maj3} and \texttt{min3}).
+We define triads as a datatype \texttt{Triad} with roman numeral
+constructors representing the lowest note (called the \emph{root}).
+For instance, \texttt{I} in C major represents the triad C-E-G.

### 2020-03-03 doc/icfp20/counterpoint.tex
improve comparison section

@@
 
-The type-theoretical formalization of counterpoint has previously
-been attempted by \citet{cong-cp}.
-They represent correct counterpoint as an Agda datatype akin to
-a list, where the base cases enforce a valid ending and the inductive
-case guarantees correct uses of intervals and motion (by means of
-implicit arguments representing proofs).
+The type-theoretical formalization of counterpoint has first
+been attempted by \citet{szamozvancev-welltyped}.
+They encode the rules for intervals and motion by declaring
+type classes \texttt{ValidHarmInterval} and \texttt{ValidMotion},
+and specifying which instances are invalid using GHC's support
+for custom type errors.
 
-Our representation of counterpoint improves on the existing
-implementation in two ways.
-First, while \citet{cong-cp} rely on the Agda type checker to report
-errors in the counterpoint, we use our own type checker (implemented
-as the \texttt{checkXXX} functions) and type errors (defined as the
-\texttt{XXXError} datatypes).
-This allows us to produce error messages that clearly state what
-kind of mistake the programmer has made.
+The availability of the custom error feature is an advantage of using
+Haskell, but the language's type system does not allow producing
+detailed error messages.
 Suppose we have replaced the second interval of Figure~\ref{fig:fs} by
@@
 
-\noindent In \citet{cong-cp}'s implementation, on the other hand,
-the invalid motion results in an unsolved metavariable \texttt{\_13},
-meaning that Agda failed to construct a proof required by the
-counterpoint constructor:
+\noindent Observe that the message clearly tells us which intervals
+form what kind of invalid motion.
+
+In Szamozvancev and Gale's implementation, the same example
+results in the following error:
+
+\begin{alltt}
+Direct motion into a perfect octave is forbidden.
+\end{alltt}
+
+\noindent The lack of the precise information about the intervals
+and motion is due to the absence of native dependent types in
+Haskell.
+
+More recently, \citet{cong-cp} give an implementation of first
+species counterpoint using dependent types in Agda.
+Their idea is to represent correct counterpoint as a list-like
+datatype, where the base cases enforce a valid ending and the
+inductive case guarantees correct uses of intervals and motion
+(by means of implicit arguments representing proofs).
+
+Our representation of counterpoint improves on their
+implementation in two ways.
+First, while \citet{cong-cp} rely on the Agda type checker to report
+errors in the counterpoint, we use our own type checker (implemented
+as the \texttt{checkXXX} functions) and type errors (defined as the
+\texttt{XXXError} datatypes).
+This allows us to produce helpful error messages while keeping
+types simple.
+Indeed, for the parallel octave example discussed above,
+\citet{cong-cp} would just report the existence of an unsolved
+metavariable \texttt{\_13}, meaning that Agda failed to construct
+a proof required by the counterpoint constructor:
 
@@
 \noindent For a non-expert user, it may be difficult to connect
-this message to the corresponding musical error \footnote{
-  The attempt to produce customized musical errors has also
-  appeared in \citet{szamozvancev-welltyped}.
-  They formalize various musical rules, including those for
-  intervals and motion, using GHC's custom type error feature.
-  However, their error messages are not as precise as ours:
-  e.g., for the parallel octave example, they simply report
-  ``\texttt{Direct motion into a perfect octave is forbidden}'',
-  without the precise information about the two intervals and
-  the motion between them.}.
+this message to the corresponding musical error.
 

### 2020-03-04 doc/icfp20/harmony.tex
major work on harmony section

@@
 
+\newcommand{\HarmonizationMT}{
+  \begin{figure}
+    \includegraphics[width=5.5cm]{fig/piston.png}
+    \hspace{1cm}
+    \includegraphics[width=5cm]{fig/fharm.png} \\
+    \begin{flushleft}
+    \begin{small}
+     \hspace{2.35cm} V \ \ \ \ \ \ \ \,I \ \ \ \ \ \ \ \,IV \,VI \ \,III \ IV \ \ \,V
+     \hspace{2.6cm} V \ \ \ \ \ \ \ \,I \ \ \ \ \ \,IV VI \,III\,IV \ \ V
+    \end{small}
+    \end{flushleft}
+    \caption{Harmonization by Music Tools (to be replaced by real figure)}
+    \Description{Harmonization by Music Tools}
+    \label{fig:harmonizationMT}
+  \end{figure}
+}
+
 \section{Harmony}
@@
 
-Harmony is a particularly fundamental area of music theory with numerous texts, for
-example \citet{piston-harmony} and \citet{aldwell2018harmony}, devoted
-to it. It has also been the focus of most of the Haskell-based
-work~\citep{magalhaes-harmtrace,koops-fharm,magalhaes-fcomp}, and as a
-first step into this huge area it worth trying to implement some of
-this functionality in Agda. Indeed \citet{magalhaes-harmtrace} states in its final section
-``It would be interesting to see if we could easily port our system to
-a dependently-typed setting'' and explicitly mentions Agda and Idris
-as candidates. Our work, while currently very preliminary, provides a
-glimpse into what that can look like.
+Harmony refers to simultaneously sounding tones (chords) and its study
+is primarily concerned with the progression of chords throughout a
+piece of music. It is in essence the of dual of counterpoint. In this
+section we formalize a small part of the classic text
+\citet{piston-harmony}, in particular part of Chapter~9 which is
+concerned with techniques for harmonizing a given melody.
 
-\citet{magalhaes-harmtrace} aim to creating a grammar for
-harmonic analysis and a corresponding parser that takes as input a
-series of chords and tries to produce a harmonic analysis that fits
-the progression. The grammar makes crucial use of dependent types and
-is thus awkward to represent in Haskell. We have implemented those
-parts in Agda (\texttt{Fmmh.agda}) and not surprisingly they are simple and
-elegant. However \citet{magalhaes-harmtrace} also makes use of
-sophisticated Haskell libraries for parser combinators and generics,
-which are unavailable in Agda and would either have to be natively
-implemented or called through Agda's Haskell FFI, both of which involve
-considerable work, so we have not yet pursued that route.
+\subsection{Harmonic Progression}
+\label{sec:harmony:prog}
 
-Instead we first focus on \fharm~\citep{koops-fharm}, which aims to create a
-harmonize a given melody. This is the subject of Chapter~9 of 
-\citet{piston-harmony}, and, along with three exercises,
-\fharm harmonizes the main Example~9.1 of the chapter,
-shown in Figure~\ref{fig:harmonization}.
+The primary Example~9.1 in Piston's text is shown on the left
+side of Figure~\ref{fig:harmonization} which shows the melody with a
+sample four part SATB harmonization in C~major. The melody is the
+series of highest notes, the soprano (S) line, while the other four
+parts alto (A), tenor (T) and bass (B) are the melodies below
+it. Roman numerals denote the names of the triads at each vertical
+slice. Note that even though there are four voices there are only
+three distinct pitches sounding at each vertical slice (where tones an
+octave apart are considered to have the same relative pitch). As is
+typical the root of the triad is always doubled. Also note that all
+triads are in root position (meaning the root is in the bass voice),
+as Piston restricts to this case initially in the chapter (he does
+relax it later). The goal of a harmonization is both to create an
+aesthetically pleasing chord progression as well as fluid melodies in
+each voice (called voice leading) and independence of each music
+line (in other words good counterpoint). In this chapter as well as
+throughout the text Piston gives advice on how to achieve these goals.
 
 \Harmonization
+
+We would like to formalize some of this advice and apply it to
+automate harmonization of melodies. We start with selection of the
+sequence of triads, in other words the harmonic progression. Earlier
+in the text on page~23 Piston presents a Table of Usual Root
+Prrogressions, noting which triads usually follow a given one and
+which are sometimes used or rare. It is easy to encode this in Agda.
+
+\begin{alltt}
+data Triad : Set where
+  I : Triad; II : Triad; III : Triad; IV : Triad; V : Triad; VI : Triad; VII : Triad
+
+record NextTriad : Set where
+  constructor nextTriadb
+  field
+    usual : List Triad; sometimes : List Triad; rare : List Triad
+
+rootProgression : Triad \(\rightarrow\) NextTriad
+rootProgression I   = nextTriad (IV :: V :: []) (VI :: []) (II :: III :: [])
+rootProgression II  = nextTriad (V :: []) (IV  :: VI :: []) (I :: III :: [])
+-- other cases omitted for space
+\end{alltt}
+
+It is easiest to create a harmonic progression in reverse, to
+ensure one ends a musical phrase with either a full (V--I) or half (V)
+cadence, and work backward from there. So we would like to invert this
+table to give a list of likely triads to precede a given triad. This
+would normally be straightfoward but here we see a disadvantage of
+using Agda. As is typical with languages based on dependent type
+theories, equality is a very subtle concept and even determing if a
+term is a member of a list is nontrivial.
+
+In Haskell this would be easy--there is an \texttt{Eq} typeclass and a
+method for automatically generating equalities. In the Agda Standard
+Library~\citep{agda-stdlib} is a relation with constructive proof of
+membership, and requires far more effort than we need. Alternatives
+include defining decidable equality for triads (which requires writing
+out 49 cases) or mapping to natural numbers and using already
+implemented decidable or boolean equality for those. For now we simply
+write the inverse function manually, excluding the rare cases.
+
+\begin{alltt}
+previousTriads : Triad → List Triad
+previousTriads I   = V :: IV :: VII :: []
+previousTriads II  = VI :: IV :: []
+-- other cases omitted for space
+\end{alltt}
+
+\subsection{Completing the Harmonization}
+\label{sec:harmony:complete}
+
+TODO: Finish this.
+
+See Figure~\ref{fig:harmonizationMT}, which is currently just a copy of
+the other figure, but included here now for space.
+
+\HarmonizationMT
+
+\subsection{Comparison with Previous Work}
+\label{sec:harmony:compare}
+
+Most of the the Haskell-based
+work~\citep{magalhaes-harmtrace,koops-fharm,magalhaes-fcomp} has
+focused on harmony. In particular \fharm~\citep{koops-fharm} tackles
+the problem of harmonization of a melody, and their harmonization of
+Piston's example is shown on the right side of
+Figure~\ref{fig:harmonization}. Their method is quite different from
+ours. Instead of using Piston's table of root progressions they use a
+more sophisticated hierarchical grammar from previous
+work~\citep{magalhaes-harmtrace}.
 
@@
 
-Not having an implementation of the grammar and parser we explore a
-different approach, based closer on Piston's.
-
-TODO: More here.
+TODO: More comparison.

### 2020-03-04 doc/icfp20/music.tex
major work on harmony section

@@
 
-A \emph{chord} refers to a set of simultaneously sounding notes.
-In this paper we are particularly interested in \emph{triads},
-which are chords consisting of three notes separated by the 3rd
-intervals (i.e., \texttt{maj3} and \texttt{min3}).
-We define triads as a datatype \texttt{Triad} with roman numeral
-constructors representing the lowest note (called the \emph{root}).
-For instance, \texttt{I} in C major represents the triad C-E-G.
+A \emph{chord} refers to a set of simultaneously sounding notes.  In
+this paper we are particularly interested in \emph{triads}, which are
+chords consisting of three notes separated by the 3rd intervals (i.e.,
+\texttt{maj3} and \texttt{min3}).  We define triads as a datatype
+\texttt{Triad} with roman numeral constructors representing the scale
+degree of the lowest note (called the \emph{root}). The other two
+notes are then members of the scale a third and a fifth above the
+root. For instance, \texttt{I} in C major represents the triad C-E-G.

### 2020-03-04 doc/icfp20/harmony.tex
harmony half done

@@
 
-TODO: Finish this.
+As Piston does we require the harmonizing triad to include the melody
+note. A function \texttt{harmonizations} is used to generate a list of
+all candidate harmonizations for a given melody. For space
+considerations the function is not shown here but is available in the
+supplement \texttt{Harmony.agda}. Here again we encounter difficulties
+with equality and membership. We use a bit vector to represent the
+set of scale degrees in a triad where each scale degree is mapped to
+an element of \texttt{Fin 7}.
+
+We next restrict to progressions ending in a half cadence (V) to match
+Piston's example. This results in 25 candidate progressions.
 
@@
 Their technique is to first require the melody notes to be members of
-the harmonizing chords, then use their previous work to try to
-create a harmonic analysis of every possible sequence of chords that
-arises, choosing the one that most closely fits their
-grammar. Although the result is not bad, it clearly leaves room for
-improvement. Noted in the paper is that no attention is given to
-voicing (how each of the four parts flows horizontally and how the
-lines interact---in other words the counterpoint). Not noted is that
-the melody note is always doubled, whereas typically in four-part
-harmony it is preferred to double the root note. Also minimizing the
-number of parse errors as a means to choose the best harmonization
-does not necessarily have a musical meaning.
+the harmonizing chords, then use their previous work to try to create
+a harmonic analysis of every possible sequence of chords that arises,
+choosing the one that most closely fits their grammar. Although the
+result is not bad, it clearly leaves room for improvement. Noted in
+the paper is that no attention is given to voicing (how each of the
+four parts flows horizontally and how the lines interact---in other
+words the counterpoint). They use inversions (in which the
+third or fifth of the chord is in the bass) to ensure each melody line
+does not move too much, but as the harmonizing triads are all in close
+postion separated from the melody the result sounds like a soprano
+melody with block chord accompaniment rather than four independent
+lines.
+
+In addition note that the melody note is always doubled, whereas
+typically in four-part harmony it is preferred to double the root note
+as Piston does. Also minimizing the number of parse errors as a means
+to choose the best harmonization does not necessarily have a musical
+meaning.
 

### 2020-03-04 doc/icfp20/harmony.tex
tweak harmony section

--- 2020-03-04 doc/icfp20/harmony.tex
@@
 
-The primary Example~9.1 in Piston's text is shown on the left
-side of Figure~\ref{fig:harmonization} which shows the melody with a
+The primary Example~9.1 in Piston's text
+(Figure~\ref{fig:harmonization} left) shows a
 sample four part SATB harmonization in C~major. The melody is the
-series of highest notes, the soprano (S) line, while the other four
-parts alto (A), tenor (T) and bass (B) are the melodies below
+series of highest notes, i.e., the soprano (S) line, while the other three
+parts alto (A), tenor (T), and bass (B) are the melodies below
 it. Roman numerals denote the names of the triads at each vertical
@@
 each voice (called voice leading) and independence of each music
-line (in other words good counterpoint). In this chapter as well as
-throughout the text Piston gives advice on how to achieve these goals.
+line (in other words good counterpoint). In this chapter
+Piston gives advice on how to achieve these goals.
 
@@
 sequence of triads, in other words the harmonic progression. Earlier
-in the text on page~23 Piston presents a Table of Usual Root
-Prrogressions, noting which triads usually follow a given one and
-which are sometimes used or rare. It is easy to encode this in Agda.
+in the text (on page~23) Piston presents a Table of Usual Root
+Progressions, noting which triads usually follow a given one and
+which are sometimes used or rare. It is easy to encode this in Agda
+using the \texttt{Triad} datatype from Section~\ref{sec:music}.
 
 \begin{alltt}
-data Triad : Set where
-  I : Triad; II : Triad; III : Triad; IV : Triad; V : Triad; VI : Triad; VII : Triad
-
 record NextTriad : Set where
@@
 It is easiest to create a harmonic progression in reverse, to
-ensure one ends a musical phrase with either a full (V--I) or half (V)
+ensure that one ends a musical phrase with either a full (V--I) or half (V)
 cadence, and work backward from there. So we would like to invert this
@@
 
-In Haskell this would be easy--there is an \texttt{Eq} typeclass and a
+In Haskell this would be easy---there is an \texttt{Eq} typeclass and a
 method for automatically generating equalities. In the Agda Standard
-Library~\citep{agda-stdlib} is a relation with constructive proof of
+Library~\citep{agda-stdlib} is a relation with a constructive proof of
 membership, and requires far more effort than we need. Alternatives
@@
 
-TODO: Finish this.
+As Piston does we require the harmonizing triad to include the melody
+note. We use a function \texttt{harmonizations} to generate a list of
+all candidate harmonizations for a given melody. For space
+considerations we refer the reader to the supplement
+\texttt{Harmony.agda} for its definition. Here again we encounter difficulties
+with equality and membership. We use a bit vector to represent the
+set of scale degrees in a triad, where each scale degree is mapped to
+an element of \texttt{Fin 7}.
+
+We next restrict to progressions ending in a half cadence (V) to match
+Piston's example. This results in 25 candidate progressions.
 
@@
 
-Most of the the Haskell-based
+Most of the Haskell-based
 work~\citep{magalhaes-harmtrace,koops-fharm,magalhaes-fcomp} has
 focused on harmony. In particular \fharm~\citep{koops-fharm} tackles
-the problem of harmonization of a melody, and their harmonization of
-Piston's example is shown on the right side of
-Figure~\ref{fig:harmonization}. Their method is quite different from
+the problem of harmonization of a melody;
+Figure~\ref{fig:harmonization} right shows their harmonization of
+Piston's example. Their method is quite different from
 ours. Instead of using Piston's table of root progressions they use a
@@
 work~\citep{magalhaes-harmtrace}.
+More specifically, they first require the melody notes to be members of
+the harmonizing chords, then use their previous work to try to create
+a harmonic analysis of every possible sequence of chords that arises,
+choosing the one that most closely fits their grammar.
 
-Their technique is to first require the melody notes to be members of
-the harmonizing chords, then use their previous work to try to
-create a harmonic analysis of every possible sequence of chords that
-arises, choosing the one that most closely fits their
-grammar. Although the result is not bad, it clearly leaves room for
-improvement. Noted in the paper is that no attention is given to
-voicing (how each of the four parts flows horizontally and how the
-lines interact---in other words the counterpoint). Not noted is that
-the melody note is always doubled, whereas typically in four-part
-harmony it is preferred to double the root note. Also minimizing the
-number of parse errors as a means to choose the best harmonization
-does not necessarily have a musical meaning.
+Although the result produced by \fharm is not bad, it clearly leaves
+room for improvement. Noted in
+the paper is that no attention is given to voicing (how each of the
+four parts flows horizontally and how the lines interact---in other
+words the counterpoint). They use inversions (in which the
+third or fifth of the chord is in the bass) to ensure each melody line
+does not move too much, but as the harmonizing triads are all in close
+postion separated from the melody, the result sounds like a soprano
+melody with block chord accompaniment, rather than four independent
+lines.
+
+Furthermore, \fharm always double the melody note, not the root note
+as Piston does (and as preferred in four-part harmony in general).
+Also, minimizing the number of parse errors as a means
+to choose the best harmonization does not necessarily have a musical
+meaning.
 

### 2020-03-04 doc/icfp20/harmony.tex
fix typo

@@
 record NextTriad : Set where
-  constructor nextTriadb
+  constructor nextTriad
   field
@@
 table to give a list of likely triads to precede a given triad. This
-would normally be straightfoward but here we see a disadvantage of
+would normally be straightforward but here we see a disadvantage of
 using Agda. As is typical with languages based on dependent type
-theories, equality is a very subtle concept and even determing if a
+theories, equality is a very subtle concept and even determining if a
 term is a member of a list is nontrivial.
@@
 does not move too much, but as the harmonizing triads are all in close
-postion separated from the melody, the result sounds like a soprano
+position separated from the melody, the result sounds like a soprano
 melody with block chord accompaniment, rather than four independent

### 2020-03-04 doc/icfp20/harmony.tex
save work

--- 2020-03-04 doc/icfp20/harmony.tex
@@
 method for automatically generating equalities. In the Agda Standard
-Library~\citep{agda-stdlib} is a relation with constructive proof of
-membership, and requires far more effort than we need. Alternatives
-include defining decidable equality for triads (which requires writing
-out 49 cases) or mapping to natural numbers and using already
-implemented decidable or boolean equality for those. For now we simply
-write the inverse function manually, excluding the rare cases.
+Library~\citep{agda-stdlib} list membership is a relation with
+constructive evidence of membership, and requires far more effort we
+would like given that we currently don't need to do
+proofs. Alternatives include defining decidable equality for triads
+(which requires writing out 49 cases) or mapping to natural numbers
+and using already implemented decidable or boolean equality for
+those. For now we simply write the inverse function manually,
+excluding the rare cases.
 

### 2020-03-04 doc/icfp20/harmony.tex

--- 2020-03-04 doc/icfp20/harmony.tex
@@
 method for automatically generating equalities. In the Agda Standard
-Library~\citep{agda-stdlib} is a relation with a constructive proof of
-membership, and requires far more effort than we need. Alternatives
-include defining decidable equality for triads (which requires writing
-out 49 cases) or mapping to natural numbers and using already
-implemented decidable or boolean equality for those. For now we simply
-write the inverse function manually, excluding the rare cases.
+Library~\citep{agda-stdlib} list membership is a relation with
+constructive evidence of membership, and requires far more effort we
+would like given that we currently don't need to do
+proofs. Alternatives include defining decidable equality for triads
+(which requires writing out 49 cases) or mapping to natural numbers
+and using already implemented decidable or boolean equality for
+those. For now we simply write the inverse function manually,
+excluding the rare cases.
 

### 2020-03-04 doc/icfp20/counterpoint.tex
tweak comparison to Haskell '17 work

@@
 
-The type-theoretical formalization of counterpoint has first
-been attempted by \citet{szamozvancev-welltyped}.
-They encode the rules for intervals and motion by declaring
-type classes \texttt{ValidHarmInterval} and \texttt{ValidMotion},
-and specifying which instances are invalid using GHC's support
-for custom type errors.
+The type-theoretical formalization of counterpoint has first been
+attempted by \citet{szamozvancev-welltyped}.
+They encode the rule for intervals by declaring a type class
+\texttt{ValidHarmInterval} that classifies valid intervals, and then
+specifying which instances are invalid using GHC's support for
+custom type errors.
+They also implement the restriction on motion based on a similar idea.
 
-The availability of the custom error feature is an advantage of using
-Haskell, but the language's type system does not allow producing
-detailed error messages.
+The availability of the custom type error feature is an advantage of
+using Haskell: it allows the rule writer to leverage Haskell's type checker.
+We believe that extending Agda with such support would be
+beneficial in music formalization and for many other purposes.
+
+More recently, \citet{cong-cp} give an implementation of first
+species counterpoint using dependent types in Agda.
+Their idea is to represent correct counterpoint as a list-like
+datatype, where the base cases enforce a valid ending and the
+inductive case guarantees correct uses of intervals and motion
+(by means of implicit arguments representing proofs).
+
+Our representation of counterpoint improves on Cong and Leo's
+in two ways.
+First, while Cong and Leo rely on the Agda type checker to report
+errors, we use our own type checker (implemented as the
+\texttt{checkXXX} functions) and type errors (defined as the
+\texttt{XXXError} datatypes).
+This allows us to produce more user-friendly error messages.
 Suppose we have replaced the second interval of Figure~\ref{fig:fs} by
@@
 
-\noindent Observe that the message clearly tells us which intervals
-form what kind of invalid motion.
-
-In Szamozvancev and Gale's implementation, the same example
-results in the following error:
-
-\begin{alltt}
-Direct motion into a perfect octave is forbidden.
-\end{alltt}
-
-\noindent The lack of the precise information about the intervals
-and motion is due to the absence of native dependent types in
-Haskell.
-
-More recently, \citet{cong-cp} give an implementation of first
-species counterpoint using dependent types in Agda.
-Their idea is to represent correct counterpoint as a list-like
-datatype, where the base cases enforce a valid ending and the
-inductive case guarantees correct uses of intervals and motion
-(by means of implicit arguments representing proofs).
-
-Our representation of counterpoint improves on their
-implementation in two ways.
-First, while \citet{cong-cp} rely on the Agda type checker to report
-errors in the counterpoint, we use our own type checker (implemented
-as the \texttt{checkXXX} functions) and type errors (defined as the
-\texttt{XXXError} datatypes).
-This allows us to produce helpful error messages while keeping
-types simple.
-Indeed, for the parallel octave example discussed above,
-\citet{cong-cp} would just report the existence of an unsolved
-metavariable \texttt{\_13}, meaning that Agda failed to construct
-a proof required by the counterpoint constructor:
+\noindent In contrast, Cong and Leo would just report the existence
+of an unsolved metavariable \texttt{\_13}, meaning that Agda failed
+to construct a proof required by the counterpoint constructor.
+Thus, the user would see the following message in the Agda buffer:
 
@@
 incorporating all the rules into a counterpoint datatype, we define
-each of them as a separate function and combine them using a
-record type.
+each rule as a separate function and combining them using a record type.
 This makes it easy to switch between strict and relaxed rule sets:

### 2020-03-04 doc/icfp20/harmony.tex
add harmonization figure
harmonization section complete except for diagram

@@
 \begin{alltt}
-previousTriads : Triad → List Triad
+previousTriads : Triad \(\rightarrow\) List Triad
 previousTriads I   = V :: IV :: VII :: []
@@
 set of scale degrees in a triad, where each scale degree is mapped to
-an element of \texttt{Fin 7}.
+an element of \texttt{Fin 7}. However we are able to make use of
+dependent types in the form of vectors to ensure the length of each
+harmonization matches the melody, and continue to use vectors in the
+following functions where appropriate.
 
 We next restrict to progressions ending in a half cadence (V) to match
-Piston's example. This results in 25 candidate progressions.
+Piston's example (\texttt{filter halfCadence}). This results in 25
+candidate progressions. For each pitch and triad pair we construct a
+\texttt{harmonizingChord} in root position with the root doubled. The
+function \texttt{voiceChord} spaces the notes nicely. All of this work
+ins done in the function \texttt{voicedHarmonizations}.
 
-See Figure~\ref{fig:harmonizationMT}, which is currently just a copy of
-the other figure, but included here now for space.
+\begin{alltt}
+voicedHarmonizations : \{n : N\} \(\rightarrow\) Vec Pitch n \(\rightarrow\) List (Vec (Vec Pitch 4) n)
+voicedHarmonizations \{n\} ps =
+  let ds = map pitchToDegreeCMajor ps
+      hs : List (Vec Triad n)
+      hs = filter halfCadence (harmonizations ds)
+      in map (\(\lambda\) ts \(\rightarrow\) map (\(\lambda\) pt \(\rightarrow\)
+             \(\text{proj}_1\) pt :: harmonizingChord (\(\text{proj}_1\) pt) (\(\text{proj}_2\) pt)) (zip ps ts)) hs
+\end{alltt}
+
+To determine the ``best'' of these harmonizations, we now make use of
+\texttt{motionCheck} from the counterpoint work. We compare all six
+pairs of melodies, count the number of motion errors present, and pick
+the harmonization which minimizes these flaws. The functon to
+determine all errors is straightfoward.
+
+\begin{alltt}
+motionErrors : \{n : N\} \(\rightarrow\) Vec (Vec Pitch 4) n \(\rightarrow\) List MotionError
+motionErrors xs =
+  let ss = map (head) xs
+      as = map (head \(\circ\) tail) xs
+      ts = map (head \(\circ\) tail \(\circ\) tail) xs
+      bs = map (head \(\circ\) tail \(\circ\) tail \(\circ\) tail) xs
+
+      sas = map pitchPairToPitchInterval (toList (zip as ss))
+      sts = map pitchPairToPitchInterval (toList (zip ts ss))
+      sbs = map pitchPairToPitchInterval (toList (zip bs ss))
+      ats = map pitchPairToPitchInterval (toList (zip ts as))
+      abs = map pitchPairToPitchInterval (toList (zip bs as))
+      tbs = map pitchPairToPitchInterval (toList (zip bs ts))
+  in concatMap checkMotion (sas :: sts :: sbs :: ats :: abs :: tbs :: [])
+\end{alltt}
+
+The harmonization produced is shown in Figure~\ref{fig:harmonizationMT}.
 
@@
 
-TODO: More comparison.
+Our harmonization follows Piston's rules more closely and has arguably
+better spacing of voices; readers can decide for themselves which is
+more pleasing.

### 2020-03-04 doc/icfp20/music.tex

@@
 
-[TODO] Maybe need to explain degrees, instead of duration and notes.
-
 \paragraph{Pitches}
 
-A \emph{pitch} tells us how high or low a sound is.
+The most basic concept in melody is \emph{pitches}, which tell us
+how high or low a tone is.
 We represent pitches as natural numbers, but throughout the paper,
@@
 
+\paragraph{Scale Degrees}
+
+While pitches uniquely determine the position of a tone on e.g.
+a piano keyboard, \emph{scale degrees} stand for the relative position
+of a tone in a scale.
+For instance, a diatonic scale consists of seven scale degrees, which
+corresponds to C, D, E, F, G, A, B in the case of C major.
+We represent scale degrees as a datatype with seven constructors
+\texttt{d1} - \texttt{d7}.
+
 \paragraph{Duration}
 
-\emph{Duration} denotes an unspecified unit of time during which
-a sound or silence lasts.
+For rhythm, the fundamental concept is \emph{duration}, an
+unspecified unit of time during which a sound or silence lasts.
 We represent duration as a natural number, and use this value to
@@
 
-\paragraph{Notes}
+% \paragraph{Notes}
 
-Combining pitches and duration gives us \emph{notes}.
-We represent notes as a datatype \texttt{Note} with two constructors:
-\texttt{tone} for notes with sound and \texttt{rest} for those without.
-For example, \texttt{tone whole (c 5)} is a whole note whose sound
-is the middle C.
+% Combining pitches and duration gives us \emph{notes}.
+% We represent notes as a datatype \texttt{Note} with two constructors:
+% \texttt{tone} for notes with sound and \texttt{rest} for those without.
+% For example, \texttt{tone whole (c 5)} is a whole note whose sound
+% is the middle C.
 
@@
 We represent intervals as natural numbers, but in this paper, we
-refer to them using more informative names such as \texttt{maj3}
-and \texttt{per8}.
+refer to them using names such as \texttt{maj3} and \texttt{per8},
+which are more reader-friendly.
 
@@
 
-A \emph{chord} refers to a set of simultaneously sounding notes.  In
-this paper we are particularly interested in \emph{triads}, which are
+A \emph{chord} refers to a set of simultaneously sounding notes.
+In this paper we are particularly interested in \emph{triads}, which are
 chords consisting of three notes separated by the 3rd intervals (i.e.,
-\texttt{maj3} and \texttt{min3}).  We define triads as a datatype
-\texttt{Triad} with roman numeral constructors representing the scale
-degree of the lowest note (called the \emph{root}). The other two
-notes are then members of the scale a third and a fifth above the
-root. For instance, \texttt{I} in C major represents the triad C-E-G.
+\texttt{maj3} and \texttt{min3}).
+We define triads as a datatype with roman numeral constructors
+representing the scale degree of the lowest note (called the \emph{root}).
+The other two notes are then members of the scale a third and a fifth
+above the root.
+For instance, the triad \texttt{I} in C major corresponds to the chord
+C-E-G.
+Triads can also be arranged in a way that a non-root note appears as
+the lowest note.
+Such arrangement is called \emph{inversion}; if a triad is not applied
+inversion, it is in \emph{root position}.

### 2020-03-04 doc/icfp20/harmony.tex
restore some code

@@
      \hspace{2.35cm} V \ \ \ \ \ \ \ \,I \ \ \ \ \ \ \ \,IV \,VI \ \,III \ IV \ \ \,V
-     \hspace{2.6cm} V \ \ \ \ \ \ \ \,I \ \ \ \ \ \,IV VI \,III\,IV \ \ V
+     \hspace{2.6cm} V \,III \ \ I \ III \ \,II IV \,III\,IV \ \,V
     \end{small}
@@
   \begin{figure}
-    \includegraphics[width=5.5cm]{fig/piston.png}
-    \hspace{1cm}
-    \includegraphics[width=5cm]{fig/fharm.png} \\
+    \includegraphics[width=5.5cm]{fig/harmmt.png}
     \begin{flushleft}
     \begin{small}
-     \hspace{2.35cm} V \ \ \ \ \ \ \ \,I \ \ \ \ \ \ \ \,IV \,VI \ \,III \ IV \ \ \,V
-     \hspace{2.6cm} V \ \ \ \ \ \ \ \,I \ \ \ \ \ \,IV VI \,III\,IV \ \ V
+     \hspace{5.4cm} I \ \ V \ \ \ VI \ \,V \ \ IV \ I \ \ \ V  \,IV \ \ \ \ V
     \end{small}
     \end{flushleft}
-    \caption{Harmonization by Music Tools (to be replaced by real figure)}
+    \caption{Harmonization by Music Tools}
     \Description{Harmonization by Music Tools}
@@
 rootProgression : Triad \(\rightarrow\) NextTriad
-rootProgression I   = nextTriad (IV :: V :: []) (VI :: []) (II :: III :: [])
+rootProgression I   = nextTriad (IV :: V :: []) (VI :: []) (II  :: III : [])
 rootProgression II  = nextTriad (V :: []) (IV :: VI :: []) (I :: III :: [])
--- other cases omitted for space
+rootProgression III = nextTriad (VI :: []) (IV :: []) (I :: II :: V :: [])
+rootProgression IV  = nextTriad (V :: []) (I :: II :: []) (III :: VI :: [])
+rootProgression V   = nextTriad (I :: []) (IV :: VI :: []) (II :: III :: [])
+rootProgression VI  = nextTriad (II :: V :: []) (III :: IV :: []) (I :: [])
+rootProgression VII = nextTriad (I  :: III :: []) (VI :: []) (II :: IV  :: V :: [])
 \end{alltt}
@@
 previousTriads II  = VI :: IV :: []
--- other cases omitted for space
+previousTriads III = VI :: VII :: []
+previousTriads IV  = I :: V :: II :: III :: []
+previousTriads V   = I :: IV :: II :: VI :: []
+previousTriads VI  = IV :: I :: II :: V :: VII :: []
+previousTriads VII = []
 \end{alltt}

### 2020-03-04 doc/icfp20/harmony.tex
minor tweaks to harmony

@@
 it. Roman numerals denote the names of the triads at each vertical
-slice. Note that even though there are four voices there are only
+slice. Note that even though the example consists of four voices there are only
 three distinct pitches sounding at each vertical slice (where tones an
@@
 typical the root of the triad is always doubled. Also note that all
-triads are in root position (meaning the root is in the bass voice),
+triads are in root position (i.e., the root is in the bass voice),
 as Piston restricts to this case initially in the chapter (he does
@@
 
-In Haskell this would be easy---there is an \texttt{Eq} typeclass and a
+In Haskell this would be easy---there is an \texttt{Eq} type class and a
 method for automatically generating equalities. In the Agda Standard
@@
 function \texttt{voiceChord} spaces the notes nicely. All of this work
-ins done in the function \texttt{voicedHarmonizations}.
+is done in the function \texttt{voicedHarmonizations}.
 
@@
 pairs of melodies, count the number of motion errors present, and pick
-the harmonization which minimizes these flaws. The functon to
-determine all errors is straightfoward.
+the harmonization which minimizes these flaws. The function to
+determine all errors is straightforward.
 

### 2020-03-04 doc/icfp20/harmony.tex
fix typos

--- 2020-03-04 doc/icfp20/harmony.tex
@@
 Library~\citep{agda-stdlib} list membership is a relation with
-constructive evidence of membership, and requires far more effort we
+constructive evidence of membership, and requires far more effort than we
 would like given that we currently don't need to do
@@
 We next restrict to progressions ending in a half cadence (V) to match
-Piston's example (\texttt{filter halfCadence}). This results in 25
+Piston's example. This results in 25
 candidate progressions. For each pitch and triad pair we construct a

### 2020-03-04 doc/icfp20/conclusion.tex
final tweaks & swap figures

@@
 
-We believe that functional programming, and in particular functional
-programming with dependent types, is a promising way to do software
-engineering going forward. The languages and tools are perhaps not yet
-as finely polished and rich as those typically used in production
-software, but they are fundamentally far more powerful, and thus once
-the infrastructure (and training of programmers) catches up should
-become the languages of choice.
+% We believe that functional programming, and in particular functional
+% programming with dependent types, is a promising way to do software
+% engineering going forward. The languages and tools are perhaps not yet
+% as finely polished and rich as those typically used in production
+% software, but they are fundamentally far more powerful, and thus once
+% the infrastructure (and training of programmers) catches up should
+% become the languages of choice.
 
-Music is an especially valuable domain in which to explore these
-ideas. We have found it to be a microcosm of issues that arise
+In this paper we have reported our experience in using Agda for
+formalizing music theory.
+We have found music to be a microcosm of issues that arise
 everywhere in software engineering: modularity, composability,
-correctness, and data issues such as handling equivalent formulations of
-the same data as well slightly different formulations. If we can
+correctness, and data issues such as handling equivalent or slightly
+different formulations of the same data. If we can
 demonstrate that functional programming with dependent types works
@@
 For now we have explicitly converted between equivalent representations
-and explicitly lifted functions, to get a feel for the amount of work
+and lifted functions, to get a feel for the amount of work
 required.  However in the world of dependent types there are known

### 2020-03-04 doc/icfp20/counterpoint.tex
add scale degree & inversion to glossary
final tweaks & swap figures

@@
 counterpoint follows certain rules.
-For space reason, we only describe two variants of species counterpoint;
-other variants can be formalized in an analogous way.
+Due to the space limitation, we only show some essential parts of
+our implementation; see our anonymized supplement
+\texttt{Counterpoint.agda} for full details.
 
@@
 \begin{alltt}  
-first : PitchInterval -- Pitch \(\times\) Interval
+first : PitchInterval -- defined as Pitch \(\times\) Interval
 first = (c 5 , per8)
@@
 As we saw in Figure~\ref{fig:interval}, consonant intervals include
-the unison, 3rd, 5th, 6th, and 8th, and among these, the unison is
-clearly an obstacle to distinguishing between the two lines of music.
+the unison, 3rd, 5th, 6th, and 8th.
+Among these, the unison is clearly an obstacle to distinguishing
+between the two lines of music.
 Therefore, the middle bars must consist only of the latter four intervals.
@@
 To rule out such motion, we define the \texttt{checkMotion}
-function, which inspects all pairs of two adjacent intervals (using an
-auxiliary function \texttt{pairs}) and returns a list of erroneous patterns.
+function, which inspects all pairs of two adjacent intervals (using 
+auxiliary functions \texttt{motion} and \texttt{pairs}) and returns a
+list of erroneous patterns.
 
@@
 The ending of the music should express relaxation.
-There is a consensus that the unison and 8th are the most stable
-intervals, hence the music must end with either of these intervals.
+Since the unison and 8th are considered most stable, the music
+must end with either of these intervals.
 The last interval should also be approached by a \emph{cadence},
@@
 With this record type, we can show that the counterpoint in
-Figure~\ref{fig:fs} is correct, since the music is an inhabitant of
-\texttt{FirstSpecies}.
+Figure~\ref{fig:fs} is correct, by constructing the following proof term:
 
@@
 In second species, the first and last bars may have a different structure
-from middle bars.
+from the middle bars.
 More specifically, it is encouraged to begin the counterpoint line with
@@
 type for the first and last bars, and define a new type
-\texttt{PitchInterval2} for middle bars.
+\texttt{PitchInterval2} for the middle bars.
 
@@
 
-middle2 : List PitchInterval2 -- Pitch \(\times\) Interval \(\times\) Interval
+middle2 : List PitchInterval2 -- defined as Pitch \(\times\) Interval \(\times\) Interval
 middle2 =
@@
 
-Strong beats in middle bars are constrained by the same rules as in
-first species: they must all be consonant, non-unison intervals.
+Strong beats in the middle bars are constrained by the same rules
+as in first species: they must all be consonant, non-unison intervals.
 % We encode this restriction as the \texttt{checkStrongBeats}
@@
 Parallel and similar motion towards a perfect interval is prohibited
-across bars.
+across bars, that is, between the weak beat in one bar and the
+strong beat in the next bar.
 % We enforce this rule using the \texttt{checkMotion2} function,
@@
 custom type errors.
-They also implement the restriction on motion based on a similar idea.
+They also rule out invalid motion using a similar technique.
 
-The availability of the custom type error feature is an advantage of
-using Haskell: it allows the rule writer to leverage Haskell's type checker.
-We believe that extending Agda with such support would be
-beneficial in music formalization and for many other purposes.
+The idea of producing custom musical type errors underlies our
+implementation as well.
+The difference is that we have no means to leverage the Agda
+type checker due to the lack of such a language feature, which
+we consider to be a downside of using Agda for musical purposes.
 
@@
 in two ways.
-First, while Cong and Leo rely on the Agda type checker to report
-errors, we use our own type checker (implemented as the
+First, while they rely on the Agda type checker to report errors,
+we use our own type checker (implemented as the
 \texttt{checkXXX} functions) and type errors (defined as the
@@
 This allows us to produce more user-friendly error messages.
-Suppose we have replaced the second interval of Figure~\ref{fig:fs} by
-an octave (8th) of D, breaking the rule that perfect intervals cannot
-be approached by parallel motion.
+To see the difference, suppose we have replaced the second interval
+of Figure~\ref{fig:fs} by an octave (8th) of D, breaking the rule
+that perfect intervals cannot be approached by parallel motion.
 In our implementation, this causes an error in the \texttt{motionOk}
@@
 \noindent In contrast, Cong and Leo would just report the existence
-of an unsolved metavariable \texttt{\_13}, meaning that Agda failed
-to construct a proof required by the counterpoint constructor.
+of an unsolved metavariable, meaning that Agda failed to construct
+a proof required by the counterpoint constructor.
 Thus, the user would see the following message in the Agda buffer:
@@
 The second improvement from \citet{cong-cp} is that, instead of
-incorporating all the rules into a counterpoint datatype, we define
-each rule as a separate function and combining them using a record type.
+hard-coding all the rules into a datatype designed specifically for
+counterpoint, we define each rule as a separate function and
+combine them using a record type.
 This makes it easy to switch between strict and relaxed rule sets:
@@
 It also allows us to reuse some of the rules in a context other than
-counterpoint, as we will see in the next section.
+counterpoint, as we will see in Section~\ref{sec:harmony:complete}.

### 2020-03-04 doc/icfp20/harmony.tex
final tweaks & swap figures

@@
     \hspace{1cm}
-    \includegraphics[width=5cm]{fig/fharm.png} \\
+    \includegraphics[width=5.5cm]{fig/harmmt.png} \\
     \begin{flushleft}
     \begin{small}
-     \hspace{2.35cm} V \ \ \ \ \ \ \ \,I \ \ \ \ \ \ \ \,IV \,VI \ \,III \ IV \ \ \,V
-     \hspace{2.6cm} V \,III \ \ I \ III \ \,II IV \,III\,IV \ \,V
+     \hspace{2.1cm} V \ \ \ \ \ \ \ \,I \ \ \ \ \ \ \ \,IV \,VI \ \,III \ IV \ \ \,V
+     \hspace{2.65cm} I \ \ V \ \ \ VI \ V \ \ IV \ I \ \ \ V  \,IV \ \ \ \ V
     \end{small}
     \end{flushleft}
-    \caption{Harmonization by Piston (left) and \fharm (right)}
-    \Description{Harmonization by Piston (left) and \fharm (right)}
+    \caption{Harmonization by Piston (left) and Our Implementation (right)}
+    \Description{Harmonization by Piston (left) and Our Implementation (right)}
     \label{fig:harmonization}
@@
 
-\newcommand{\HarmonizationMT}{
+\newcommand{\HarmonizationF}{
   \begin{figure}
-    \includegraphics[width=5.5cm]{fig/harmmt.png}
+    \includegraphics[width=5.5cm]{fig/fharm.png}
     \begin{flushleft}
     \begin{small}
-     \hspace{5.4cm} I \ \ V \ \ \ VI \ \,V \ \ IV \ I \ \ \ V  \,IV \ \ \ \ V
+     \hspace{5.5cm} V \,III \ \ \ I \ \,III \ \,II \,IV \ \,III IV \ \ V
     \end{small}
     \end{flushleft}
-    \caption{Harmonization by Music Tools}
-    \Description{Harmonization by Music Tools}
-    \label{fig:harmonizationMT}
+    \caption{Harmonization by \fharm}
+    \Description{Harmonization by \fharm}
+    \label{fig:harmonizationF}
   \end{figure}
@@
 is primarily concerned with the progression of chords throughout a
-piece of music. It is in essence the of dual of counterpoint. In this
+piece of music. It is in essence the dual of counterpoint. In this
 section we formalize a small part of the classic text
@@
 concerned with techniques for harmonizing a given melody.
+The complete implementation can be found in \texttt{Harmony.agda}
+and \texttt{Piston.agda} of the anonymized supplement.
 
@@
 constructive evidence of membership, and requires far more effort than we
-would like given that we currently don't need to do
+would like, given that we currently do not need to do
 proofs. Alternatives include defining decidable equality for triads
-(which requires writing out 49 cases) or mapping to natural numbers
+(which requires writing out 49 cases), or mapping to natural numbers
 and using already implemented decidable or boolean equality for
@@
 note. We use a function \texttt{harmonizations} to generate a list of
-all candidate harmonizations for a given melody. For space
-considerations we refer the reader to the supplement
-\texttt{Harmony.agda} for its definition. Here again we encounter difficulties
+all candidate harmonizations for a given melody (see
+\texttt{Harmony.agda} for its definition). Here again we encounter difficulties
 with equality and membership. We use a bit vector to represent the
@@
 
-The harmonization produced is shown in Figure~\ref{fig:harmonizationMT}.
-
-\HarmonizationMT
+The harmonization produced is shown in Figure~\ref{fig:harmonization}
+right.
 
@@
 
-Most of the Haskell-based
-work~\citep{magalhaes-harmtrace,koops-fharm,magalhaes-fcomp} has
+\HarmonizationF
+
+Most of the Haskell-based work on
+music~\citep{magalhaes-harmtrace,koops-fharm,magalhaes-fcomp} has
 focused on harmony. In particular \fharm~\citep{koops-fharm} tackles
-the problem of harmonization of a melody;
-Figure~\ref{fig:harmonization} right shows their harmonization of
+the problem of harmonizing a melody;
+Figure~\ref{fig:harmonizationF} shows their harmonization of
 Piston's example. Their method is quite different from
@@
 four parts flows horizontally and how the lines interact---in other
-words the counterpoint). They use inversions (in which the
-third or fifth of the chord is in the bass) to ensure each melody line
+words the counterpoint). They use inversions to ensure each melody line
 does not move too much, but as the harmonizing triads are all in close
@@
 
-Furthermore, \fharm always double the melody note, not the root note
-as Piston does (and as preferred in four-part harmony in general).
-Also, minimizing the number of parse errors as a means
+Among those issues not discussed in the paper, one is the choice of
+notes to be doubled: \fharm always doubles the melody note, while
+in four part harmony it is generally preferred to double the root
+note as Piston does.
+Another concern is that minimizing the number of parse errors as a means
 to choose the best harmonization does not necessarily have a musical

### 2020-03-04 doc/icfp20/music.tex
final tweaks & swap figures

@@
 unspecified unit of time during which a sound or silence lasts.
-We represent duration as a natural number, and use this value to
-calculate the absolute length when the music is played at a specific
-tempo.
-As an example, the duration constant \texttt{whole} has value
-\texttt{16} and corresponds to 4 seconds if the tempo is 60 beats
-per minute.
+Examples of duration include whole, half, and quarter;
+we define them simply as natural number constants.
 
@@
 or imperfect.
-We represent intervals as natural numbers, but in this paper, we
-refer to them using names such as \texttt{maj3} and \texttt{per8},
-which are more reader-friendly.
+We represent intervals as natural numbers, and refer to them using
+conventional names such as \texttt{maj3} and \texttt{per8}.
 
@@
 A \emph{chord} refers to a set of simultaneously sounding notes.
-In this paper we are particularly interested in \emph{triads}, which are
+Of our particular interest is \emph{triads}, which are
 chords consisting of three notes separated by the 3rd intervals (i.e.,
@@
 Such arrangement is called \emph{inversion}; if a triad is not applied
-inversion, it is in \emph{root position}.
+inversion, we say that it is in \emph{root position}.

### 2020-03-04 doc/icfp20/intro.tex
final tweaks & swap figures
some more tweaks

@@
 currently has limited support for dependent types, and requires many
-extentions and tricks such as the use of singleton
+extensions and tricks such as the use of singleton
 types~\citep{eisenberg-singleton}.
@@
 final section we discuss possible ways to reduce the tedium.
+
+Our implementation is available as the anonymized supplement.

### 2020-03-04 doc/icfp20/music.tex
some more tweaks

@@
   \end{flushleft}
-  \caption{13 Kinds of Intervals}
-  \Description{13 Kinds of Intervals}
+  \caption{Intervals Within  an Octave}
+  \Description{Intervals Within an Octave}
   \label{fig:interval}
@@
 
-The most basic concept in melody is \emph{pitches}, which tell us
+One of the most basic concepts in music is \emph{pitches}, which tell us
 how high or low a tone is.
@@
 
-For rhythm, the fundamental concept is \emph{duration}, an
+\emph{Duration} is a fundamental concept for rhythm, denoting an
 unspecified unit of time during which a sound or silence lasts.

### 2020-04-21 doc/icfp20/response.txt
start response

We would like to thank the reviewers for their thorough reading and
valuable feedback.  The reviewers are right to point out our limited
use of dependent types; ...

Below is our answer to other questions.

### To Reviewer A

### To Reviewer B

* The best way to understand Line 315 is to look at the example in
  Figure 4.  In bar 5, the counterpoint line moves downwards from
  C to E (unison).  Then, in the next bar, it moves upwards from E to
  F.  We call a motion "step-wise" when its endpoints are adjacent
  notes, such as E-F.

### To Reviewer C

* The word "unspecified" in Line 87 is not a typo.  The unit of time is
  specified when the music is played at a particular tempo.

### 2020-04-22 doc/icfp20/response.txt
response draft

@@
 We would like to thank the reviewers for their thorough reading and
-valuable feedback.  The reviewers are right to point out our limited
-use of dependent types; ...
+valuable feedback. We agree with the reviewers that the use of
+dependent types in the paper is not sufficient to demonstrate their
+true value in the context of music theory. There is additional use of
+dependent types in the code that did not make it into the paper due to
+space limitations and the difficulty of fitting it into the main
+narrative, but likely this would not have been enough to change
+anyone's mind. We have much more planned that should demononstrate how
+compelling dependent types are in this context, but this is still
+future work.
+
+All reviewers raise other interesting points that we intend to
+address, and we thank them for pointing these out. Reviewers B and C
+raise doubts about whether the extrinsic approach in this paper is
+really superior to the intrinsic approach of our previous
+work. Although our experience so far has indicated the extrinsic
+approach seems both more flexible and more user-friendly, a more
+detailed comparison is warranted. We were unaware that Agda's
+reflection mechanism allows custom error messages, so thanks to
+Reviewer B for mentioning this and we will try it out. It is enticing
+to take advantage of the built-in type checker as much as possible,
+but it can also be constraining if one wants to for example explore
+multiple rule systems at once. The two approaches are not so different
+as indices can be represented as predicates attached to a sigma
+(record) type, and with an appropriate equivalence can be converted
+back and forth, which we also intend to explore. As Reviewer C notes
+we need to be more clear about what advantages there are to running
+the check at compile time rather than run-time in Haskell.
+
+Reviewers B and C also mention constraint-based search in which the type
+system may be used to partially constrain the search, and this is also
+planned future work.
+
+Reviewer A suggests implementing previoudTriads as a relation instead
+of a function, and the tradeoff between these approaches, and indeed
+many other aspects of practical programming using dependent types, are
+areas we hope to explore in detail going forward.
+
+[Note to Youyou: We could just leave out specific answers to questions
+unless you think the reviewers are really interested in this.]
 

### 2020-04-23 doc/icfp20/response.txt
fix typo; address one more minor question

@@
 narrative, but likely this would not have been enough to change
-anyone's mind. We have much more planned that should demononstrate how
+anyone's mind. We have many more plans that should demonstrate how
 compelling dependent types are in this context, but this is still
@@
 All reviewers raise other interesting points that we intend to
-address, and we thank them for pointing these out. Reviewers B and C
-raise doubts about whether the extrinsic approach in this paper is
-really superior to the intrinsic approach of our previous
+address, and we thank them for pointing these out.  Below we would
+like to respond to the main points.
+
+### Extrinsic vs. intrinsic approaches (Reviewers B and C)
+
+The reviewers raise doubts about whether the extrinsic approach
+in this paper is really superior to the intrinsic approach of previous
 work. Although our experience so far has indicated the extrinsic
@@
 multiple rule systems at once. The two approaches are not so different
-as indices can be represented as predicates attached to a sigma
+as indices can be represented as predicates attached to a $\Sigma$
 (record) type, and with an appropriate equivalence can be converted
@@
 
-Reviewers B and C also mention constraint-based search in which the type
-system may be used to partially constrain the search, and this is also
+### Constraint-based search (Reviewers B and C)
+
+The reviewers also mention constraint-based search in which the type
+system may be used to partially constrain the search.  This is also
 planned future work.
 
-Reviewer A suggests implementing previoudTriads as a relation instead
-of a function, and the tradeoff between these approaches, and indeed
+### Triads as a relation (Reviewer A)
+
+The reviewer suggests implementing previousTriads as a relation instead
+of a function. The tradeoff between these approaches, and indeed
 many other aspects of practical programming using dependent types, are
@@
 
-Below is our answer to other questions.
+[Note to John: I agree that these questions are not very important,
+but I would prefer not to ignore them.]
 
-### To Reviewer A
+Lastly, let us answer some minor questions.
 
@@
 
-* The best way to understand Line 315 is to look at the example in
-  Figure 4.  In bar 5, the counterpoint line moves downwards from
-  C to E (unison).  Then, in the next bar, it moves upwards from E to
-  F.  We call a motion "step-wise" when its endpoints are adjacent
-  notes, such as E-F.
+* The reviewer asks how to interpret the sentence in Line 315.  The
+  best way to understand it is to look at the example in Figure 4.
+  In bar 5, the counterpoint line moves downwards from C to E (unison).
+  Then, in the next bar, it moves upwards from E to F. We call a motion
+  "step-wise" when its endpoints are adjacent notes, such as E-F.
+
+* The reviewers wonders in what way harmony is dual to counterpoint.
+  As Cong and Leo [FARM '19] suggest, we may view harmony as a
+  horizontal view of a musical score, and counterpoint as a vertical view.
 

### 2020-04-23 doc/icfp20/response.txt
response submitted version

@@
 multiple rule systems at once. The two approaches are not so different
-as indices can be represented as predicates attached to a $\Sigma$
+as indices can be represented as predicates attached to a sigma
 (record) type, and with an appropriate equivalence can be converted
@@
 
-[Note to Youyou: We could just leave out specific answers to questions
-unless you think the reviewers are really interested in this.]
-
-[Note to John: I agree that these questions are not very important,
-but I would prefer not to ignore them.]
-
 Lastly, let us answer some minor questions.
@@
 
-* The reviewers wonders in what way harmony is dual to counterpoint.
+* The reviewer wonders in what way harmony is dual to counterpoint.
   As Cong and Leo [FARM '19] suggest, we may view harmony as a

### 2022-08-10 doc/farm22/abstract.tex
farm22 doc from haskell

\documentclass[sigplan,screen]{acmart}
\settopmatter{printccs=false, printacmref=false}

\setcopyright{none}
\acmPrice{}
\acmDOI{}
\acmYear{}
\copyrightyear{}
\acmISBN{}
\acmConference[FARM '22]{}
\acmBooktitle{}

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
\citestyle{acmauthoryear}   %% For author/year citations

%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption
\usepackage{alltt}

\begin{document}

%% Title information
\title{Demo: Counterpoint Analysis and Synthesis}
%\subtitle{Functional Pearl}

%% Author information
%% Contents and number of authors suppressed with 'anonymous'.

\author{John Leo}
\affiliation{
  \institution{Halfaya Research}
  \city{Bellevue}
  \state{WA}
  \country{USA}
}
\email{leo@halfaya.org}

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
We present a Haskell library to help analyze
and synthesize musical counterpoint. The tool allows expression of
generic constraints in a higher-level musical language which are
translated to a lower level for use both to find rule violations in
existing music and to generate (using an SMT solver) new
music satisfying the constraints. The tool is intended for use by
musicians who need only have a basic knowledge of Haskell.
\end{abstract}

%% 2012 ACM Computing Classification System (CSS) concepts
%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10010405.10010469.10010475</concept_id>
<concept_desc>Applied computing~Sound and music computing</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10011007.10011006.10011008.10011009.10011012</concept_id>
<concept_desc>Software and its engineering~Functional languages</concept_desc>
<concept_significance>500</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Applied computing~Sound and music computing}
\ccsdesc[500]{Software and its engineering~Functional languages}
%% End of generated code

%% Keywords
%% comma separated list
\keywords{Counterpoint, Haskell, SMT}  %% \keywords are mandatory in final camera-ready submission

%% \maketitle
%% Note: \maketitle command must come after title commands, author
%% commands, abstract environment, Computing Classification System
%% environment and commands, and keywords command.
\maketitle

\section{Introduction}

We demonstrate work in progress on a tool to assist in the analysis and
synthesis of musical counterpoint. Since the mid-18th century, the
composition of counterpoint has been guided by principles enunciated
in Fux's \textit{Gradus ad Parnassum} \citep{Fux1965}, first published
in 1725. Fux presents an increasingly sophisticated series of
``species'' (one note against one note, two notes against one note,
etc.) along with rules governing intervals between notes and motion
between intervals designed to ensure consonance and independence of
voices. It is well documented \citep{Mann1987} that composers
including Haydn, Mozart and Beethoven both studied and taught from
this text, and its fundamentals continue to be taught to music students
today (for example \cite{Kennan1999, Aldwell2018}).

In previous work, Cong and Leo \citep{CongLeo2019} encode the rules of
first-species (note against note) counterpoint as type constructors in
the dependently-typed language Agda. This enforces
correct-by-construction counterpoint and allows use of the Agda
typechecker to return errors with no additional effort required. On
the down side, these errors can be difficult to interpret for those
less familiar with type errors. Furthermore, encoding rules into
constructors is awkward for handling more complex species and more
global constraints. One could separate the construction of the pure
music from the constraints which can be added as a local or global
predicate (as in refinement types), but one may also wish to
deliberately violate some of the rules of strict counterpoint, as
composers often do in practice, and simply be informed of where the
violations occur without being prohibited from incorporating them.

Another approach then is to write a special-purpose ``type checker''
which can be run on previously-created music and which can generate
clear and precise error messages which are musically meaningful. The
checker is ideally easily customizable in terms of what constraints
one would like to impose. It turns out these constraints can be
expressed (at a low level) in the quantifier-free logic of linear
arithmetic and uninterpreted functions (QF-UFLIA \cite{Barrett2010}),
which allows one to not only analyze existing counterpoint, but also
synthesize counterpoint satisfying the constraints using an SMT
solver.

This tool, still a work in progress, has been implemented in Haskell
and is available at \cite{HaskellCounterpoint}. Haskell was chosen as
a high-quality and feature-rich SMT library called SBV \citep{SBV} is
available. The tool makes heavy use of functional programming, but
dependent types are sometimes wished for and the goal is to eventually
port the work to the Agda Music Tools \cite{MusicTools} library.

The following sections give an overview of current
functionality. Future plans include integration with Liquid Haskell
\citep{LiquidHaskell} (for analysis) and Synquid \citep{Synquid} (for
synthesis). In addition to handling higher species we would like to
also incorporate the rules and conventions of galant schemata
\citep{Gjerdingen2007}, which informed much of the music composed in
the era of Haydn and Mozart. Ideally one could use the tool to compose
convincing music in the galant style.

Concurrently Cong \citep{Cong2022a} is exploring another approach
using a different representation of constraints within a type system.

\section{An Example}

We briefly describe one way in which this tool can be used for both
analysis and synthesis. We first examine the example numbered 146 in
the critical edition of Beethoven's studies with Haydn
(\cite{BeethovenWerke13}; see also \cite{Nottebohm1971}, p.\ 31 and
\cite{Mann1987}, p.\ 115). It is shown in modern notation in
Figure~\ref{fig:b146}.

\begin{figure}
  \includegraphics[width=8cm]{figures/b146.png}
  \caption{Beethoven Exercise 146}
  \Description{Beethoven Exercise 146}
  \label{fig:b146}
\end{figure}

Here Haydn has supplied the top voice, marked ``C.F.'' for cantus
firmus, and Beethoven's task was to supply the remaining three voices
following first species counterpoint rules. He uses only tones from
triads, all of them complete save the last and all in root position
save the second triad in bar 4. Since the notes form triads they
automatically satisfy consonance rules (only perfect intervals, thirds
or sixths between any pair of notes vertically); note that a perfect
fourth is prohibited in two part strict counterpoint but allowed with
more voices. Perfect fourths can be found between the top two voices
in the first, fourth and last bars.

There are boundary rules that the top and bottom voices must form
octaves at the beginning and end, and motion rules which are designed
to ensure independence of voices. In particular parallel and similar
(both voices moving in the same direction) motion into fifths and
octaves is prohibited. Here we can see that Beethoven has in fact made
two errors in the top two voices: similar motion into a fifth in bar 3
and then into an octave between bars 3 and 4. Haydn fixes both errors
by changing the F in the second voice to an A.

For simplicity we now focus only on the top two voices. Feeding
Beethoven's notes into the tool as two part first species
counterpoint, it easily finds and reports errors in the use of perfect
fourths, missing octaves at the boundaries, and similar motion into
fifths and octaves. We can disable the boundary rules as they are not
relevant and relax the consonance rule to allow perfect fourths, leaving
only the motion errors. We can replace the note F that Haydn fixed
with a hole, and run the tool in synthesis mode using the same
rules. It makes the same fix Haydn did. Perhaps we gave it too much
guidance, so instead we can create three holes in a row near the
error, which generates the solution shown in
Figure~\ref{fig:b146fix3}. Note that this introduces a series of five
parallel 6ths, and there is a ``soft'' rule that one should not have
long sequences of 3rds or 6ths as that also diminishes independence
of voices. However we can easily add a constraint limiting parallel
intervals to at most two in a row; the solution returned changes the G
in bar three to an A, breaking the chain with a fifth, a small
improvement but one might hope for more.

\begin{figure}
  \includegraphics[width=8cm]{figures/b146fix3.png}
  \caption{Three Notes Changed}
  \Description{Three Notes Changed}
  \label{fig:b146fix3}
\end{figure}

This can be done by adding a constraint to increase contrary motion
(in which both voices move in opposite directions), and in fact we can
try generating our own two part counterpoint following Haydn's cantus
firmus, fixing only the first and last of Beethoven's notes. To help
create better quality counterpoint we add constraints to limit the
number of leaps (horizontal intervals of more than a major third) to
at most one and require at least six instances of contrary motion,
which turns out to be maximal. Note that it would be extremely
difficult for a human being with no computer assistance to generate
counterpoint following such restrictions, but the SMT solver instantly
returns the solution shown in Figure~\ref{fig:b146gen3}.

\begin{figure}
  \includegraphics[width=8cm]{figures/b146gen3.png}
  \caption{Generated Counterpoint}
  \Description{Generated Counterpoint}
  \label{fig:b146gen3}
\end{figure}

\section{Selected Features}

We describe at a high level some of the key features of the
library. Details can be found in the code repository
\citep{HaskellCounterpoint}.

At a high level music is represented as a list of voices, which in
turn are a list of bars, which are in turn a list of beats. Here
dependent types would be useful to ensure for example that every voice
has the same number of bars, but a run-time check can also be used to
enforce this. Each of these elements is parameterized, allowing a
number of pitch-like elements to be used to construct music. These
could be algebraic data types representing name (like C\#) and
octave, or a numeric representation of pitch. Either datum can then be
embedded into a wrapper, such as \texttt{MPitch} which can either be a
fixed pitch or a named hole to be solved for.

The above representation of music allows natural handling of first,
second and third species counterpoint (one to four notes per cantus
firmus note) as well as multi-part counterpoint. At a lower level the
music is translated into a simpler form---for example first-species
two part counterpoint is most simply represented as a list of pairs.

We plan to try to implement a constraint language as an embedded
DSL. However for now Haskell typeclasses are used, which have the
advantage that constraints can be written in natural Haskell; the only
drawback is that type signatures are somewhat abstract. The main
challenge is that we want to write constraints only once and use them
for both analysis and synthesis. The former expects low-level integers
and booleans while the latter their symbolic counterparts. This is
handled by using typeclasses which can be instantiated with either.

Analysis and synthesis then both take the same music and list of
constraints. The former checks the constraints returning error
messages in the higher-level language used to express the music. The
latter generates (if possible) music by filling in the given holes
using the constraints and the SMT solver. MIDI is also generated,
which can then be opened and viewed with notation software such as
MuseScore.

\section*{Acknowledgements}

Thanks to Youyou Cong for many enlightening discussions about type
theory and music over the past several years, and for suggesting to
look at Beethoven's exercises as a source of examples. Thanks also to
Emina Torlak and Stephen Rumph for their generosity in allowing me to
audit their classes at the University of Washington. CSE 507 was
invaluable for learning the fundamentals of SMT 
solvers, and MUHST 211 and 430 introduced me to galant schemata.

%% Bibliography
\bibliography{abstract.bib}

\end{document}

### 2022-08-10 doc/farm22/abstract.tex
work on paper

@@
 \acmISBN{}
-\acmConference[FARM '22]{}
+\acmConference[FARM '22]{Proceedings of the 10th ACM SIGPLAN International Workshop on Functional Art, Music, Modeling, and Design}{September 15, 2022}{Ljubljana, Slovenia}
 \acmBooktitle{}
@@
 \begin{abstract}
-We present a Haskell library to help analyze
+We present an Agda library to help analyze
 and synthesize musical counterpoint. The tool allows expression of
@@
 music satisfying the constraints. The tool is intended for use by
-musicians who need only have a basic knowledge of Haskell.
+musicians who need only have a basic knowledge of Agda.
 \end{abstract}
@@
 %% comma separated list
-\keywords{Counterpoint, Haskell, SMT}  %% \keywords are mandatory in final camera-ready submission
+\keywords{Counterpoint, Agda, SMT}  %% \keywords are mandatory in final camera-ready submission
 
@@
 synthesize counterpoint satisfying the constraints using an SMT
-solver.
+(Satisfiablity Modulo Theories) solver.
 
-This tool, still a work in progress, has been implemented in Haskell
-and is available at \cite{HaskellCounterpoint}. Haskell was chosen as
-a high-quality and feature-rich SMT library called SBV \citep{SBV} is
-available. The tool makes heavy use of functional programming, but
-dependent types are sometimes wished for and the goal is to eventually
-port the work to the Agda Music Tools \cite{MusicTools} library.
+The initial prototype \citep{HaskellCounterpoint} of this tool was
+written in Haskell, chosen due to the availability of the
+high-quality and feature-rich SMT library SBV \citep{SBV}. It was then
+ported to the Agda Music Tools library \cite{MusicTools}, using Agda's
+Haskell foreign function interface (FFI) to call SBV. There is also an
+Agda interface \citep{Schmitty} to SMT and we might convert to using
+that at some point in the future.
 
 The following sections give an overview of current
-functionality. Future plans include integration with Liquid Haskell
-\citep{LiquidHaskell} (for analysis) and Synquid \citep{Synquid} (for
-synthesis). In addition to handling higher species we would like to
-also incorporate the rules and conventions of galant schemata
-\citep{Gjerdingen2007}, which informed much of the music composed in
-the era of Haydn and Mozart. Ideally one could use the tool to compose
-convincing music in the galant style.
-
-Concurrently Cong \citep{Cong2022a} is exploring another approach
-using a different representation of constraints within a type system.
+functionality.
 
@@
 and then into an octave between bars 3 and 4. Haydn fixes both errors
-by changing the F in the second voice to an A.
+by changing the F in the second voice (marked with the red X) to an A.
 
@@
 number of pitch-like elements to be used to construct music. These
-could be algebraic data types representing name (like C\#) and
+could be algebraic data types representing name (like G\#) and
 octave, or a numeric representation of pitch. Either datum can then be
@@
 
+\section{Future Plans}
+
+Future plans include possible integration with Liquid Haskell
+\citep{LiquidHaskell} (for analysis) and Synquid \citep{Synquid} (for
+synthesis). In addition to handling higher species we would like to
+also incorporate the rules and conventions of galant schemata
+\citep{Gjerdingen2007}, which informed much of the music composed in
+the era of Haydn and Mozart. Ideally one could use the tool to compose
+convincing music in the galant style.
+
+Concurrently Cong \citep{Cong2022a,Cong2022c} is exploring another approach
+using a different representation of constraints within a type system.
+Recent work by Tanaka \citep{Tanaka2022} uses integer programming to
+express constraints, and we plan to compare our method with this.
+
 \section*{Acknowledgements}
@@
 solvers, and MUHST 211 and 430 introduced me to galant schemata.
+Finally, thanks the anonymous referees for many helpful comments and
+suggestions.
 

### 2022-08-16 doc/farm22/abstract.tex
2nd diagram fixed

@@
 Haskell foreign function interface (FFI) to call SBV. There is also an
-Agda interface \citep{Schmitty} to SMT and we might convert to using
+Agda interface \citep{Schmitty} to SMT; we might convert to using
 that at some point in the future.
@@
 or sixths between any pair of notes vertically); note that a perfect
-fourth is prohibited in two part strict counterpoint but allowed with
+fourth is prohibited in strict two part counterpoint but allowed with
 more voices. Perfect fourths can be found between the top two voices
@@
 error, which generates the solution shown in
-Figure~\ref{fig:b146fix3}. Note that this introduces a series of five
-parallel 6ths, and there is a ``soft'' rule that one should not have
-long sequences of 3rds or 6ths as that also diminishes independence
-of voices. However we can easily add a constraint limiting parallel
-intervals to at most two in a row; the solution returned changes the G
-in bar three to an A, breaking the chain with a fifth, a small
-improvement but one might hope for more.
+Figure~\ref{fig:b146fix3}, with the three generated notes inside the
+blue box. Although there are several possible solutions, the SMT
+solver has chosen to retain the third A while changing the first two
+notes to A as well; the only difference with Haydn's correction is
+that the first note was changed from G to A as well. Although this
+creates a satisfying oblique motion into a perfect fifth between bars
+2 and 3, this same motion occurs between bars 5 and 6; furthermore we
+now have four A notes in a row in addition to the three G notes at the
+end of the second voice. For a middle voice this is not too bad, but
+viewed as just two voices it is a big dull.
 
@@
 
-This can be done by adding a constraint to increase contrary motion
-(in which both voices move in opposite directions), and in fact we can
-try generating our own two part counterpoint following Haydn's cantus
-firmus, fixing only the first and last of Beethoven's notes. To help
-create better quality counterpoint we add constraints to limit the
-number of leaps (horizontal intervals of more than a major third) to
-at most one and require at least six instances of contrary motion,
-which turns out to be maximal. Note that it would be extremely
-difficult for a human being with no computer assistance to generate
-counterpoint following such restrictions, but the SMT solver instantly
-returns the solution shown in Figure~\ref{fig:b146gen3}.
+This can be fixed by adding a constraint to limit the number of repeated
+notes, but we can also try generating our own two part counterpoint
+following Haydn's cantus firmus, fixing only the first and last of
+Beethoven's notes. To help create better quality counterpoint we add
+constraints to limit the number of leaps (horizontal intervals of more
+than a major third) to at most one and require at least six instances
+of contrary motion, which turns out to be maximal. Note that it would
+be extremely difficult for a human being with no computer assistance
+to generate counterpoint following such restrictions, but the SMT
+solver instantly returns the solution shown in
+Figure~\ref{fig:b146gen3}.
 

### 2022-08-16 doc/farm22/abstract.tex
figures complete

@@
 error, which generates the solution shown in
-Figure~\ref{fig:b146fix3}, with the three generated notes inside the
-blue box. Although there are several possible solutions, the SMT
-solver has chosen to retain the third A while changing the first two
-notes to A as well; the only difference with Haydn's correction is
+Figure~\ref{fig:b146fix3}, with the three synthesized notes inside the
+blue box. The only difference with Haydn's correction is
 that the first note was changed from G to A as well. Although this
@@
 end of the second voice. For a middle voice this is not too bad, but
-viewed as just two voices it is a big dull.
+viewed as just two voices it is a bit dull.
 
@@
   \includegraphics[width=8cm]{figures/b146fix3.png}
-  \caption{Three Notes Changed}
-  \Description{Three Notes Changed}
+  \caption{Three Notes Synthesized}
+  \Description{Three Notes Synthesized}
   \label{fig:b146fix3}
@@
 than a major third) to at most one and require at least six instances
-of contrary motion, which turns out to be maximal. Note that it would
+of contrary motion, which turns out to be maximal. We also disallow
+perfect fourths except at the boundary. Note that it would
 be extremely difficult for a human being with no computer assistance
@@
 solver instantly returns the solution shown in
-Figure~\ref{fig:b146gen3}.
+Figure~\ref{fig:b146gen3}. Contrary motion is shown by pairs of blue
+arrows and the single leap by a red line.
 
 \begin{figure}
-  \includegraphics[width=8cm]{figures/b146gen3.png}
+  \includegraphics[width=8cm]{figures/b146gen.png}
   \caption{Generated Counterpoint}

### 2022-08-18 doc/farm22/abstract.tex
revised FARM paper

@@
 We describe at a high level some of the key features of the
-library. Details can be found in the code repository
-\citep{HaskellCounterpoint}.
+library; it is under development so details may change. The
+latest code is available at \citep{MusicTools}.
 
-At a high level music is represented as a list of voices, which in
-turn are a list of bars, which are in turn a list of beats. Here
-dependent types would be useful to ensure for example that every voice
-has the same number of bars, but a run-time check can also be used to
-enforce this. Each of these elements is parameterized, allowing a
-number of pitch-like elements to be used to construct music. These
-could be algebraic data types representing name (like G\#) and
-octave, or a numeric representation of pitch. Either datum can then be
-embedded into a wrapper, such as \texttt{MPitch} which can either be a
-fixed pitch or a named hole to be solved for.
+The core component is a simple constraint language (\texttt{Expr}) of
+booleans and integers. All constraints must be expressed at a low level
+in this language. Integers represent absolute pitches and there are
+named variables for synthesized pitches. Integer functions include
+addition, subtraction and modulo. Equalities and inequalities convert
+from integers to booleans; a monomorphic \texttt{if\_then\_else\_} allows
+the reverse.
 
-The above representation of music allows natural handling of first,
-second and third species counterpoint (one to four notes per cantus
-firmus note) as well as multi-part counterpoint. At a lower level the
-music is translated into a simpler form---for example first-species
-two part counterpoint is most simply represented as a list of pairs.
+At a higher level, constraints (\texttt{Constraint}) are represented as
+simple algebraic data types. For example \texttt{contrary}
+denotes the \texttt{MotionConstraint} that two pairs of pitches
+(each representing an vertical interval) represent contrary
+motion. Constraints are compiled into boolean expressions, and only the
+constraint author needs to understand these details; the user can
+employ the high-level constraints directly.
 
-We plan to try to implement a constraint language as an embedded
-DSL. However for now Haskell typeclasses are used, which have the
-advantage that constraints can be written in natural Haskell; the only
-drawback is that type signatures are somewhat abstract. The main
-challenge is that we want to write constraints only once and use them
-for both analysis and synthesis. The former expects low-level integers
-and booleans while the latter their symbolic counterparts. This is
-handled by using typeclasses which can be instantiated with either.
+At a higher level still, counterpoint rules (\texttt{Counterpoint})
+take music as input and generate a list of constraints. For analysis of
+music (no unknown pitches), constraints can be compiled and evaluated;
+those which evaluate to false fail to hold. This can be done
+interactively within emacs, for example.
 
-Analysis and synthesis then both take the same music and list of
-constraints. The former checks the constraints returning error
-messages in the higher-level language used to express the music. The
-latter generates (if possible) music by filling in the given holes
-using the constraints and the SMT solver. MIDI is also generated,
-which can then be opened and viewed with notation software such as
-MuseScore.
+For synthesis, the code must currently be compiled and run. The Agda
+constraint language is converted to Haskell and compiled into symbolic
+integers and booleans using SBV (\texttt{Smt}). An SMT solver
+(currently only Z3 is supported) is called to solve for the variable
+pitches, if possible. The result is then output as MIDI
+(\texttt{SmtInterface}) which can then be opened and viewed with
+notation software such as MuseScore.
 
